Acute avoidance of hydrogen sulfide is modulated by external and internal states in C. elegans

Version 1

Abstract

Abstract

Hydrogen sulfide (H2S) can act as an energy source, a poison and a gasotransmitter in organisms. We used the robust locomotory responses to H2S in Caenorhabditis elegans to delineate the molecular mechanisms governing sensory and adaptive responses to H2S exposure. We found that C. elegans exhibited transiently increased locomotory activity and turning behavior as a strategy to escape the noxious H2S stimulation. The behavioral responses to H2S were modulated by a complex network of signaling pathways, including cyclic GMP signaling in ciliated sensory neurons, calcineurin, nuclear hormone receptors, to the major starvation regulators such as insulin and TGF-β signaling. The response to H2S was substantially affected by the ambient O2 levels and their prior experience in low O2 environments, suggesting an intricate interplay between O2 and H2S sensing mechanisms. Prolonged exposure to H2S robustly evoked H2S detoxification coupled with reduced locomotory response to the subsequent H2S challenges. Intriguingly, the expression of genes involved in iron homeostasis, including ftn-1 and smf-3, was substantially modified in exposure to H2S, implying that labile iron levels are affected by H2S. In support of this, iron supplement significantly bolstered the behavioral response to H2S. In addition, mitochondria, one of the central hubs for H2S metabolism, played a crucial role in adaptive responses to H2S. In summary, our study provides molecular insights into the mechanisms through which C. elegans detects, modulates, and adapts its response to H2S.

Introduction

The detection and response to chemical signals are crucial to interact effectively with the environment in various organisms, spanning from bacteria to humans. During the late Proterozoic era, the rise in oxygen (O2) levels eliminated hydrogen sulfide (H2S) from most of the habitats. Yet, low O2 levels together with high concentrations of H2S and carbon dioxide (CO2) can still persist in enclosed environments where bacteria actively decompose organic matter (Olson & Straub, 2016). H2S can easily penetrate biological membranes and interfere with various cellular processes, leading to toxicity. One of its detrimental effects is the disruption of cellular respiration by binding to cytochrome c oxidase (COX) and inhibiting the mitochondrial electron transport chain (Cooper & Brown, 2008; Khan et al., 1990; Nicholls & Kim, 1982b).

Animals inhabiting environments rich in H2S have evolved mechanisms to either avoid or adapt to high concentrations of H2S, providing them with survival advantages. Species that encounter periodic increases in H2S often possess sensing abilities and exhibit avoidance behaviors. For example, benthic species undertake daily vertical migrations between sulfidic and non-sulfidic waters (Abel, Koenig, & Davis, 1987; Salvanes, Utne-Palm, Currie, & Braithwaite, 2011). Certain metazoan species have seized the ecological opportunities presented by H2S-rich environments through the development of cellular adaptations, allowing them to occupy a unique ecological niche that is inaccessible to the other organisms. Livebearing fish species, for instance, have evolved COX subunits that are more resistant to H2S and are more efficient in H2S detoxification (Kelley et al., 2016; Pfenninger et al., 2014).

Free living nematodes thrive in decaying matter where they are exposed to significant fluctuations in O2, CO2, and H2S levels over short distances (Rodriguez-Kabana, Jordan, & Hollis, 1965). Nematodes like C. elegans have evolved the ability to utilize these gasous stimuli to navigate their environment, locate bacterial food sources, or escape the dangers (Bretscher, Busch, & de Bono, 2008; Carrillo, Guillermin, Rengarajan, Okubo, & Hallem, 2013). Acute exposure to H2S triggers transiently increased locomotion in C. elegans (Budde & Roth, 2011). However, prolonged exposure to H2S induces cellular adaptation to survive the toxic environment. C. elegans can tolerate H2S levels up to 50 ppm, but the concentrations exceeding 150 ppm prove lethal (Budde & Roth, 2010, 2011; Fawcett, Hoyt, Johnson, & Miller, 2015; Horsman, Heinis, & Miller, 2019; Miller, Budde, & Roth, 2011a; Miller & Roth, 2007). Extended periods of low O2 and/or high H2S exposure stabilize the hypoxia-inducible factor HIF-1, leading to transcriptional reprogramming (Budde & Roth, 2010). Through HIF-1-dependent and independent mechanisms, acclimation of C. elegans to 50 ppm H2S extends lifespan and improves the response to hypoxia-induced protein aggregation (Budde & Roth, 2010; Fawcett et al., 2015; Horsman et al., 2019; Miller & Roth, 2007; Qabazard et al., 2014).

In this study, we investigated how the nematode C. elegans responds acutely to avoid an increase of H2S levels, as well as how it adapts at the molecular and physiological levels if they fail to escape the H2S exposure.

Results

cGMP signaling in ASJ cilia contributes to H2S avoidance

When encountering aversive cues such as noxious gases, C. elegans initiates a pirouette followed by an increase of its locomotory activity to escape the noxious stimuli. When acutely exposed to 150 ppm H2S, C. elegans exhibited an increased locomotory speed and enhanced turning behavior, while decreasing the frequency of reversals, as a strategy to escape the noxious stimuli (Figure 1—figure supplement 1A–C). The maximum speed was reached after 6–8 minutes in H2S, and returned to the baseline over a period of approximately 1 hour (Figure 1—figure supplement 1D). The reduced locomotory activity after 2 hours in H2S was fully reversible if these animals were immediately challenged by a different stimulus such as hypoxia (Figure 1—figure supplement 1E). The magnitude of the avoidance responses appeared to be concentration dependent. Under our assay conditions, acute behavioral response to 50 ppm H2S was minimal and lacked consistency, whereas the response to 150 ppm H2S was robust and reproducible (Figure 1—figure supplement 1F).

To gain molecular insight into H2S evoked locomotory response, we performed a candidate screen for mutants that failed to increase their locomotory speed when H2S levels acutely increased to 150 ppm (Supplementary file 1). We discovered that mutants defective in ciliogenesis exhibited diminished H2S responses (Figure 1A–C). These included daf-19 mutants lacking cilia, dyf-3 mutants with reduced cilia length, and dyf-7 mutants where ciliogenesis occurs but the cilia are not anchored to their sensilla and are not exposed to the external environment (Heiman & Shaham, 2009; Murayama, Toh, Ohshima, & Koga, 2005; Starich et al., 1995; Swoboda, Adler, & Thomas, 2000). Our observations suggest that cilia exposure to the external environment is necessary for H2S detection. Cilia-mediated sensory responses often involve the activation of guanylate cyclases and the opening of cGMP-gated channels (Ferkey, Sengupta, & L’Etoile, 2021). We observed attenuated H2S responses in tax-4 or tax-2 mutants, lacking the key subunits of cilia-enriched cGMP-gated channels (Figure 1D and E). Comparable observations have also been reported in acute response to CO2, suggesting that locomotory responses to CO2 and H2S are similar (Bretscher et al., 2008; Hallem et al., 2011; Hallem & Sternberg, 2008). Consistent with this, mutations in genes encoding the calcineurin subunits TAX-6 or CNB-1, or nuclear hormone receptor NHR-49, impaired both H2S and CO2 responses (Hallem & Sternberg, 2008) (Figure 1—figure supplement 2A–C). Additionally, locomotory response to H2S was also repressed following a 24-hour period of starvation, as observed for CO2 (Figure 1—figure supplement 2D) (Bretscher et al., 2008; Hallem & Sternberg, 2008). In C. elegans, the regulation of starvation-responses involves the reduced activity of two major pathways: insulin and TGF-β signaling. Disrupting the genes encoding the insulin receptor DAF-2 or the TGF-β ligand DAF-7 eliminated locomotory activity in response to acute H2S (Figure 1F and G, Figure 1—figure supplement 2E). Insulin and TGF-β signaling involve the inhibition of downstream transcription factors DAF-16 or DAF-3 respectively. H2S avoidance defects of daf-2 and daf-7 mutants were fully suppressed in daf-2; daf-16 or in daf-7; daf-3 double mutants (Figure 1F and G).

cGMP signaling in ASJ neurons contributes to H2S avoidance

(A–H) Locomotory responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT (N2) and daf-19(m86) (A); WT and dyf-3(m185) (B); WT and dyf-7(m539) (C); WT, daf-11(m47) and tax-4(p678) (D); WT and tax-2(p691) (E); WT, daf-2(e1370), daf-16(mgDf47) and daf-2(e1370); daf-16(mgDf47) double mutants (F); WT, daf-7(e1372), daf-3(mgDf90) and daf-7(e1372); daf-3(mgDf90) double mutants (G); WT and gcy-9(tm7632) (H). In this and subsequent figures for H2S assay, red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes, labeled as a and b respectively in Figure 1D, F, G, J and K) used for statistical analysis unless otherwise indicated. **** = p < 0.0001, *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test. (I) Comparison between the acute locomotory response of WT animals to 150 ppm H2S balanced with 7% O2 and the acute locomotory response to 5% CO2 balanced with 7% O2. Red bars on the x-axis indicate two intervals (3-4 minutes and 11-12 minutes) used for statistical analysis. * = p < 0.05, ns = not significant, Mann–Whitney U test. (J) Locomotory responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT, daf-11(m47) and transgenic daf-11(m47) expressing daf-11 genomic DNA in ASJ neurons (two independent lines). *** = p < 0.001, ** = p < 0.01, * = p < 0.05, Mann–Whitney U test. (K) Locomotory responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT, tax-4(p678) and transgenic tax-4(p678) expressing tax-4 cDNA in ASJ neurons (two independent lines). *** = p < 0.001, ** = p < 0.01, * = p < 0.05, Mann–Whitney U test.

These observations prompted us to further investigate if the same molecular and circuit mechanisms underlay the acute avoidance to CO2 and H2S. The response to CO2 requires receptor guanylate cyclase GCY-9 (Hallem et al., 2011). Interestingly, GCY-9 was dispensable for the locomotory responses to H2S (Figure 1H). In addition, animals responded more rapidly to CO2 than to H2S, but with a reduced magnitude (Figure 1I), implying that distinct mechanisms likely drive these two responses. In a comprehensive survey of all guanylate cyclase mutants, the daf-11 strain emerged as the only mutant showing defects in the avoidance of H2S (Figure 1D, Supplementary file 1). Selective expression of daf-11 in ASJ neurons, but not in other neurons, partially restored the locomotory response to acute H2S (Figures 1J, Figure 1—figure supplement 2F–I). Consistently, the H2S response defect of tax-4 mutants was also rescued by selective expression of its cDNA in ASJ neurons (Figure 1K), suggesting that the receptor guanylate cyclase DAF-11 and the CNG channel TAX-4 partially act in ASJ neurons to promote acute response to H2S. This differed from the site of action of the CO2 sensor GCY-9, which primarily operates in BAG neurons (Hallem et al., 2011). Taken together, these observations suggest that common signaling pathways modulate the locomotory responses to both H2S and CO2, although the molecular and circuit underpinning of H2S and CO2 sensing are different.

Activation of the O2 sensing circuit antagonizes H2S avoidance

The sensory responses of wild C. elegans are adapted to their living environment, which is characterized by low O2 levels and high concentrations of CO2. In comparison to the domesticated lab strain N2, wild isolates of C. elegans exhibit a strong response to high O2 but are relatively indifferent to CO2 (Beets et al., 2020; Bretscher et al., 2008; Hallem & Sternberg, 2008). Local H2S concentrations could also be significantly higher in decomposing substances where wild C. elegans thrives. We suspected that wild strains could display a modified response to H2S in comparison to standard laboratory N2 animals. Indeed, wild isolates CB4855, CB4856, and CB4858 exhibited attenuated responses to H2S (Figure 2A, Figure 2—figure supplement 1A). One predominant genetic trait in these strains is a naturally occurring variation of neuropeptide receptor NPR-1 215F, which is less active than NPR-1 215V in N2 animals (de Bono & Bargmann, 1998). This led us to investigate whether the NPR-1 signaling plays a role in modulating acute avoidance of H2S. Mutating npr-1 in N2 strain background was sufficient to abolish locomotory responses to H2S, as observed in the wild isolates (Figure 2B). NPR-1 primarily acts within the RMG interneurons to modulate responses to environmental stimuli by reducing signal output from these neurons (Laurent et al., 2015; Macosko et al., 2009). The selective expression of the highly active npr-1(215V) variant in RMG neurons effectively restored H2S evoked locomotory response in npr-1 mutant worms (Figure 2C). Furthermore, optogenetic stimulation of RMG interneurons using channelrhodopsin-2 (ChR2) significantly dampened locomotory response to H2S of N2 animals, suggesting high RMG activity inhibits H2S responses (Figure 2—figure supplement 1B).

Activation of O2 sensing circuit antagonizes H2S avoidance

(A) Locomotory responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for WT animals, wild isolates CB4858 and CB4855. Red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes, labeled as a and b respectively), used for statistical analyses. ** = p < 0.01, * = p < 0.05, Mann–Whitney U test. (B, C) Locomotory responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT and npr-1(ad609) (B); WT, npr-1(ad609) and transgenic npr-1(ad609) expressing npr-1 in RMG neurons using Cre-LoxP system (Macosko et al., 2009) (C). Red bars on the x-axis indicate two intervals (4-5 minutes and 11-12 minutes) used for statistical analysis. **** = p < 0.0001, *** = p < 0.001, ** = p < 0.01, Mann–Whitney U test. (D–G) Locomotory responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT, npr-1(ad609), gcy-35(ok769) and npr-1(ad609); gcy-35(ok769) double mutants (D); WT, npr-1(ad609), tax-4(p678) and npr-1(ad609); tax-4(p678) double mutants (E); WT, npr-1(ad609), qaIs2241(genetic ablation of AQR, PQR and URX neurons) and npr-1(ad609); qaIs 2241 (F); WT, npr-1(ad609), npr-1(ad609); gcy-35(ok769) double mutants and transgenic npr-1(ad609); gcy-35(ok769) double mutants expressing gcy-35 cDNA under gcy-37 promoter, which drives expression in O2 sensing neurons (G). Red bars on the x-axis indicate two intervals (4-5 minutes and 11-12 minutes) used for statistical analysis. *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test.

The sensation of 21% O2 involves the soluble guanylate cyclases GCY-35/GCY-36 and the downstream cyclic nucleotide gated (CNG) channels TAX-4/TAX-2 (Busch et al., 2012; Cheung, Cohen, Rogers, Albayram, & de Bono, 2005; Couto, Oda, Nikolaev, Soltesz, & de Bono, 2013; Gray et al., 2004; Laurent et al., 2015; Persson et al., 2009; Zimmer et al., 2009). We examined if disrupting O2 sensing machinery had the same effect as the inhibition of RMG signaling by NPR-1(215V). We found that silencing O2 sensing neurons by mutating gcy-35 and tax-4, or ablating these neurons by cell-specific expression of the pro-apoptotic gene egl-1 restored H2S evoked behavioral response in npr-1 null mutants (Figure 2D–F). Specific expression of gcy-35 cDNA in URX, AQR, PQR was sufficient to repress acute response to H2S in npr-1; gcy-35 double mutants (Figure 2G). Finally, enhancing the presynaptic activities of URX, AQR, PQR neurons by the expression of a gain-of-function PKC-1(A160E) (Dekker, Mcintyre, & Parker, 1993; Hiroki et al., 2022) significantly dampened H2S regulated locomotory activity (Figure 2—figure supplement 1C). Taken together, these observations indicate that activation of O2 sensing circuit is inhibitory to H2S evoked locomotory response.

H2S exposure reprograms gene expression in C. elegans

Prolonged exposure to H2S has a dramatic impact on the physiology of C. elegans: 50 ppm H2S exposure extends its lifespan, while 150 ppm H2S is lethal (Budde & Roth, 2010, 2011; Fawcett et al., 2015; Horsman et al., 2019; Miller & Roth, 2007 Miller, 2011 #5). We conducted a comparative analysis of the transcriptome profiles in animals exposed to either 50 ppm or 150 ppm H2S for the durations of 1 hour, 2 hours, or 12 hours. Our RNA-seq analysis revealed that a one-hour exposure to either 50 ppm or 150 ppm H2S was sufficient to trigger significant changes in gene expression, with 518 or 304 genes showing differential expression, respectively (Figure 3A, Supplementary file 2, p<1e-10). Genes induced by 1-hour and 2-hour H2S exposure displayed a considerable overlap (Figure 3B, Figure 3— figure supplement 1A). Gene Ontology (GO) analysis revealed that biological processes such as defense against bacteria and cysteine biosynthesis from L-serine were immediately activated upon H2S exposure (Figure 3C and D, Figure 3—figure supplement 1B and C). As expected, we observed a robust induction of genes involved in H2S detoxification (Figure 3E and F) (Miller, Budde, & Roth, 2011b; Niu et al., 2011; Vora et al., 2022). H2S is detoxified in mitochondria by sulfide:quinone oxidoreductase SQRD-1 to persulfide, which is then metabolized by sulfur dioxygenase ETHE-1 to generate sulfite. Thiosulfate transferase further oxidizes sulfite to produce thiosulfate (Hildebrandt & Grieshaber, 2008). Glutathione S-transferases (GSTs) remove accumulated sulfur molecules during H2S oxidation (Jackson, Melideo, & Jorns, 2012). Notably, gst-19 and sqrd-1 were among the most significantly upregulated genes after one or two hours of exposure to either 50 ppm or 150 ppm H2S, while ethe-1 showed a weak increase (Figure 3E and F, Figure 3—figure supplement 1D and E, Supplementary file 2). H2S activates the HIF-1 pathway through the interaction of sulhydrylase/cysteine synthase CYSL-1 with the HIF-1 proline hydroxylase EGL-9 (Ma, Vozdek, Bhatla, & Horvitz, 2012). We found that a proportion of genes with HIF-1 regulated promoters displayed increased expression after 1 or 2 hours of H2S exposure (Figure 3— figure supplement 1F) (Miller et al., 2011b; Vora et al., 2022). These genes included the above-mentioned detoxifying enzymes GST-19, SQRD-1, and ETHE-1, as well as enzymes involved in cysteine metabolism, CYSL-1, CYSL-2, and CDO-1 (Supplementary file 2). Mutations increasing SKN-1 activity bypass the requirement for HIF-1 to survive in high H2S (Horsman et al., 2019). Consistent with early observations (Miller et al., 2011a), we also observed that a set of SKN-1 targets were differentially expressed in H2S (Figure 3—figure supplement 1G) (Niu et al., 2011). This included the heat shock protein encoding genes hsp-16.2 and hsp-16.41 (Figure 3E and F, Figure 3—figure supplement 1D and E).

Prolonged H2S exposure reprograms gene expression in C. elegans

(A) Venn diagram displaying the number of differentially expressed genes after 1-hour exposure to 50 ppm or 150 ppm H2S balanced with 7% O2 in WT animals. p<1e-10. (B) Venn diagram displaying the number of differentially expressed genes after 1-hour, 2-hour and 12-hour exposure to 150 ppm H2S balanced with 7% O2 in WT animals. p<1e-10. (C, D) Significantly enriched GO categories for differentially expressed genes with adjusted p value <1e-10 in WT animals exposed to 50 ppm (C) or 150 ppm (D) H2S balanced with 7% O2 for 1 hour. (E, F) Volcano plots showing the differentially expressed genes with adjusted p value <1e-10 in WT animals exposed to 50 ppm (E) or 150 ppm (F) H2S balanced with 7% O2 for 1 hour. A set of genes involved in H2S detoxification, cysteine metabolism and stress response were highlighted in red.

Surprisingly, we also observed a rapid increase in the expression of genes associated with intracellular H2S production, such as semo-1 and mpst-3 genes (Figure 3E and F, Figure 3—figure supplement 1D and E) (Philipp et al., 2022; Qabazard et al., 2014). However, these data aligned with earlier observations made in sulfide spring fish (Kelley et al., 2016; Mathew, Schlipalius, & Ebert, 2011; Qabazard et al., 2014).

When exposed to the H2S for 12 hours, the differentially expressed genes showed reduced overlap with those identified at earlier time points (Figure 3B, Figure 3—figure supplement 1A). This suggests that different defense strategies against H2S might be employed at various stages of exposure. Although similar sets of genes were induced after 1h or 2h at 50 ppm and 150 ppm H2S, the overlap between 50ppm and 150ppm decreased after 12 hours exposure (Figure 3—figure supplement 1H–J). This observation suggests that defense mechanisms might also differ for prolonged exposure at low and high levels of H2S. In particular, we noted that 150 ppm H2S specifically triggered the expression of the HIF-1 regulator rhy-1 and the HIF-1 target nhr-57 at all exposure time points, which was less pronounced in response to 50 ppm H2S (Figures 3E and F, Figure 3—figure supplement 1D, E, K and L).

Acute response to H2S is modulated by HIF-1 signaling

Consistent with early observations (Budde & Roth, 2010), our RNA-seq data suggested that H2S exposure activates HIF-1 pathway (Figure 3—figure supplement 1F). Stabilization of HIF-1 by H2S is critical for animals to survive H2S exposure by the induction of detoxification genes as well as other adaptative responses (Ma et al., 2012). Such an adaptation is usually accompanied with the reduction in sensory sensitivity to this stressor. Therefore, we suspected that HIF-1 stabilization might dampen the acute response to H2S. Under normoxia conditions, the conserved proline-4-hydroxylase PHD/EGL-9 hydroxylates HIF-1, which is subsequently recognized by the von Hippel-Lindau (VHL) tumor suppressor protein for degradation (Kaelin & Ratcliffe, 2008; Semenza, 2010). We found that disrupting hif-1 resulted in extremely transient behavioral response to H2S characterized by an initial bout of activity and a rapid decline of locomotory speed to the basal level (Figure 4A). While eliminating PHD/EGL-9 or VHL activity, which stabilizes HIF-1, the locomotory response to H2S was completely inhibited (Figure 4B). egl-9; hif-1 or vhl-1; hif-1 double mutants had the identical response as hif-1 mutants (Figure 4C and D), confirming that mutating egl-9 or vhl-1 inhibits acute response to H2S by activating HIF-1. Acclimating animals in 1% O2 for more than 12 hours also significantly decreased acute response to H2S (Figure 4E). In addition, animals carrying the non-degradable form of HIF-1 in all neurons were defective in H2S evoked locomotory responses (Figure 4F). Taken together, these observations suggest that constitutive activation of HIF-1 promotes the adaption to H2S and inhibits the behavioral responses to H2S stimulation.

Acute response to H2S is modulated by HIF-1 signaling

(A–D) Locomotory responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT and hif-1(ia4) (A); WT, egl-9(sa307) and vhl-1(ok161) (B); WT, egl-9(sa307), hif-1(ia4) and egl-9(sa307); hif-1(ia4) double mutants (C); WT, vhl-1(ok161), hif-1(ia4) and vhl-1(ok161); hif-1(ia4) double mutants (D). Red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes, labeled as a and b respectively), used for statistical analyses. *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test. (E) Locomotory responses of WT animals to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 after 0, 12, 24 or 36 hours in 1% O2. Red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes, labeled as a and b respectively), used for statistical analyses. *** = p < 0.001, ** = p < 0.01, ns = not significant, Mann–Whitney U test. (F) Locomotory responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals carrying the non-degradable form of HIF-1. Non-degradable form of HIF-1 (P621A and P621G) are expressed from a pan-neuronal promoter unc-14 or form hif-1 endogenous promoter, respectively. Red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes, labeled as a and b respectively), used for statistical analyses. ** = p < 0.01, ns = not significant, Mann–Whitney U test. (G–J) Locomotory responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT and sqrd-1(tm3378) (G); WT and ethe-1(yum2895) (H); WT, cysl-1(ok762), cysl-2(ok3516) and cysl-3(yum4) (I); WT and semo-1(yum2889) (J). Red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes, labeled as a and b respectively), used for statistical analyses. *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test.

The transient behavioral response to H2S in hif-1 mutants suggested that animals deficient of HIF-1 can still detect the presence of H2S. This observation together with the importance of HIF-1 signaling in H2S detoxification provoked us to explore whether detoxification enzymes were also required for the sustained locomotory activity in H2S. Similar to hif-1 mutants, both sqrd-1 and ethe-1 mutants displayed a transient response to H2S (Figure 4G and H). We next examined the contribution of other upregulated genes in response to H2S, including the genes encoding sulfhydrylase/cysteine synthases CYSL-1, CYSL-2, and CYSL-3 as well as SEMO-1. Disrupting CYSLs and SEMO-1 encoding genes phenocopied hif-1 or sqrd-1 mutants (Figure 4I and J). Our data suggest that while H2S detoxification system is not essential for H2S sensing, it is required for preserving animals’ locomotory activity in the presence of H2S.

Labile iron pool is critical to sustain the locomotory activity in H2S

C. elegans tightly regulates iron homeostasis through the control of iron transporter SMF-3, as well as the regulation of iron storage protein Ferritins FTN-1 and FTN-2 (Cha’on et al., 2007; Gourley, Parker, Jones, Zumbrennen, & Leibold, 2003; Kim, Cho, Yoo, & Ahnn, 2004; Pekec et al., 2022). Among the most differentially regulated genes, ftn-1 appeared to be consistently downregulated by H2S (Figure 5A and B, Figure 5—figure supplement 1A–D). The inhibited expression of ftn-1 and ftn-2 as well as the elevated expression of smf-3 (Figure 5A and B, Figure 5—figure supplement 1A–D) prompted us to explore if labile iron levels play a role in behavioral response to H2S exposure. The labile iron pool is chelatable and redox-active. In the presence of iron chelator 2,2’-dipyridyl (BP), H2S evoked locomotory response was fully inhibited (Figure 5C). This was not caused by the general sickness of BP-exposed animals as they responded robustly to acute hypoxia (Figure 5—figure supplement 1E). By contrast, the supplementation of ferric ammonium citrate (FAC) significantly prolonged animals’ locomotory activity in H2S (Figure 5C). In addition, disrupting ftn-1, which was expected to increase labile iron levels (Anderson & Leibold, 2014; Romney, Newman, Thacker, & Leibold, 2011), promoted the locomotory activity of animals in H2S (Figure 5D), Conversely, mutating smf-3, which impaired iron uptake (Anderson & Leibold, 2014; Romney et al., 2011), led to a faster decline of locomotory activity in H2S (Figure 5D). Furthermore, depleting labile iron pool by ftn-1 overexpression resulted in an attenuated response to H2S (Figure 5E), while it did not affect acute response to hypoxia (Figure 5—figure supplement 1F). These data suggests that high levels of free iron sustain locomotory activity in H2S, whereas low iron levels inhibit behavioral responses to H2S. It has been reported that hypoxia inhibits the expression of ftn-1 and ftn-2, while upregulating smf-3 expression (Ackerman & Gems, 2012; Gourley et al., 2003; Romney et al., 2011). Additionally, hif-1 mutants have been reported to display reduced intracellular iron content and increased susceptibility to iron starvation (Rajan et al., 2019; Romney et al., 2011). We explored if the locomotory defect of hif-1 mutants in H2S could be explained by iron deficiency (Rajan et al., 2019; Romney et al., 2011). The presence of FAC significantly improved H2S evoked locomotory activity of hif-1 mutants (Figure 5F). In addition, disrupting ftn-1 partially suppressed the locomotory defect of hif-1 mutants in H2S (Figure 5G). These data confirm that labile iron pool is critical to maintain the locomotory activity of animals in H2S.

Labile iron pool is critical to sustain the locomotory activity in H2S

(A, B) Volcano plots showing the relative expression of the genes involved in the regulation of iron homeostasis in WT animals after 2-hour exposure with 50 ppm (A) or 150 ppm (B) H2S balanced with 7% O2. (C) Locomotory responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for WT animals, WT animals pretreated with 100 μM 2,2′-Bipyridyl (BP) and WT animals pretreated with 5 mg/ml ferric ammonium citrate (FAC) in the presence of food for 16 hours. Red bars on the x-axis represent two intervals (4–5 minutes and 14–15 minutes, labeled as a and b respectively), used for statistical analyses. ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test. (D) Locomotory responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT, smf-3(ok1305) and ftn-1(ok3625). Red bars on the x-axis represent two intervals (4–5 minutes and 14–15 minutes, labeled as a and b respectively), used for statistical analysis. *** = p < 0.001, * = p < 0.05, ns = not significant, Mann–Whitney U test. (E) Locomotory responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for WT animals and animals overexpressing ftn-1 genomic DNA under its own promoter (#1 and #2 indicate two independent lines). Red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes, labeled as a and b respectively), used for statistical analysis. **** = p < 0.0001, ** = p < 0.01, ns = not significant, Mann–Whitney U test. (F, G) Locomotory responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT, hif-1(ia4) and hif-1(ia4) mutants pretreated with 5 mg/ml ferric ammonium citrate (FAC) for 16 hours (F); WT, hif-1(ia4), ftn-1(ok3625) and hif-1(ia4); ftn-1(ok3625) double mutants (G). Red bars on the x-axis represent two intervals (4–5 minutes and 8–9 minutes) used for statistical analysis. * = p < 0.05, ns = not significant, Mann– Whitney U test.

Iron deficiency causes mitochondrial dysfunction including decreased cytochrome concentration and respiratory efficiency (Dallman, 1986; Masini, Salvioli, et al., 1994; Masini, Trenti, et al., 1994; Walter et al., 2002). H2S is known to exert both positive and negative effects on the mitochondria, which are not only the primary targets of H2S but also represent one of the main cellular hubs of H2S detoxification (Cooper & Brown, 2008; Khan et al., 1990; Nicholls & Kim, 1982a). We probed the possibility that impairing mitochondrial function might alter locomotory response to H2S. To this end, we tested mutants with the disruption of gas-1, clk-1, mev-1, or isp-1 genes. Similar to iron depletion, these mutants abrogated locomotory response to H2S (Figure 6A–D). Additionally, exposing animals to the mitochondrial complex I inhibitor rotenone fully repressed locomotory response to H2S (Figure 6E). While rotenone exposure inhibits H2S response, it evoked increased locomotory activity over 2 hours (Figure 6E). These observations confirm the critical role of mitochondria in supporting locomotory response in exposure to H2S in C. elegans.

Mitochondrial function is required for acute response to H2S

(A–D) Locomotory responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT and gas-1(fc21) (A); WT and clk-1(qm30) (B); WT and mev-1(kn1) (C); WT and isp-1(qm150) (D). Red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes) used for statistical analysis. **** = p < 0.0001, ** = p < 0.01, * = p < 0.05, Mann–Whitney U test. (E) Locomotory responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for WT animals pretreated with 10 μM rotenone for 0 minute, 10 minutes, 30 minutes, 60 minutes or 120 minutes. Red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes, labeled as a and b respectively), used for statistical analysis. *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test.

Discussion

Animals’ behavior and physiology are profoundly influenced by the surrounding environment, in which they reside. Wild C. elegans isolates thrive in the decomposing matters, where the local concentrations of O2 are low while the levels of CO2 and H2S could be high. These animals have adapted their behavior in such an environment, displaying increased sensitivity to high O2 exposure but dampened responses to CO2. In contrast, the standard laboratory strain N2, which is constantly exposed to high O2 but low CO2, is relatively insensitive to elevated levels of O2 but robustly avoids CO2 exposure (Beets et al., 2020; Bretscher et al., 2008; Carrillo et al., 2013; Hallem & Sternberg, 2008). As observed in response to CO2, N2 animals exhibit strong reorientation and acceleration to escape the H2S exposure, while wild isolates display a significantly diminished response to H2S.

The responses to CO2 and H2S in N2 animals are modulated by common signaling pathways, which include the ciliogenesis, the cGMP-gated channels TAX-2 and TAX-4, calcineurin subunit TAX-6/CNB-1, and nuclear hormone receptor NHR-49. Additionally, these two responses are affected by the nutritional state of the animals. In C. elegans, starvation is signaled by reduced activity in two major signaling pathways: the insulin and TGF-β pathways. H2S and CO2 responses are abolished in starved animals or mutants that disrupt insulin or TGF-β signaling. Therefore, the modulation of behavioral responses to H2S and CO2 by external and internal factors appears highly similar. However, CO2 response is mediated by the guanylate cyclase GCY-9 in BAG neurons (Hallem et al., 2011), which appears to be dispensable for H2S sensing. We observed a prominent role for the guanylate cyclase DAF-11 expressed in the chemosensory neurons ASJ for H2S responses. daf-11 signaling in ASJ also mediates chemosensory responses to volatile chemicals and to NO (Birnby et al., 2000; Hao et al., 2018). However, DAF-11 controls the normal larval development of C. elegans and may have pleiotropic indirect effects on sensory functions (Murakami, Koga, & Ohshima, 2001). In absence of more direct evidence, its role as H2S sensor in ASJ remain uncertain.

Long-term exposure to H2S can elicit adaptive responses including changes in gene expression patterns to mitigate the toxic effects of H2S. Within just one hour of H2S exposure, we observed the upregulation of genes associated with H2S detoxification. This response enables animals to detoxify H2S and maintain cellular homeostasis under sulfidic conditions. This adaptive strategy is accompanied by the reduced locomotory response to H2S exposure. The HIF-1 (hypoxia-inducible factor) signaling is a major pathway involved in the adaption during prolonged H2S exposure (Budde & Roth, 2011; Miller et al., 2011b). When the HIF-1 pathway is constitutively induced in the nervous system, the behavioral responses to H2S are eliminated. Similarly, reduced behavioral responses to H2S are observed after 12 hours of exposure to 1% O2. It has been proposed that persulfides, polysulfides, and thiosulfate, which are the by-products during sulfide detoxification, play a role in H2S signaling (Mishanina, Libiad, & Banerjee, 2015). However, in the case of locomotion responses to H2S in C. elegans, this is unlikely to be the mechanism for the initial response, as mutants lacking detoxification enzymes still exhibit an acute and transient response to H2S. Interestingly, the sustained locomotory activity was abolished in these mutants, suggesting that either the by-products of H2S detoxification contribute to maintain the behavioral response or the lack of detoxification enzyme results in rapid cellular toxicity, preventing sustained locomotory activity in H2S (Budde & Roth, 2011).

Interestingly, we provide evidence that labile iron pool is critical for H2S responses. Removal of labile iron pool by the chelator BP or ftn-1 overexpression attenuated both acute and sustained responses to H2S, suggesting that intracellular iron is required for the sensation of H2S. While increasing labile iron levels sustains the locomotory activity in H2S. In C. elegans, labile iron homeostasis is predominantly regulated by the HIF-1 signaling, the activation of which decreases the transcription of ftn-1 and increases the expression of smf-3 (Ackerman & Gems, 2012; Romney et al., 2011). Consistent with this, we have observed a downregulation of the ferritins FTN-1 and FTN-2, as well as an upregulation of the iron transporter SMF-3 during exposure to H2S. This suggests that H2S may control the expression of FTN-1 and SMF-3, and consequently the labile iron pool, through the HIF-1 pathway.

Although the exact mechanisms are not yet fully understood, several possibilities shed light on the importance of labile iron for H2S responses. Fe2+ acts as a critical cofactor for the PHD/EGL-9 enzymes, which are involved in the hydroxylation-dependent destabilization of HIF-1 (Myllyharju & Kivirikko, 1997). PHD inactivation by iron sequestration by ferritin can increase HIF-1 signaling, suggesting a potential role of labile iron as a negative feedback loop to limit HIF-1 activity (Siegert et al., 2015). In addition, iron is necessary for the biosynthesis of iron-sulfur (Fe-S) clusters, critical cofactors involved in redox reactions and mitochondrial respiration (Read, Bentley, Archer, & Dunham-Snary, 2021; Xu & Moller, 2011). Labile iron is beneficial for the biogenesis of Fe-S clusters despite its oxidative properties under normal O2 conditions (Liochev, 1996). Interestingly, hypoxia has been observed to enhance Fe-S cluster biogenesis by increasing the availability of bioactive iron (Ast et al., 2019). Therefore, low oxidative stress may alleviate the need to maintain a low labile iron pool. In this context, the antioxidant properties of H2S or the interruption of mitochondrial respiration by H2S might provide condition favorable to high labile iron pool (Shefa, Kim, Jeong, & Jung, 2018; Sunda, Kieber, Kiene, & Huntsman, 2002).

The primary mechanism underlying the toxic effects of H2S has been suggested to be the direct inhibition of cytochrome c oxidase, a critical enzyme involved in mitochondrial respiration (Cooper & Brown, 2008; Khan et al., 1990; Nicholls & Kim, 1982b). Intriguingly, we have observed that genetic and pharmacological inhibition of mitochondrial respiration prevents the locomotory response of C. elegans to H2S. This effect may be mediated by the induction of the HIF-1 pathway, which has been observed in mutants with impaired mitochondrial respiration (Lee, Hwang, & Kenyon, 2010). Furthermore, acute inhibition of mitochondrial respiration by rotenone also enhances locomotion while suppressing the response to H2S. Reducing mitochondrial respiration has been found to stimulate aversive behavior in C. elegans (Melo & Ruvkun, 2012). Therefore, the acute inhibition of mitochondrial respiration by H2S could transiently promote aversive behavior in C. elegans, while long-term inhibition of mitochondrial respiration may contribute to the induction of the HIF-1 pathway and adaptation to H2S.

In summary, this study unveils a novel behavior of C. elegans in their avoidance to H2S. The response to H2S is intricately shaped by their environment, where it integrates with other sensory signals, including food availability, CO2 and O2 levels. We emphasize the vital role of H2S detoxification system, particularly the involvement of mitochondria and labile irons, in sustaining the locomotory activity in H2S. In the future, it will be exciting to elucidate the neuronal mechanisms driving H2S avoidance and to explore how the changes of labile iron pools modulate cellular and organismal behavior responses to noxious sensory stimuli.

Materials and Methods

Strains

C. elegans were maintained using standard protocols (Brenner, 1974). Strains used in this study are listed in Supplementary file 1 and Supplementary file 3.

Preparation of H2S gas mixture

Different H2S concentrations were created as previously described (Miller & Roth, 2007). Briefly, the H2S-containing gas mixture were prepared by diluting 5000 ppm H2S in nitrogen (N2) with 7% O2 balanced with N2. The gas flow was tuned by Sierra Smart-Trak 100 mass flow controllers. H2S concentration in the mixture was measured using two H2S detectors (MSA ALTAIR 2X gas detector for H2S and Clip Single Gas Detector, SDG, CROWCON). The pre-defined gas mixtures of 7% O2, 1% O2 and 5% CO2 with N2 were purchased from Air Liquide Gas AB. 5000 ppm H2S stock in N2 was obtained from Linde Gas AB. Gas mixtures in all experiments were hydrated before use.

Molecular Biology

The Multisite Gateway system (Thermo Fisher Scientific, United States) was used to generate expression vectors. Promoters, including gcy-37 (2.7kb), trx-1 (1 kb), ftn-1(2 kb), daf-11(3 kb), odr-3(2.7 kb), gpa-11(3 kb), sra-6(3 kb), odr-1(2.4 kb), flp-21(4.1kb), and ocr-2(2.4 kb) were amplified from N2 genomic DNA and cloned into pDONR P4P1 using BP clonase. gcy-35, tax-4 and pkc-1 cDNAs were amplified using the first strand cDNA library as the template, while daf-11 and ftn-1 genomic sequences were amplified using genomic DNA as the template. The entry clones of these genes pDONR 221 were generated using BP reaction. To generate the gain of function mutation of pkc-1 (A160E), Q5 Site-Directed Mutagenesis Kit (NEB) was used according to manufacturer instructions. All expression vectors were generated using LR reaction. The primer sequences were displayed in Supplementary file 4.

To generate transgenic animals, the daf-11 expression vectors were injected at the concentration of 20 ng/μl supplemented with 50 ng/μl of a coelomocyte co-injection marker (unc-122p::GFP) and 30 ng/μl of 1kb DNA ladder. For tax-4 gene, the injection mixtures were prepared using 40 ng/μl of tax-4 expression vectors, 50 ng/μl of a coelomocyte co-injection marker and 10 ng/μl of 1kb DNA ladder. The rest of expression constructs were injected at 50 ng/μl together with 50 ng/μl of a coelomocyte marker.

CRISPR/Cas9 genome editing

Genes were disrupted using CRISPR/Cas-9 mediated genome editing as described (Dokshin, Ghanta, Piscopo, & Mello, 2018; Ghanta & Mello, 2020). The strategy involved the utilization of homology-directed insertion of custom designed single strand DNA template (ssODN), which had two homology arms of 35bp flanking the targeted PAM site. Between two homology arms, a short sequence containing a unique restriction enzyme cutting site and in frame as well as out of frame stop codons was included. The insertion of ssODN template would delete 16 bases of coding sequence, ensuring the proper gene disruption. To prepare the injection cocktail, 0.5 μl of Cas-9 protein (IDT) was mixed with 5 μl of 0.4μg/μl tracrRNA (IDT, United States) and 2.8 μl of 0.4 μg/μl crRNA (IDT). The mixture was incubated at 37°C for at least 10 minutes before 2.2 μl of 1μg/μl ssODN (or 500 ng dsDNA) and 2 μl of 0.6 μg/μl rol-6 co-injection marker were added. Nuclease-free water was used to bring the final volume to 20 μl. The injection mixture was centrifuged for 2 minutes before use.

Behavioral assays

H2S evoked locomotion activity was monitored as described previously (Laurent et al., 2015; Zhao et al., 2022). Briefly, OP50 bacteria were seeded on the assay plates 16 hours before use. The border of bacterial lawn was removed using a PDMS stamp. For each assay, 25 to 30 day-one adult animals were picked onto assay plates, allowed to settle down for 15 minutes, and subsequently sealed within microfluidic chambers. A syringe pump (PHD 2000, Harvard Apparatus) was employed to deliver gas mixtures into the microfluidic chamber at a constant flow of 3 ml/min. The rapid switch between different gas mixtures were controlled by Telfon valves coupled with ValveBank Perfusion Controller (AutoMate Scientific). The locomotory activity at different gas mixtures were monitored using a high-resolution camera (FLIR) mounted on a Zeiss Stemi 508 microscope. Videos were captured at a rate of 2 frames per second, starting with a 2-minute recording in 7% O2, followed by 11 minutes in H2S, and ending with a final 5-minute period in 7% O2. For the long-term recording in H2S, video capture commenced with a 2-minute period in 7% O2, followed by a duration of 148 minutes in 150 ppm H2S, and concluded with a final 2-minute interval in 7% O2.

To assess the effects of rotenone on H2S evoked locomotory response, day-one adult animals were subjected to 10 μM rotenone for 10min, 30min, 1 hour and 2 hours on the drug containing plates. Subsequently, the rotenone-treated animals were assayed on the standard assay plates without the drug. To explore the impact of ferric ammonium citrate (FAC) (F5879, Sigma-Aldrich) and 2,2′-Bipyridyl (BP) (D216305, Sigma-Aldrich) on behavioral response to H2S, L4 animals were exposed to 100 μM FAC or 5mg/ml BP for 16 hours in the presence of bacterial food. The FAC or BP treated animals were assayed on FAC or BP containing plates.

Optogenetic experiments were performed as previously outlined (Zhao et al., 2022). Transgenic L4 animals were exposed to 100μM all-trans retinal (ATR) (R2500, Sigma) in the dark for 16 hours. The ATR-fed animals were assayed with the illumination of 70 μW/mm2 blue light during the whole assay period. Blue light was emitted from an ultra-high-power LED lamp (UHP-MIC-LED-460, Prizmatix), and the light intensity was determined using a PM50 Optical Power Meter (ThorLabs). To minimize the effect of transmitted light on the transgenic animals, a long-pass optical filter was used to eliminate the lights with short wavelengths during picking. All videos of behavioral analysis were analyzed using a home-made MATLAB program Zentracker (https://github.com/wormtracker/zentracker). For all behavior analyses, at least three assays were performed for each strain with more than 80 worms in total.

RNA extraction and sequencing

To obtain RNA samples for RNA-seq, thirty young adult animals were picked to each fresh plate and allowed to lay eggs for 2 hours, after which the adult animals were removed and eggs were allowed to develop into day one adults. Day one animals were subsequently challenged with H2S for three different time points (1 hour, 2 hours and 12 hours). For each time point, five plates of day one animals were exposed to 50 ppm H2S, 150 ppm H2S or 7% O2 as a control group. Subsequently, the animals were immediately collected and rinsed three times with M9 buffer. After washing, the worm pellet was frozen in liquid nitrogen. Animals were homogenized using Bullet Blender (Next Advance) in the presence of Qiazol Lysis Reagent and 0.5mm Zirconia beads at 4°C. RNA was prepared using RNeasy Plus Universal Mini Kit (Qiagen). Six independent RNA samples were prepared for all different treatments. The 2100 Bioanalyzer instrument (Agilent) was used for the RNA quality control with a microfluidic chip specific for RNA (Agilent RNA 6000 Nano Kit). Then 1 μg of qualified RNA samples in RNase-free ddH2O was used for library preparation. Library was constructed and sequenced by Novogene.

RNA-seq analysis

The RNA-seq data was aligned using STAR v2.7.9a (Dobin et al., 2013) and gene expression counts extracted by featureCounts v2.0.3 (Liao, Smyth, & Shi, 2014). Differential expression was calculated using DESeq2 (Love, Huber, & Anders, 2014). GO analysis was performed using the EnrichR R package (Kuleshov et al., 2016) on genes having p.adj < 1e-10. For brevity, several highly similar GO categories were manually omitted (full EnrichR output available at Github repository).

Data availability

Raw sequencing data has been deposited at ArrayExpress # E-MTAB-13296. The R code is available at https://github.com/henriksson-lab/ce_h2s_tc.

eLife Assessment

The manuscript aims to better understand the mechanisms underlying the behavioral responses of C. elegans to hydrogen sulfide, a toxin known to exert remarkable effects on animal physiology in a range of contexts. To this end, the authors provide a series of useful findings regarding the mechanisms by which hydrogen sulfide may be sensed, their relationships to other gas-sensing pathways, and the role of a variety of physiological pathways in responding to hydrogen sulfide exposure. While some of the findings are solid, other aspects of the paper are incomplete, such that some claims are incompletely supported, and an integrated understanding of the authors' observations does not clearly emerge.

Reviewer #1 (Public Review):

Summary:

This paper sets out to achieve a deeper understanding of the effects of hydrogen sulfide on C. elegans behavior and physiology, with a focus on behavior, detection mechanism(s), physiological responses, and detoxification mechanisms.

Strengths:

The paper takes full advantage of the experimental tractability of C. elegans, with thorough, well-designed genetic analyses.

Some evidence suggests that H2S may be directly detected by the ASJ sensory neurons.

The paper provides interesting and convincing evidence for complex interactions between responses to different gaseous stimuli, particularly an antagonistic role between H2S and O2 detection/response.

Intriguing roles for mitochondria and iron homeostasis are identified, opening the door to future studies to better understand the roles of these components and processes.

Weaknesses:

The claim that worms' behavioral responses to H2S are mediated by direct detection is incompletely supported. While a role for the chemosensory neuron ASJ is implicated, it remains unclear whether this reflects direct detection. Other possibilities, including indirect effects of ASJ and the guanylyl cyclase daf-11 on O2 responses, are also consistent with the authors' data.

The role of H2S-mediated damage in behavioral responses, particularly when detoxification pathways are disrupted, remains unclear.

The findings of the paper are somewhat disjointed, such that a clear picture of the relationships between H2S detection, detoxification mechanisms, mitochondria, and iron does not emerge from these studies. Most importantly, the relative roles of H2S detection and integration, vs. general and acute mitochondrial crisis, in generating behavioral responses are not convincingly resolved.

Reviewer #2 (Public Review):

Summary:

H2S is a gas that is toxic to many animals and causes avoidance in animals such as C. elegans. The authors show that H2S increases the frequency of turning and the speed of locomotion. The response was shown to be modulated by a number of neurons and signaling pathways as well as by ambient oxygen concentrations. The long-term adaptation involved gene expression changes that may be related to iron homeostasis as well as the homeostasis of mitochondria.

Strengths:

Overall, the authors provide many pieces that will be important for solving how H2S signals through neuronal circuits to change gene expression and physiological programs. The experiments rely mostly on a behavioral assay that measures the increase of locomotion speed upon exposure to H2S. This assay is then combined with manipulations of environmental factors, different wild-type strains, and mutants. The mutants analyzed were obtained as candidates from the literature and from transcriptional profiling that the authors carried out in worms that were exposed to H2S. These studies imply several genetic signaling pathways, some neurons, and metabolism-related factors in the response to H2S. Hence the data provided should be useful for the field.

Weaknesses:

On the other hand, many important aspects of the underlying mechanisms remain unsolved and the reader is left with many loose ends. For example, it is not clear how H2S is actually sensed, how sensory neurons are activated and signal to downstream circuits, and what the role of ciliated and RMG neurons is in this circuit. It remains unclear how signals lead to gene expression and physiological changes such as metabolic rewiring. Solving all this would clearly be beyond the scope of a single manuscript. Yet, the manuscript also does not focus on understanding one of these central aspects and rather is all over the place, which makes it harder to understand for readouts that are not in this core field. Multiple additional methods and approaches exist to dig deeper into these mechanisms in the future, such as neuronal calcium imaging, optogenetics, and metabolic analysis. To generate a story that will be interesting to a broad readership substantial additional experimentation would be required. Further, in the current manuscript, it is often difficult to understand the rationales of the experiments, why they were carried out, and how to place them into a context. This could be improved in terms of documentation, narration/explanation, and visualization.

Reviewer #3 (Public Review):

Summary:

The manuscript explores the behavioral responses of C. elegans to hydrogen sulfide, which is known to exert remarkable effects on animal physiology in a range of contexts. The possibility of genetic and precise neuronal dissection of responses to H2S motivates the study of responses in C. elegans. The manuscript is well-written in communicating the complex physiology around C. elegans behavioral responses to H2S and in appropriately citing prior and related relevant work.

There are three parts to the manuscript.

In the first, an immediate behavioral response-increased locomotory rate-upon exposure to H2S is characterized. The experimental conditions are critical, and data are obtained from exposure of animals to 150ppm H2S at 7% O2. The authors provide evidence that this is a chemosensory response to H2S, showing a requirement for genes encoding components of the cilia apparatus and implicating a role for tax-4 and daf-11. Neuron-specific rescue in the ASJ neurons suggests the ASJ neurons contribute to the response to H2S. One caveat is that previous work has shown that the dauer-constitutive phenotype of daf-11 mutants can be suppressed by ASJ ablation, suggesting that there may be pervasive changes in animal nervous system signaling that are ASJ-dependent in daf-11 mutants, which may indirectly alter chemosensory responses to H2S. More direct methods to assess whether ASJ senses H2S, e.g. using calcium imaging, would better assess a direct role for the ASJ neurons in a behavioral response to H2S. The authors also point out interesting parallels between the response to H2S and CO2 though provide some genetic data separating the two responses. Importantly, the authors note that when aerotaxis (O2-sensing and movement) in the presence of bacterial food is intact, as in npr-1 215F animals, the response to H2S is abrogated. Mutation in gcy-35 in the npr-1 215F background restores the H2S chemosensory response.

There is a second part of the paper that conducts transcriptional profiling of the response to H2S that corroborates and extends prior work in this area.

The final part of the paper is the most intriguing, but for me, also the most problematic. The authors examine how H2S-evoked locomotory behavioral responses are affected in mutants defective in the stress and detoxification response to H2S, most notably hif-1. Prior genetic studies have established the pathways leading to HIF-1 activation/stabilization, as well as potential downstream mechanisms. The authors conduct logical genetic analysis to complement studies of the hif-1 mutant and in part motivated by their transcriptional profiling studies, examine the role of iron sequestration/free iron in the locomotory response to H2S, and further speculate on how the behavior of mutants defective in mitochondrial function might be affected by exposure to H2S.

In some regard, this part of the manuscript is interesting because the analysis begins to connect how the behavior of an animal to a toxic compound is affected by mutations that affect sensitivity to the toxic compound. However, what is unclear is what is being studied at this point. In the context, of noting that H2S at 150ppm is known to be lethal, its addition to mutants clearly sensitized to its effects would be anticipated to have pervasive effects on animal physiology and nervous system function. The authors note that the continued increased locomotion of wild-type animals upon H2S exposure might be due to the byproducts of detoxification or the detrimental effects of H2S. The latter explanation seems much more likely, in which case what one may be observing is the effects of general animal sickness, or even a bit more specifically, neuronal dysfunction in the presence of a toxic compound, on locomotion. As such, what is unclear is what conclusions can be taken away from this part of the work.

Strengths:

1. Characterization of a motor behavior response to H2S

2. Transcriptional profiling of the response to H2S corroborating prior work.

Weaknesses:

Unclear significance and experimental challenges regarding the study of locomotory responses to animals sensitized to the toxic effects of H2S under exposure to H2S.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

Summary:

This paper sets out to achieve a deeper understanding of the effects of hydrogen sulfide on C. elegans behavior and physiology, with a focus on behavior, detection mechanism(s), physiological responses, and detoxification mechanisms.

Strengths:

The paper takes full advantage of the experimental tractability of C. elegans, with thorough, welldesigned genetic analyses.

Some evidence suggests that H2S may be directly detected by the ASJ sensory neurons. The paper provides interesting and convincing evidence for complex interactions between responses to different gaseous stimuli, particularly an antagonistic role between H2S and O2 detection/response. Intriguing roles for mitochondria and iron homeostasis are identified, opening the door to future studies to better understand the roles of these components and processes.

We thank the reviewer for the supportive comments.

Weaknesses:

The claim that worms' behavioral responses to H2S are mediated by direct detection is incompletely supported. While a role for the chemosensory neuron ASJ is implicated, it remains unclear whether this reflects direct detection. Other possibilities, including indirect effects of ASJ and the guanylyl cyclase daf-11 on O2 responses, are also consistent with the authors' data.

We thank the reviewer for the insightful comment and agree that the role of ASJ neurons in H2S detection was not clear. We included new experiments and revised our text to make it clearer.

Since our initial analyses suggest a role of ASJ neurons in H2S-evoked locomotory responses (Figure 2F and G), We thought that this would offer us a starting point to dissect the neuronal circuit involved in H2S responses. Expression of the tetanus toxin catalytic domain in ASJ, which blocks neurosecretion, inhibited H2S-evoked locomotory speed responses (Figure 2H), suggesting that neurosecretion from ASJ promotes H2S-evocked response (Lines 162–165). We then performed calcium imaging of ASJ neurons in response to H2S exposure. However, while we observed CO2-evoked calcium transients in ASJ using GCaMP6s, we did not detect any calcium response to H2S, under several conditions, including animals on food, off food, and with different H2S concentrations and exposure times (Figure2—Figure supplement 2E and F) (Lines 166–170). Since signaling from ASJ neurons regulates developmental programs that modify sensory functions in C. elegans (Murakami et al., 2001), the involvement of ASJ neurons is not specific to H2S and ASJ neurons are unlikely to serve as the primary H2S sensor (Discussed in Line 449–458). Therefore, the exact sensory neuron, circuit and molecular triggers mediating acute H2S avoidance remain to be elucidated.

Our subsequent investigation on mitochondrial components suggests that a burst of mitochondrial ROS production may be the trigger for H2S avoidance, as transient exposure to rotenone substantially increases baseline locomotory speed (Figure 7E) (Line 391–396). However, to initiate avoidance behavior to H2S, mitochondrial ROS could potentially target multiple neurons and cellular machineries, making it challenging to pinpoint specific sites of action. Nevertheless, we agree that further dissection of the neural circuits and mitochondrial signaling in H2S avoidance will be important and should be explored in future studies.

The role of H2S-mediated damage in behavioral responses, particularly when detoxification pathways are disrupted, remains unclear.

We thank the reviewer for the insightful comment and fully agree with the concern raised. The same issue was also noted by the other reviewers. We agree that decreased locomotory responses in H2S-sensitized animals can arise from distinct causes, either systemic toxicity or behavioral adaptation, and distinguishing between these is critical. We have included new experiments and revised the text to clarify this issue.

Our data suggest that increased initial omega turns and a rapid loss of locomotion in hif-1 and detoxification-defective mutants including sqrd-1 and ethe-1 likely reflect an enhanced sensitivity to H2S toxicity due to their failure to induce appropriate adaptative responses (Figure 5D–F, Figure 5J–L, Figure 5—Figure supplement 1F–P). Supporting this, hif-1 mutants become less responsive to unrelated stimuli (near-UV light) after 30 minutes of H2S exposure (Figure 5I).

In contrast, egl-9 and SOD-deficient animals show reduced initial omega-turn and reduced speed responses (Figure 5B, Figure 7G, Figure 5—Figure supplement 1A and B, and Figure 7—Figure supplement 1F and G), although both egl-9 and sod mutants respond normally to the other stimuli prior or after H2S exposure (Figure 5I, Figure 5—Figure supplement 1C, and Figure 7—Figure supplement 1H). Since disrupting egl-9 stabilizes HIF-1 and upregulates the expression of numerous genes involved in cellular defense against H2S toxicity, the enhanced detoxification capacity in egl-9 mutants likely increases animals’ tolerance to H2S, thereby reducing avoidance to otherwise toxic H2S levels. Similarly, persistently high ROS in SOD deficient animals activates a variety of stress-responsive signaling pathways, including HIF-1, NRF2/SKN-1 and DAF-16/ FOXO signaling (Lennicke & Cocheme, 2021; Patten et al., 2010), facilitating cellular adaptation to redox stress and reducing animals’ responsiveness to toxic H2S levels. Taken together, these findings support the view that reduced locomotory speed during H2S exposure can arise from distinct mechanisms: early systemic toxicity in hif-1 and detoxificationdefective mutants, versus enhanced cellular adaptation in egl-9 and SOD mutants. We have integrated the relevant information across the result section and discussed this in Lines 485-536.

The findings of the paper are somewhat disjointed, such that a clear picture of the relationships between H2S detection, detoxification mechanisms, mitochondria, and iron does not emerge from these studies. Most importantly, the relative roles of H2S detection and integration, vs. general and acute mitochondrial crisis, in generating behavioral responses are not convincingly resolved.

We thank the reviewer for this comment and agree that our presentation did not fully connect different findings into a cohesive picture. To address this, we have acquired new data, and revised the abstract, results and discussion sections to clarify two phases of H2S-evoked responses: an initial avoidance behavior upon H2S exposure, followed by a later phase of adaption and detoxification when the escape is not successful.

In brief, we began with the basic characterization of H2S-induced locomotory speed response, followed by a candidate gene screen to identify key molecules and pathways involved in initial speed response to H2S. Subsequently, we focused on three major intersecting pathways that contributed to the acute behavioral response to H2S. These include cGMP signaling, which led to the identification of ASJ neurons; nutrient-sensitive pathways that modulate behavioral responses to both H2S and CO2; and O2sensing signaling, whose activation inhibits responses to H2S. However, the molecules and neurons in these pathways, including ASJ, likely play modulatory roles and are unlikely to serve as the primary H2S sensors. Our subsequent analysis, however, suggests that mitochondria play a critical role in triggering avoidance behavior upon H2S exposure. Brief treatment with rotenone, a potent inducer of ROS, led to marked increase in locomotory speed (Figure 7E). This suggests the possibility that a burst of ROS production triggered toxic levels of H2S (Jia et al., 2020) may initiate the avoidance behavior.

When the initial avoidance fails, H2S detoxification programs are induced as a long-term survival strategy. The induction of detoxification programs appears to enhance tolerance to H2S exposure and contributes to the gradual decrease of locomotory speed in H2S. We now provide a clearer image of how different pathways modulate H2S detoxification and adaptation (see our responses to other comments). Briefly, mutants defective in detoxification, such as hif-1 and other detoxification-defective mutants, showed stronger initial omega-turn response and a rapid loss of locomotion. This loss of locomotion is likely caused by early cellular toxicity as the mutants failed to respond to other unrelated stimuli (nearUV light) after 30 minutes of H2S exposure (Figure 5I). Likewise, smf-3 mutants and BP-treated animals were hypersensitive to H2S (Figure 6D and E, and Figure 6—Figure supplement 1G and I), likely due to impaired H2S detoxification under low iron conditions, as iron is a co-factor required for the activity of the H2S detoxification enzyme ETHE-1 (Figure 5K and Figure 5—Figure supplement 1E).

In contrast, reduced locomotion and response in other contexts such as egl-9 mutants and SODdeficient animals reflect H2S-induced adaptive mechanism rather than toxicity as they remain responsive to the other stimuli after H2S exposure. Since disrupting egl-9 stabilizes HIF-1 and upregulates the expression of numerous genes involved in cellular defense against H2S toxicity, the enhanced detoxification capacity in egl-9 mutants likely increases animals’ tolerance to H2S, thereby reducing avoidance to otherwise toxic H2S levels. Similarly, persistently high ROS in SOD deficient animals activates a variety of stress-responsive signaling pathways, including HIF-1, NRF2/SKN-1 and DAF-16/ FOXO signaling (Lennicke & Cocheme, 2021; Patten et al., 2010), facilitating cellular adaptation to redox stress and reducing animals’ responsiveness to toxic H2S levels. Therefore, different animals decline their locomotory speed to the effects of H2S through distinct mechanisms. We have integrated the relevant information across the result section and discussed this in Lines 485-536.

Reviewer #2 (Public Review):

Summary:

H2S is a gas that is toxic to many animals and causes avoidance in animals such as C. elegans. The authors show that H2S increases the frequency of turning and the speed of locomotion. The response was shown to be modulated by a number of neurons and signaling pathways as well as by ambient oxygen concentrations. The long-term adaptation involved gene expression changes that may be related to iron homeostasis as well as the homeostasis of mitochondria.

Strengths:

Overall, the authors provide many pieces that will be important for solving how H2S signals through neuronal circuits to change gene expression and physiological programs. The experiments rely mostly on a behavioral assay that measures the increase of locomotion speed upon exposure to H2S. This assay is then combined with manipulations of environmental factors, different wild-type strains, and mutants. The mutants analyzed were obtained as candidates from the literature and from transcriptional profiling that the authors carried out in worms that were exposed to H2S. These studies imply several genetic signaling pathways, some neurons, and metabolism-related factors in the response to H2S. Hence the data provided should be useful for the field.

We thank the reviewer for the supportive comments.

Weaknesses:

On the other hand, many important aspects of the underlying mechanisms remain unsolved and the reader is left with many loose ends. For example, it is not clear how H2S is actually sensed, how sensory neurons are activated and signal to downstream circuits, and what the role of ciliated and RMG neurons is in this circuit. It remains unclear how signals lead to gene expression and physiological changes such as metabolic rewiring. Solving all this would clearly be beyond the scope of a single manuscript. Yet, the manuscript also does not focus on understanding one of these central aspects and rather is all over the place, which makes it harder to understand for readouts that are not in this core field. Multiple additional methods and approaches exist to dig deeper into these mechanisms in the future, such as neuronal calcium imaging, optogenetics, and metabolic analysis. To generate a story that will be interesting to a broad readership substantial additional experimentation would be required. Further, in the current manuscript, it is often difficult to understand the rationales of the experiments, why they were carried out, and how to place them into a context. This could be improved in terms of documentation, narration/explanation, and visualization.

We thank the reviewer for the comment, which has also been raised by the other reviewers. We agree that our initial submission was poorly presented. We also acknowledge the fact that some aspects, such as detailed neural circuit and sensory transduction, still remain unresolved. We have now included additional experiments and revised the manuscript to clarify the logic of our experiments, provided better context for our findings, and improved both the narrative flow and data visualization to make the manuscript more accessible to readers. We now provide a clearer image of how different pathways interact to modulate the initial avoidance response, and the H2S detoxification and behavioral habituation during prolonged H2S exposure. The following response is similar to the one for reviewer #1.

In brief, we began with the basic characterization of H2S-induced locomotory speed response, followed by a candidate gene screen to identify key molecules and pathways involved in initial speed response to H2S. Subsequently, we focused on three major intersecting pathways that contributed to the acute behavioral response to H2S. These include cGMP signaling, which led to the identification of ASJ neurons; nutrient-sensitive pathways that modulate behavioral responses to both H2S and CO2; and O2sensing signaling, whose activation inhibits responses to H2S. However, the molecules and neurons in these pathways, including ASJ, likely play modulatory roles and are unlikely to serve as the primary H2S sensors. Our subsequent analysis, however, suggests that mitochondria play a critical role in triggering avoidance behavior upon H2S exposure. Brief treatment with rotenone, a potent inducer of ROS, led to marked increase in locomotory speed (Figure 7E). This suggests the possibility that a burst of ROS production triggered toxic levels of H2S (Jia et al., 2020) may initiate the avoidance behavior.

When the initial avoidance fails, H2S detoxification programs are induced as a long-term survival strategy. The induction of detoxification programs appears to enhance tolerance to H2S exposure and contributes to the gradual decrease of locomotory speed in H2S. We now provide a clearer image of how different pathways modulate H2S detoxification and adaptation (see our responses to other comments). Briefly, mutants defective in detoxification, such as hif-1 and other detoxification-defective mutants, showed stronger initial omega-turn response and a rapid loss of locomotion. This loss of locomotion is likely caused by early cellular toxicity as the mutants failed to respond to other unrelated stimuli (nearUV light) after 30 minutes of H2S exposure (Figure 5I). Likewise, smf-3 mutants and BP-treated animals were hypersensitive to H2S (Figure 6D and E, and Figure 6—Figure supplement 1G and I), likely due to impaired H2S detoxification under low iron conditions, as iron is a co-factor required for the activity of the H2S detoxification enzyme ETHE-1 (Figure 5K and Figure 5—Figure supplement 1E).

In contrast, reduced locomotion and response in other contexts such as egl-9 mutants and SODdeficient animals reflect H2S-induced adaptive mechanism rather than toxicity as they remain responsive to the other stimuli after H2S exposure. Since disrupting egl-9 stabilizes HIF-1 and upregulates the expression of numerous genes involved in cellular defense against H2S toxicity, the enhanced detoxification capacity in egl-9 mutants likely increases animals’ tolerance to H2S, thereby reducing avoidance to otherwise toxic H2S levels. Similarly, persistently high ROS in SOD deficient animals activates a variety of stress-responsive signaling pathways, including HIF-1, NRF2/SKN-1 and DAF-16/ FOXO signaling (Lennicke & Cocheme, 2021; Patten et al., 2010), facilitating cellular adaptation to redox stress and reducing animals’ responsiveness to toxic H2S levels. Therefore, different animals decline their locomotory speed to the effects of H2S through distinct mechanisms. We have integrated the relevant information across the result section and discussed this in Lines 485-536.

Reviewer #3 (Public Review):

Summary:

The manuscript explores the behavioral responses of C. elegans to hydrogen sulfide, which is known to exert remarkable effects on animal physiology in a range of contexts. The possibility of genetic and precise neuronal dissection of responses to H2S motivates the study of responses in C. elegans. The manuscript is well-written in communicating the complex physiology around C. elegans behavioral responses to H2S and in appropriately citing prior and related relevant work.

There are three parts to the manuscript.

In the first, an immediate behavioral response-increased locomotory rate-upon exposure to H2S is characterized. The experimental conditions are critical, and data are obtained from exposure of animals to 150ppm H2S at 7% O2. The authors provide evidence that this is a chemosensory response to H2S, showing a requirement for genes encoding components of the cilia apparatus and implicating a role for tax-4 and daf-11. Neuron-specific rescue in the ASJ neurons suggests the ASJ neurons contribute to the response to H2S. One caveat is that previous work has shown that the dauer-constitutive phenotype of daf-11 mutants can be suppressed by ASJ ablation, suggesting that there may be pervasive changes in animal nervous system signaling that are ASJ-dependent in daf-11 mutants, which may indirectly alter chemosensory responses to H2S. More direct methods to assess whether ASJ senses H2S, e.g. using calcium imaging, would better assess a direct role for the ASJ neurons in a behavioral response to H2S. The authors also point out interesting parallels between the response to H2S and CO2 though provide some genetic data separating the two responses. Importantly, the authors note that when aerotaxis (O2sensing and movement) in the presence of bacterial food is intact, as in npr-1 215F animals, the response to H2S is abrogated. Mutation in gcy-35 in the npr-1 215F background restores the H2S chemosensory response.

There is a second part of the paper that conducts transcriptional profiling of the response to H2S that corroborates and extends prior work in this area.

The final part of the paper is the most intriguing, but for me, also the most problematic. The authors examine how H2S-evoked locomotory behavioral responses are affected in mutants defective in the stress and detoxification response to H2S, most notably hif-1. Prior genetic studies have established the pathways leading to HIF-1 activation/stabilization, as well as potential downstream mechanisms. The authors conduct logical genetic analysis to complement studies of the hif-1 mutant and in part motivated by their transcriptional profiling studies, examine the role of iron sequestration/free iron in the locomotory response to H2S, and further speculate on how the behavior of mutants defective in mitochondrial function might be affected by exposure to H2S.

In some regard, this part of the manuscript is interesting because the analysis begins to connect how the behavior of an animal to a toxic compound is affected by mutations that affect sensitivity to the toxic compound. However, what is unclear is what is being studied at this point. In the context, of noting that H2S at 150ppm is known to be lethal, its addition to mutants clearly sensitized to its effects would be anticipated to have pervasive effects on animal physiology and nervous system function. The authors note that the continued increased locomotion of wild-type animals upon H2S exposure might be due to the byproducts of detoxification or the detrimental effects of H2S. The latter explanation seems much more likely, in which case what one may be observing is the effects of general animal sickness, or even a bit more specifically, neuronal dysfunction in the presence of a toxic compound, on locomotion. As such, what is unclear is what conclusions can be taken away from this part of the work.

Strengths:

(1) Characterization of a motor behavior response to H2S

(2) Transcriptional profiling of the response to H2S corroborating prior work.

We thank the reviewer for the supportive comments.

Weaknesses:

Unclear significance and experimental challenges regarding the study of locomotory responses to animals sensitized to the toxic effects of H2S under exposure to H2S.

We thank the reviewer for the comment, which has also been raised by the other reviewers. We agree that our initial submission left several important questions open, and we acknowledge the fact that some aspects, such as detailed neural circuit and sensory transduction, still remain unresolved. Nevertheless, we acquired new data and revised our text, aiming to clarify the distinct mechanisms underlying the reduced locomotion in different mutants during prolonged H2S exposure.

Our data suggest that increased initial omega turns and a rapid loss of locomotion in hif-1 and detoxification-defective mutants including sqrd-1 and ethe-1 likely reflect an enhanced sensitivity to H2S toxicity due to their failure to induce appropriate adaptative responses (Figure 5D–F, Figure 5J–L, Figure 5—Figure supplement 1F–P). Supporting this, hif-1 mutants become less responsive to unrelated stimuli (near-UV light) after 30 minutes of H2S exposure (Figure 5I).

In contrast, egl-9 and SOD-deficient animals show reduced initial reorientation and reduced speed responses (Figure 5B, Figure 7G, Figure 5—Figure supplement 1A and B, and Figure 7—Figure supplement 1F and G), although both egl-9 and sod mutants respond normally to the other stimuli prior or after H2S exposure (Figure 5I, Figure 5—Figure supplement 1C, and Figure 7—Figure supplement 1H). Since disrupting egl-9 stabilizes HIF-1 and upregulates the expression of numerous genes involved in cellular defense against H2S toxicity, the enhanced detoxification capacity in egl-9 mutants likely increases animals’ tolerance to H2S, thereby reducing avoidance to otherwise toxic H2S levels. Similarly, constant high ROS in SOD deficient animals activates a variety of stress-responsive signaling pathways, including HIF-1, NRF2/SKN-1 and DAF-16/ FOXO signaling (Lennicke & Cocheme, 2021; Patten et al., 2010), facilitating cellular adaptation to redox stress and reducing animals’ responsiveness to toxic H2S levels. Taken together, these findings support the view that reduced locomotory speed during H2S exposure can arise from distinct mechanisms: early systemic toxicity in hif-1 and detoxification-defective mutants, versus enhanced cellular adaptation in egl-9 and SOD mutants. We have integrated the relevant information across the result section and discussed this in Lines 485-536.

Reviewer #1 (Recommendations For The Authors):

To better substantiate a role for H2S detection, it would be useful for the authors to image Ca responses to H2S in ASJ in WT and unc-13, and to rule out the possibility that the requirement for daf-11 in ASJ reflects a role in O2 rather than H2S detection.

We thank the reviewer for this comment. As suggested, we performed calcium imaging of ASJ neurons using GCaMP6s. As previously described, 3% CO2 evoked a calcium transient in ASJ (Figure 2—figure supplement 2F). To investigate whether H2S evoked a calcium transient in ASJ neurons, we tested several conditions, including animals on food or off food, with different H2S concentrations (~75 or ~150ppm), and different exposure time (4 or 8 mins). However, we did not detect a calcium response to H2S in ASJ under any of the conditions tested (Figure2—figure supplement 2E) (Lines 166–168). Given that neuronspecific rescue of daf-11 or tax-4 mutants pointed to a role of ASJ neurons in promoting H2S responses, we sought to determine how ASJ neurons were involved. Expression of the tetanus toxin catalytic domain in ASJ neurons, which blocks neurosecretion, inhibited H2S-evoked locomotory speed responses (Figure 2H), similar to the phenotypes observed in daf-11 and daf-7 mutants (Figure 2C and D) (Lines 162–165). These results confirm that ASJ activity and neurosecretion contribute to the H2S responses, although ASJ is unlikely to serve as the primary H2S sensor. One potential explanation is that DAF-7 released by ASJ controls the starvation program, which in turn modulates the animal’s response to H2S. We also discussed this in Lines 449–458.

The paper would be significantly strengthened by testing the possibility (as the authors acknowledge in lines 348-52) that disruption of detoxification mechanisms reduces sustained behavioral responses to H2S because of physiological damage. Authors use acute exposure to high O2 for this purpose earlier in the paper, but not to probe the consequences of loss of hif-1 and detoxification factors.

We thank the reviewer for the valuable suggestion. As the reviewer highlighted, we attributed the brief locomotory speed responses to H2S observed in hif-1 mutants to the lack of detoxification response, leading to the rapid intoxication of the animals. Several lines of evidence support this conclusion. First, we observed that hif-1 and the detoxification mutants displayed a stronger initial reorientation response (omega turns) and a more rapid decline in speed and reversals compared to wild type (Figure 5 D–F). Second, to test if hif-1 mutants were indeed more susceptible to H2S toxicity, we exposed WT and hif-1 animals to H2S for 30 mins and subsequently tested their ability to respond to near-UV light. Unlike WT animals, the speed response to near-UV light was inhibited in hif-1 mutants (Figure 5I), suggesting that exposure to H2S for 30 min causes a stronger toxicity in animals deficient of HIF-1 signaling. Third, hif-1 and detoxification mutants displayed a sustained high speed in response to 1% O2 , suggesting the specific impairment of H2S response. The data were presented in Lines 318–347, and were further discussed this in Lines 485–508.

To better understand whether mitochondrial damage has a role in H2S-evoked behavior, it might be useful for the authors to determine whether general ROS response pathways are important for H2S behavioral responses.

We thank the reviewer for this insightful comment. As suggested, we investigated whether ROS detoxification pathways contribute to H2S-evoked locomotory speed responses by analyzing mutants in the superoxide dismutase (SOD) family. These experiments, together with other observations, suggest that mitochondrial ROS play a dual role in H2S-evoked locomotion. The relevant results were presented in Lines 401–425, and were further discussed in Lines 509–536.

First, we found that increased mitochondrial ROS formation, either induced pharmacologically by rotenone or genetically in mitochondrial electron transport chain (ETC) mutants (Ishii et al., 2013; Ochi et al., 2016; Ramsay & Singer, 1992; Yang & Hekimi, 2010; Zorov, Juhaszova, & Sollott, 2014), suppressed the behavioral response to toxic H2S (Figure 7A–E). This indicates that mitochondrial ROS plays a significant role in H2S-evoked responses. One likely explanation is that high ROS formation may dampen the H2S-triggered ROS spike, or may impair other H2S signaling processes required to initiate avoidance. Second, consistent with previous reports (Onukwufor et al., 2022), we observed that shortterm rotenone exposure (<1 hour) significantly increased baseline locomotory speed. Given that toxic H2S levels promote ROS formation (Jia et al., 2020), our findings suggest that acute mitochondrial ROS production by toxic levels of H2S exposure may serve as a trigger for the avoidance response.

In contrast, animals with sustained mitochondrial ROS production do not have an increased baseline locomotory speed. This effect was observed after 2 hours of rotenone exposure, in mitochondrial ETC mutants, and in animals lacking all SOD enzymes (Figures 7A–K). A likely explanation for the reduced basal locomotory speed during sustained mitochondrial ROS production is the activation of ROSresponsive signaling pathways including HIF-1, NRF2/SKN-1, and DAF-16/FOXO (Lennicke & Cocheme, 2021; Patten, Germain, Kelly, & Slack, 2010), which may promote adaptation to prolonged oxidative stress (Figure 7H). Notably, unlike hif-1 mutants, SOD-deficient animals remained as responsive as WT to other stimuli after 30 minutes of H2S exposure (Figure 7—figure supplement 1H), indicating that elevated ROS levels do not compromise overall viability or the ability to detoxify H2S.

Taken together, these results support a model in which mitochondrial ROS exerts a biphasic effect on H2S-induced avoidance. It enhances detection and avoidance under acute stress but contributes to locomotory suppression when ROS levels remain elevated chronically.

Reviewer #2 (Recommendations For The Authors):

The way the manuscript is presented could be improved without much effort by rewriting/editing. For the reader, it is hard at present to understand the rationales of the experiments, why they were carried out, and how to place them into a context. This could be improved on three levels:

(1) Documentation

(2) Narration/Explanation

(3) Visualization

(1) Documentation

Not all of the results in the text are well documented. The results should be described with more details in the written text and improved documentation and quantification of the results. Example:

Turning behavior is mentioned as an important aspect of the response to H2S. There is no citation given but this effect is not well documented. The authors image the animals and could provide video footage of the effect, could quantify eg turning/pirouettes, and provide the data. At the moment the manuscript largely relies on measuring the increase in speed, but the reader is left wondering what other behavioral effects occur and how this is altered in all of the mutant and other conditions tested. Just quantifying speed reduces the readout and seems like an oversimplification to characterize the behavioral response.

We are grateful for this comment. We now provide a video footage of the H2S effects (Figure 1—Video 1). As suggested, we analyzed the recordings to extract reorientation (omega-turns) and reversals. These analyses are now included in the Supplemental file 1 with representative panels displayed in Figure 5 and supplements to Figures 2, 3, 5, 6 and 7. Even though the mutant effects on omega-turns were often subtle, and reversal responses showed considerable variability (likely due to differences in population density, food availability, or animals’ physiological state prior to the assay), this analysis has proven valuable for distinguishing mutants that exhibit adaptation from those that display hypersensitivity to H2S toxicity. For instance, although both SOD-deficient and BP-treated animals failed to increase their locomotory speed in H2S (Figure 6E and Figure 7G), they exhibited distinct omega-turn responses (Figure 6—figure supplement 1I and Figure 7—figure supplement 1F), suggesting that different mechanisms likely underlie the locomotory defects of these two animals. We have integrated the omega-turn and reversal data into the text and discussed under relevant contexts.

(2) Narration/Description.

Generally, the description of the results part is very brief and it is often not clear why a certain experiment was carried out and how. Surely it is possible to check the methods but this interrupts the flow of reading and it would be easier for the reader to be guided through the results with more information what the initial motivation for an experiment is, what the general experimental outline is, and what specific experiments are carried out.

We apologize for the lack of clarity and logical structure in the initial submission. In the revised manuscript, we have thoroughly revised the text to improve its organization and readability.

Examples:

Line 97ff: The authors performed a candidate screen yet it is not described why which genes were chosen. Are there also pathways that were tested that turned out to not be involved?

We thank the reviewer for the suggestion. To address this, we have added a new section, explaining the rationale for selecting genes and pathways in our candidate screen. Briefly, we focused on genes known or predicted to be involved in sensory responses to gaseous stimuli in C. elegans and mammals, including globins and guanylate cyclases (21% O2 sensing), potassium channels (acute hypoxia), and nutrientsensitive pathways (CO2 responses). We also included mutants defective in sensory signal transduction and neurotransmission. In addition, mitochondrial mutants were analyzed because mitochondria play a central role in H2S detoxification. The pathways that contributed to the acute H2S response included cGMP, insulin, and TGF-β signaling, as well as mitochondrial components. In contrast, globins, potassium channels, and biogenic amine signaling did not appear to play significant roles under our assay conditions. The results of the candidate screen are described in Lines 106–138 and summarized in Supplementary File 1.

line 262ff: the paragraph starts with explaining ferritin genes that are important for iron control but the reader does not yet know why. Then it is explained that a ferritin gene is DE in the H2S transcriptomes. then a motivation to look into the labile iron pool is described. Why not first explain what genes are strongly regulated and why they are selected based on their DE? Then explain what is known about these genes and pathways, and then motivate a set of experiments.

We agree with the reviewer that our initial description could have been more logically organized. We reframed this section to first present the RNA-seq data, followed by an explanation of their known biological functions and the motivation for the subsequent experiments (Lines 350–357).

nhr-49 appears suddenly in the results part and it is not clear why it was tested and how the result links. Is nhr-49 a key transcription factor that is activated by H2S sensory or physiological response, and does it control the signaling or protective changes induced by H2S?

We thank the reviewer for the comment. As suggested, we revised the text to present the information more clearly. In our candidate gene screen, a set of mutants exhibiting reduced speed responses to H2S has previously been shown to be defective in response to CO2 stimulation (Hallem & Sternberg, 2008). These included animals deficient in nutrient-sensitive pathways, including insulin, TGF-beta, and NHR49, which were reported by Sternberg’s lab to exhibit dampened responses to CO2 (Hallem & Sternberg, 2008) (Lines 173–179). We also included a simply cartoon to further illustrate this (Figure 3C).

The nuclear hormone receptor NHR-49 has been implicated in a variety of stress responses, including starvation (Van Gilst, Hadjivassiliou, & Yamamoto, 2005), bacterial pathogen (Naim et al., 2021; Wani et al., 2021), and hypoxia (Doering et al., 2022). The nhr-49 mutants exhibited a rapid decline in locomotory speed during H2S exposure, implicating a role in sustaining high speed in the presence of H2S. Furthermore, we observed that fmo-2, a well-characterized target gene of NHR-49, was significantly upregulated after 1 hour of exposure to 50 and 150 ppm H2S (Supplementary file 2), suggesting that NHR-49 signaling is rapidly activated by H2S exposure. Exactly how NHR-49 contributes to H2S response requires further investigation.

(3) Visualization

Adding a model/cartoon summary that describes the pathways tested and their interaction would be helpful in some of the figures for the reader to keep an overview of the pathways that were tested. Also, a final summary cartoon that integrates all the puzzle pieces into one larger picture would be helpful. Such a final cartoon overview could also point to the key open questions of the underlying mechanisms.

We thank the reviewer for this suggestion. We have added a series of models/cartoons to illustrate the different pathways and their interactions. These include starvation regulatory mechanisms (Figure 3C), 21% O2 sensing mechanisms (Figure 3G), HIF-1 signaling and detoxification (Figure 5—figure supplement 1E), HIF-1 signaling and the regulation of labile iron (Figure 6H), as well as ROS signaling and regulation (Figure 7L). To help interpretation and to elaborate on these models, we have also included explanatory sentences in the corresponding figure legends.

Other comments:

Introduction and line 93: The authors mention that 50 ppm H2S has beneficial effects on lifespan yet does not have a detectable phenotype." Are there any concentrations of H2S that cause attraction of C. elegans and what is the preferred range if it exists? Could this be measured in an H2S gradient?

We thank the reviewer for the insightful comment. We performed an H2S gradient assay, which suggests that wild type animals are attracted toward low concentrations of H2S around 40 ppm (Figure 1G and H) (Lines 95–104). These results suggest that H2S acts as a strong repellent for C. elegans at high concentrations but as an attractant at low levels. This dual role may be ecologically relevant, as wild C. elegans lives in complex and dynamic environments where H2S levels likely fluctuate over short distances (Adams, Farwell, Pack, & Bamesberger, 1979; Budde & Roth, 2011; Morra & Dick, 1991; Patange, Breen, Arsuffi, & Ruvkun, 2025; Rodriguez-Kabana, Jordan, & Hollis, 1965; Romanelli-Cedrez, Vairoletti, & Salinas, 2024).

Line 146: "Local H2S concentrations could also be significantly higher in decomposing substances where wild C. elegans thrives" please provide a citation.

As suggested, we included a set of references that have described the H2S enrichment in the natural environment in early field studies (Adams et al., 1979; Morra & Dick, 1991; Rodriguez-Kabana et al., 1965), as well as references that have discussed and implied this in C. elegans studies (Budde & Roth, 2011; Patange et al., 2025; Romanelli-Cedrez et al., 2024). They can be found in the introduction (Lines 59–62) and in the result (Lines 197–199).

Line 311 "Wild C. elegans isolates thrive in the decomposing matters, where the local concentrations of O2 are low while the levels of CO2 and H2S could be high. These animals have adapted their behavior in such an environment, displaying increased sensitivity to high O2 exposure but dampened responses to CO2." Please provide citations for these statements.

As suggested, we cited the relevant articles or books describing the variation of O2 and CO2 levels in the decomposing matters including several C. elegans papers that mentioned this in Lines 197–199 (Bretscher, Busch, & de Bono, 2008; Gea, Barrena, Artola, & Sanchez, 2004; Hallem & Sternberg, 2008; Oshins, Michel, Louis, Richard, & Rynk, 2022), and the above-mentioned articles for H2S (Adams et al., 1979; Budde & Roth, 2011; Morra & Dick, 1991; Patange et al., 2025; Rodriguez-Kabana et al., 1965; Romanelli-Cedrez et al., 2024).

For C. elegans’ sensitivity to O2 and CO2, these articles were cited in Lines 201–203 (Beets et al., 2020; Bretscher et al., 2008; Carrillo, Guillermin, Rengarajan, Okubo, & Hallem, 2013; Hallem & Sternberg, 2008; Kodama-Namba et al., 2013; McGrath et al., 2009).

Reviewer #3 (Recommendations For The Authors):

More work could be conducted establishing the neuronal circuitry involved in the initial, tractable response to H2S.

We thank the reviewer for the insightful comment. Since our initial analyses suggest a role of ASJ neurons in H2S-evoked locomotory responses (Figure 2F and G), We thought that this would offer us an entry point to dissect the neuronal circuit involved in H2S responses. Expression of the tetanus toxin catalytic domain in ASJ, which blocks neurosecretion, inhibited H2S evoked locomotory responses (Figure 2H), suggesting that neurosecretion from ASJ promotes the speed response to H2S (Lines 162– 165). We then performed calcium imaging of ASJ neurons in response to H2S exposure. However, while we observed CO2 -evoked calcium transients in ASJ using GCaMP6s, we did not detect any calcium response to H2S, under several conditions, including animals on food, off food, and with different H2S concentrations and exposure times (Figure2—Figure supplement 2E and 2F) (Lines 166–168). Since signaling from ASJ neurons regulates developmental programs that modify sensory functions in C. elegans, including CO2 and O2 responses (Murakami, Koga, & Ohshima, 2001), the involvement of ASJ neurons is not specific to H2S responses and ASJ neurons are unlikely to serve as a primary H2S sensor (Discussed in Line 449–458). Therefore, the exact sensory neuron, circuit and molecular triggers mediating acute H2S avoidance behavior remain to be elucidated.

Our subsequent investigation on mitochondrial components suggests that a burst of mitochondrial ROS production may be the trigger for H2S avoidance, as transient exposure to rotenone substantially increases baseline locomotory activity (Figure 7E) (Line 391–396). However, mitochondrial ROS could potentially target multiple neurons and cellular machineries to initiate avoidance behavior to H2S, making it challenging to pinpoint specific sites of action. Nevertheless, we agree that further dissection of the neural circuits and mitochondrial signaling in H2S avoidance will be important and should be explored in future studies. We discussed this in Lines 509–536.

I am not sure how to overcome the challenges involved in reaching conclusions from the decreased locomotory responses of animals that are sensitized to the effects of H2S. Perhaps this conundrum could be discussed in more detail in the text.

We thank the reviewer for this important comment. We agree that decreased locomotory speed during H2S exposure can arise from distinct causes, either systemic toxicity or adaptation, and distinguishing between these is critical. We have included new experiments and revised the text to clarify this issue.

Our data suggest that increased initial omega turns and a rapid loss of locomotion in hif-1 and detoxification-defective mutants including sqrd-1 and ethe-1 likely reflect an enhanced sensitivity to H2S toxicity due to their failure to induce appropriate adaptative responses (Figure 5D–F, Figure 5J–L, Figure 5—Figure supplement 1F–P). Supporting this, hif-1 mutants become less responsive to unrelated stimuli (near-UV light) after 30 minutes of H2S exposure (Figure 5I).

In contrast, egl-9 and SOD-deficient animals show reduced initial reorientation and reduced speed responses (Figure 5B, Figure 7G, Figure 5—Figure supplement 1A and B, and Figure 7—Figure supplement 1F and G), although both egl-9 and sod mutants respond normally to the other stimuli prior or after H2S exposure (Figure 5I, Figure 5—Figure supplement 1C, and Figure 7—Figure supplement 1H). Since disrupting egl-9 stabilizes HIF-1 and upregulates the expression of numerous genes involved in cellular defense against H2S toxicity, the enhanced detoxification capacity in egl-9 mutants likely increases animals’ tolerance to H2S, thereby reducing avoidance to otherwise toxic H2S levels. Similarly, persistently high ROS in SOD deficient animals activates a variety of stress-responsive signaling pathways, including HIF-1, NRF2/SKN-1 and DAF-16/ FOXO signaling (Lennicke & Cocheme, 2021; Patten et al., 2010), facilitating cellular adaptation to redox stress and reducing animals’ responsiveness to toxic H2S levels. Taken together, these findings support the view that reduced locomotory speed during H2S exposure can arise from distinct mechanisms: early systemic toxicity in hif-1 and detoxificationdefective mutants, versus enhanced cellular adaptation in egl-9 and SOD mutants. We have integrated the relevant information across the result section and discussed this in Lines 485–536.

References

Adams, D. F., Farwell, S. O., Pack, M. R., & Bamesberger, W. L. (1979). Preliminary Measurements of Biogenic Sulfur-Containing Gas Emissions from Soils. Journal of the Air Pollution Control Association, 29(4), 380-383. doi:Doi 10.1080/00022470.1979.10470805

Beets, I., Zhang, G., Fenk, L. A., Chen, C., Nelson, G. M., Felix, M. A., & de Bono, M. (2020). NaturaL Variation in a Dendritic Scaffold Protein Remodels Experience-Dependent Plasticity by Altering Neuropeptide Expression. Neuron, 105(1), 106-121 e110. doi:10.1016/j.neuron.2019.10.001

Bretscher, A. J., Busch, K. E., & de Bono, M. (2008). A carbon dioxide avoidance behavior is integrated with responses to ambient oxygen and food in Caenorhabditis elegans. Proc Natl Acad Sci U S A, 105(23), 8044-8049. doi:10.1073/pnas.0707607105

Budde, M. W., & Roth, M. B. (2011). The response of Caenorhabditis elegans to hydrogen sulfide and hydrogen cyanide. Genetics, 189(2), 521-532. doi:10.1534/genetics.111.129841

Carrillo, M. A., Guillermin, M. L., Rengarajan, S., Okubo, R. P., & Hallem, E. A. (2013). O-2-Sensing Neurons Control CO2 Response in C. elegans. Journal of Neuroscience, 33(23), 9675-9683. doi:10.1523/Jneurosci.4541-12.2013

Doering, K. R. S., Cheng, X., Milburn, L., Ratnappan, R., Ghazi, A., Miller, D. L., & Taubert, S. (2022). Nuclear hormone receptor NHR-49 acts in parallel with HIF-1 to promote hypoxia adaptation in Caenorhabditis elegans. Elife, 11. doi:10.7554/eLife.67911

Gea, T., Barrena, R., Artola, A., & Sanchez, A. (2004). Monitoring the biological activity of the composting process: Oxygen uptake rate (OUR), respirometric index (RI), and respiratory quotient (RQ). Biotechnol Bioeng, 88(4), 520-527. doi:10.1002/bit.20281

Hallem, E. A., & Sternberg, P. W. (2008). Acute carbon dioxide avoidance in Caenorhabditis elegans. Proc Natl Acad Sci U S A, 105(23), 8038-8043. doi:10.1073/pnas.0707469105

Ishii, T., Miyazawa, M., Onouchi, H., Yasuda, K., Hartman, P. S., & Ishii, N. (2013). Model animals for the study of oxidative stress from complex II. Biochim Biophys Acta, 1827(5), 588-597. doi:10.1016/j.bbabio.2012.10.016

Jia, J., Wang, Z., Zhang, M., Huang, C., Song, Y., Xu, F., . . . Cheng, J. (2020). SQR mediates therapeutic effects of H(2)S by targeting mitochondrial electron transport to induce mitochondrial uncoupling. Sci Adv, 6(35), eaaz5752. doi:10.1126/sciadv.aaz5752

Kodama-Namba, E., Fenk, L. A., Bretscher, A. J., Gross, E., Busch, K. E., & de Bono, M. (2013). Crossmodulation of homeostatic responses to temperature, oxygen and carbon dioxide in C. elegans. PLoS Genet, 9(12), e1004011. doi:10.1371/journal.pgen.1004011

Lennicke, C., & Cocheme, H. M. (2021). Redox metabolism: ROS as specific molecular regulators of cell signaling and function. Mol Cell, 81(18), 3691-3707. doi:10.1016/j.molcel.2021.08.018

McGrath, P. T., Rockman, M. V., Zimmer, M., Jang, H., Macosko, E. Z., Kruglyak, L., & Bargmann, C. I. (2009). Quantitative mapping of a digenic behavioral trait implicates globin variation in C. elegans sensory behaviors. Neuron, 61(5), 692-699. doi:10.1016/j.neuron.2009.02.012

Morra, M. J., & Dick, W. A. (1991). Mechanisms of h(2)s production from cysteine and cystine by microorganisms isolated from soil by selective enrichment. Appl Environ Microbiol, 57(5), 14131417. doi:10.1128/aem.57.5.1413-1417.1991

Murakami, M., Koga, M., & Ohshima, Y. (2001). DAF-7/TGF-beta expression required for the normal larval development in C-elegans is controlled by a presumed guanylyl cyclase DAF-11. Mechanisms of Development, 109(1), 27-35. doi:Doi 10.1016/S0925-4773(01)00507-X

Naim, N., Amrit, F. R. G., Ratnappan, R., DelBuono, N., Loose, J. A., & Ghazi, A. (2021). Cell nonautonomous roles of NHR-49 in promoting longevity and innate immunity. Aging Cell, 20(7). doi:ARTN e13413 10.1111/acel.13413

Ochi, R., Dhagia, V., Lakhkar, A., Patel, D., Wolin, M. S., & Gupte, S. A. (2016). Rotenone-stimulated superoxide release from mitochondrial complex I acutely augments L-type Ca2+ current in A7r5 aortic smooth muscle cells. Am J Physiol Heart Circ Physiol, 310(9), H1118-1128. doi:10.1152/ajpheart.00889.2015

Onukwufor, J. O., Farooqi, M. A., Vodickova, A., Koren, S. A., Baldzizhar, A., Berry, B. J., . . . Wojtovich, A. P. (2022). A reversible mitochondrial complex I thiol switch mediates hypoxic avoidance behavior in C. elegans. Nat Commun, 13(1), 2403. doi:10.1038/s41467-022-30169-y

Oshins, C., Michel, F., Louis, P., Richard, T. L., & Rynk, R. (2022). Chapter 3 - The composting process. In R. Rynk (Ed.), The Composting Handbook (pp. 51-101): Academic Press.

Patange, O., Breen, P., Arsuffi, G., & Ruvkun, G. (2025). Hydrogen sulfide mediates the interaction between C. elegans and Actinobacteria from its natural microbial environment. Cell Reports, 44(1), 115170. doi:10.1016/j.celrep.2024.115170

Patten, D. A., Germain, M., Kelly, M. A., & Slack, R. S. (2010). Reactive oxygen species: stuck in the middle of neurodegeneration. J Alzheimers Dis, 20 Suppl 2, S357-367. doi:10.3233/JAD-2010100498

Ramsay, R. R., & Singer, T. P. (1992). Relation of superoxide generation and lipid peroxidation to the inhibition of NADH-Q oxidoreductase by rotenone, piericidin A, and MPP+. Biochem Biophys Res Commun, 189(1), 47-52. doi:10.1016/0006-291x(92)91523-s

Rodriguez-Kabana, R., Jordan, J. W., & Hollis, J. P. (1965). Nematodes: Biological Control in Rice Fields: Role of Hydrogen Sulfide. Science, 148(3669), 524-526. doi:10.1126/science.148.3669.524

Romanelli-Cedrez, L., Vairoletti, F., & Salinas, G. (2024). Rhodoquinone-dependent electron transport chain is essential for Caenorhabditis elegans survival in hydrogen sulfide environments. J Biol Chem, 300(9), 107708. doi:10.1016/j.jbc.2024.107708

Van Gilst, M. R., Hadjivassiliou, H., & Yamamoto, K. R. (2005). A Caenorhabditis elegans nutrient response system partially dependent on nuclear receptor NHR-49. Proc Natl Acad Sci U S A, 102(38), 13496-13501. doi:10.1073/pnas.0506234102

Wani, K. A., Goswamy, D., Taubert, S., Ratnappan, R., Ghazi, A., & Irazoqui, J. E. (2021). NHR- 49/PPAR-α and HLH-30/TFEB cooperate for host defense via a flavin-containing monooxygenase. Elife, 10. doi:ARTN e62775 10.7554/eLife.62775

Yang, W., & Hekimi, S. (2010). A mitochondrial superoxide signal triggers increased longevity in Caenorhabditis elegans. PLoS Biol, 8(12), e1000556. doi:10.1371/journal.pbio.1000556

Zorov, D. B., Juhaszova, M., & Sollott, S. J. (2014). Mitochondrial reactive oxygen species (ROS) and ROS-induced ROS release. Physiol Rev, 94(3), 909-950. doi:10.1152/physrev.00026.2013

Version 2

Abstract

Abstract

Hydrogen sulfide (H2S) acts as an energy source, a toxin, and a gasotransmitter across diverse biological contexts. We use the robust locomotory responses of Caenorhabditis elegans to high levels of H2S to elucidate the molecular mechanisms underlying its acute perception and adaptive responses. We find that the H2S-evoked behavioral response is shaped by multiple environmental factors including oxygen (O2) levels and nutritional state, and is modulated by various pathways such as insulin, TGF-β, and HIF-1 signaling, as well as by input from O2-sensing neurons. Prolonged exposure to H2S activates HIF-1 signaling, leading to the upregulation of stress-responsive genes, including those involved in H2S detoxification. This promotes an adaptive state in which locomotory speed is reduced in H2S, while responsiveness to other stimuli is preserved. In mutants deficient in HIF-1 signaling, iron storage, and detoxification mechanisms, animals display a robust initial response but rapidly enter a sleep-like behavior characterized by reduced mobility and diminished responsiveness to subsequent sensory stimuli. Furthermore, while acute production of mitochondria-derived reactive O2 species (ROS) appears to initiate the avoidance response to H2S, persistently high ROS promotes an adaptive state, likely by activating various stress-response pathways, without substantially compromising cellular H2S detoxification capacity. Taken together, our study provides comprehensive molecular insights into the mechanisms through which C. elegans detects, modulates, and adapts its response to H2S exposure.

Introduction

During the late Proterozoic era, the rise in oxygen (O2) levels eliminated hydrogen sulfide (H2S) from most habitats. Yet, low O2 levels, along with high concentrations of hydrogen sulfide (H2S) and carbon dioxide (CO2), can persist in enclosed environments where bacteria actively decompose organic matter (Olson & Straub, 2016). H2S can easily permeate biological membranes and interfere with various cellular processes. One of the most detrimental effects of H2S is the disruption of cellular respiration by remodeling the mitochondrial electron transport chain and inhibiting cytochrome c oxidase (COX) (Cooper & Brown, 2008; Khan et al., 1990; Nicholls & Kim, 1982; Romanelli-Cedrez, Vairoletti, & Salinas, 2024).

The detection and response to chemical signals are crucial for various organisms to interact effectively with their environment. Animals living in H2S-rich environments have evolved mechanisms to either avoid or adapt to these conditions, conferring them survival advantages. Species that encounter periodic increases in H2S often possess sensing abilities and exhibit avoidance behaviors. For example, benthic species undertake daily vertical migrations between sulfidic and non-sulfidic waters (Abel, Koenig, & Davis, 1987; Salvanes, Utne-Palm, Currie, & Braithwaite, 2011). Certain metazoan species have seized the ecological opportunities presented by H2S-rich environments through the development of cellular adaptations. Livebearing fish species, for instance, have evolved COX subunits that are more resistant to H2S and more efficient at H2S detoxification (Kelley et al., 2016; Pfenninger et al., 2014).

Free-living nematodes feed on bacteria that thrive in decaying organic matter, such as compost. This complex and dynamic environment, where species like C. elegans are commonly found, may exhibit significant fluctuations in H2S levels over short distances (Adams, Farwell, Pack, & Bamesberger, 1979; Budde & Roth, 2011; Morra & Dick, 1991; Patange, Breen, Arsuffi, & Ruvkun, 2025; Rodriguez-Kabana, Jordan, & Hollis, 1965; Romanelli-Cedrez et al., 2024). Evidence suggests that C. elegans has evolved tolerance to prolonged H2S exposure by reprogramming gene expression through H2S-induced stabilization of the hypoxia-inducible factor HIF-1 (Budde & Roth, 2010; Powell-Coffman, 2010), and by developing an H2S-resistant electron transport chain (Romanelli-Cedrez et al., 2024). In particular, low concentrations of H2S are not only well tolerated but can also be beneficial. For instance, exposure to 50 ppm H2S has been shown to extend the lifespan and enhance thermotolerance (Budde & Roth, 2010; Fawcett, Hoyt, Johnson, & Miller, 2015; Horsman, Heinis, & Miller, 2019; Miller & Roth, 2007; Qabazard et al., 2014). Moreover, endogenous H2S production in C. elegans modulates its interactions with actinobacteria (Patange et al., 2025). However, exposure to concentrations exceeding 150 ppm proved lethal (Budde & Roth, 2010, 2011; Fawcett et al., 2015; Horsman et al., 2019; Miller, Budde, & Roth, 2011a; Miller & Roth, 2007). In addition to long-term adaptive mechanisms, C. elegans exhibits an acute behavioral response to H2S exposure (Budde & Roth, 2011). In this study, we investigated whether and how the nematode C. elegans acutely avoids environments rich in H2S, how it adapts at the molecular and physiological levels to H2S exposure, and how these adaptations alter the avoidance response.

Results

C. elegans increases locomotory activity in response to acute H2S exposure

In response to aversive cues, C. elegans initiates a pirouette followed by an increase of its locomotory speed to escape the noxious stimuli. When acutely exposed to 150 ppm H2S, the laboratory N2 strain exhibited an increased locomotory speed and enhanced turning behavior, while generally decreasing the frequency of reversals, as a strategy to escape the noxious stimuli (Figure 1A–C, Figure 1—Video 1). The maximum speed was reached after 6–8 minutes in H2S and returned to the baseline over a period of approximately 1 hour (Figure 1D). The reduced locomotory speed after 30 minutes in 150 ppm H2S was reversible when these animals were immediately challenged with another stimulus, such as near-UV light (Figure 1E), suggesting that neuromuscular function remains preserved. The magnitude and dynamics of the H2S-induced avoidance responses were concentration-dependent. Under our assay conditions, 50 ppm H2S did not elicit an acute speed increase, and responses to 75 ppm were variable. In contrast, exposure to 150 ppm H2S consistently triggered a robust and reproducible escape response (Figure 1F), consistent with earlier reports that H2S at 50 ppm exerts beneficial effects, while H2S at 150 ppm is toxic to C. elegans (Budde & Roth, 2010; Fawcett et al., 2015; Horsman et al., 2019; Miller & Roth, 2007; Qabazard et al., 2014). To investigate whether low levels of H2S might be preferable or even attractive to the animals, we exposed wild type animals to a linear gradient of H2S (Figure 1G; See methods for details) (Bretscher, Busch, & de Bono, 2008; Chang, Chronis, Karow, Marletta, & Bargmann, 2006; Gray et al., 2004). Animals were allowed to move freely on a thin layer of bacteria within a microfluidic chamber containing an H2S gradient ranging from 150 ppm at one end to 0 ppm at the other. As expected, C. elegans robustly avoids 150 ppm H2S but does not accumulate at the extreme end of 0 ppm (Figure 1G, H). The highest proportion of animals was found in the region with approximately 40 ppm H2S. These data suggest that H2S acts as a potent repellent for C. elegans at high concentrations but as an attractant at low levels.

Acute locomotory responses of C. elegans to hydrogen sulfide

(A) Changes in the locomotory speed of WT animals (N2 laboratory strain) evoked by a switch from 7% O2 to 150 ppm H2S balanced with 7% O2, followed by a return to 7% O2. (B) Proportion of WT animals undergoing reorientation movements (omega turns) during a switch from 7% O2 to 150 ppm H2S balanced with 7% O2, followed by a return to 7% O2. (C) Fraction of WT animals doing backward locomotion (reversals) during a switch from 7% O2 to 150 ppm H2S balanced with 7% O2, followed by a return to 7% O2. (D) Locomotory speed of WT was recorded for 2 minutes at 7% O2, followed by 148 minutes at 150 ppm H2S balanced with 7% O2, and then at 7% O2. (E) Locomotory speed of WT animals in response to near-UV light (435 nm, 0.7 mW/mm2) exposure, before and directly after 30-minute preincubation in 150 ppm H2S balanced with 7% O2. Locomotory activity was recorded for 2 minutes before, during and after UV light exposure. (F) Locomotory speed of WT animals to different concentrations of H2S balanced with 7% O2. Red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes, labeled a and b, respectively) used for statistical analysis. **** = p < 0.0001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test. (G) Distribution of WT animals in a microfluidic device after 25 minutes of aerotaxis. The gas inputs and the five chamber sections used for scoring are indicated. (H) Distribution of WT animals in a gradient of H2S (150 ppm to 0 ppm). The bins correspond to different sections of the microfluidic chamber described in (G). N= 4 aerotaxis assays.

Candidate gene screen identifies signaling pathways required for H2S-evoked locomotion

To investigate the molecular mechanisms underlying the H2S-evoked locomotory response in C. elegans, we conducted a candidate gene screen for mutants that failed to increase their locomotory speed upon exposure to an acute rise of H2S level to 150 ppm. Our analysis prioritized genes that are either known to be, or potentially involved in, sensory responses to gaseous stimuli in both C. elegans and mammals (Supplementary file 1). We began by examining mutants affecting globins, guanylate cyclases, and cyclic nucleotide-gated (CNG) channels, which are involved in O2 or CO2 sensing in C. elegans, as well as wild isolates and other previously characterized mutants with defective responses to gas stimuli (Supplementary file 1) (Bretscher et al., 2008; Gray et al., 2004; Hallem & Sternberg, 2008; Persson et al., 2009). Given the essential role of K+ channels in acute hypoxia sensing in mammals, we also screened all potassium (K+) channel mutants (Gao, Gonzalez-Rodriguez, Ortega-Saenz, & Lopez-Barneo, 2017; Weir, Lopez-Barneo, Buckler, & Archer, 2005). To explore how the H2S sensory signals were transmitted within the neural circuit, we assayed mutants with defects in ciliogenesis or cilia-mediated signaling, as well as animals deficient in neurotransmission involving classical neurotransmitters, neuropeptides, or biogenic amines. Finally, considering that mitochondrial function is central to H2S detoxification, we included mitochondrial electron transport chain (ETC) mutants and key factors implicated in H2S and superoxide clearance. Together, these functionally diverse categories allowed us to survey a wide spectrum of potential mechanisms contributing to the behavioral response to H2S.

In the screen, a collection of mutants displayed reduced speed response, whereas only a few exhibited a lack of omega-turn and/or reversal responses to H2S (Supplementary file 1). The reversal inhibition upon H2S exposure and its rebound following H2S withdrawal appeared to vary substantially across experiments for N2, suggesting that differences in population density, food availability, and animals’ physiological state prior to the assay may matter. Nevertheless, our screen revealed that signaling pathways, including cGMP, insulin and TGF-β signaling, contributed to H2S-evoked avoidance (Supplementary file 1). A naturally occurring variant of the neuropeptide receptor gene npr-1, previously implicated in the modulation of O2 and CO2 sensing, also affected locomotory speed in H2S (Supplementary file 1). In addition, mitochondrial components, including those involved in respiration and H2S detoxification, played critical roles in triggering and maintaining escape behavior in H2S. In contrast, we did not find evidence that globins, K+ channels, or biogenic amine signaling contributed significantly to the behavioral avoidance to H2S (Supplementary file 1).

cGMP signaling in ASJ cilia contributes to H2S avoidance

One group of mutants exhibiting reduced locomotory speed in H2S had defects in ciliogenesis (Figure 2A–C). This group included daf-19 mutants lacking cilia, dyf-3 mutants with reduced cilia length, and dyf-7 mutants in which ciliogenesis occurs but the cilia are not anchored to their sensilla and are not exposed to the external environment (Heiman & Shaham, 2009; Murayama, Toh, Ohshima, & Koga, 2005; Starich et al., 1995; Swoboda, Adler, & Thomas, 2000). These findings suggest that ciliogenesis and exposure of cilia to the external environment are essential for animals to escape H2S exposure. In addition, these mutants exhibited consistently high reversal rate, likely associated with their reduced locomotory speed (Figure 2 —figure supplement 1A–F). Cilia-mediated sensory responses often involve the activation of guanylate cyclases and the opening of cGMP-gated channels (Ferkey, Sengupta, & L’Etoile, 2021). We observed significantly attenuated locomotory speed in response to H2S in tax-4 or tax-2 mutants, which lack key subunits of cilia-enriched cGMP-gated channels (Figure 2D, E). In a comprehensive survey of all guanylate cyclase mutants, the daf-11 strain emerged as the only mutant showing speed response defects in the avoidance of H2S (Figure 2D, supplementary file 1). Similar to the cilia mutants, animals deficient in daf-11 had low speed, attenuated omega turns, and constantly high reversal rate (Figure 2D, Figure 2—figure supplement 1G, H). Selective expression of daf-11 in ASJ neurons, but not in other neurons, partially restored the speed response to acute H2S (Figures 2F, Figure 2—figure supplement 2A–D). In addition, the H2S response defect of tax-4 mutants was also rescued by selective expression of its cDNA in ASJ neurons (Figure 2G), suggesting that the receptor guanylate cyclase DAF-11 and the CNG channel TAX-4 act at least partially in ASJ neurons to promote acute avoidance to H2S. To validate the role of ASJ neurons in H2S responses, we blocked neurotransmission from ASJ using the catalytic domain of Tetanus Toxin (TeTx), which impairs neurosecretion by specifically cleaving synaptobrevin (Schiavo et al., 1992). Cell-specific expression of TeTx in ASJ significantly inhibited the speed response to H2S (Figure 2H). Surprisingly, using the Ca2+ sensor GCaMP6s to monitor a direct response of ASJ neurons to H2S, we observed no H2S-induced Ca2+ transients under any of the tested conditions, whereas CO2-evoked Ca2+ increases were readily detected (Figure 2—figure supplement 2E, F). These findings suggest that although ASJ neuronal activity and neurosecretion contribute to the H2S responses, ASJ neurons are unlikely to serve as the primary H2S sensor.

Neurosecretion from ASJ neurons contributes to H2S avoidance

(A–H) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT and daf-19(m86) (A); WT and dyf-3(m185) (B); WT and dyf-7(m539) (C); WT, daf-11(m47) and tax-4(p678) (D); WT and tax-2(p691) (E); WT, daf-11(m47) and transgenic daf-11(m47) expressing daf-11 genomic DNA in ASJ neurons (two independent lines) (F); WT, tax-4(p678) and transgenic tax-4(p678) expressing tax-4 cDNA in ASJ neurons (two independent lines) (G); WT and transgenic WT expressing the catalytic domain of the tetanus toxin (TeTx) in ASJ neurons (H). For the comparison of acute locomotory speed responses between strains, red bars on the x-axis represent two intervals (4– 5 minutes and 11–12 minutes, labeled a and b, respectively, in panels D, F, and G) used for statistical analysis. **** = p < 0.0001, *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test.

Starvation modulates H2S avoidance

In the candidate gene screen, another set of mutants with reduced speed responses to H2S had previously been shown to be defective in response to CO2 (Hallem & Sternberg, 2008). Specifically, the H2S-evoked speed response was reduced in mutants deficient in nutrient-sensitive signaling, including the insulin receptor DAF-2, the TGF-β ligand DAF-7, and the nuclear hormone receptor NHR-49 (Figure 3A–C, Figure 3—figure supplement 1A). Similarly, mutants lacking other CO2 response modulators, such as the calcineurin subunits TAX-6 and CNB-1, also failed to respond to H2S (Figure 3—figure supplement 1B, C). In addition to the abolished speed response, we also observed that the H2S-evoked omega-turn response was nearly abolished in daf-2 mutants, and high proportion of animals exhibited persistently high reversal rate regardless of H2S stimulation (Figure 3—figure supplement 1D– F). The H2S avoidance defects observed in daf-2 and daf-7 single mutants were fully suppressed in daf-2; daf-16 and daf-7; daf-3 double mutants (Figure 3A, B). Therefore, it is likely that insulin and TGF-β signaling modulate H2S responses by regulating the expression of relevant genes via DAF-16 and DAF-3 transcription factors, respectively. Given that H2S-evoked speed response was modulated by the nutrient-sensitive pathways, we further explored whether the locomotory response to H2S was sensitive to the nutrient state. Similar to the response to CO2 (Bretscher et al., 2008; Hallem & Sternberg, 2008), we observed that a 24-hour period of starvation suppressed the speed response to H2S (Figure 3D). However, the CO2 sensor GCY-9 was dispensable for H2S-evoked avoidance (Figure 3—figure supplement 1G– I). These findings suggest that even though acute avoidance of CO2 and H2S are modulated through shared mechanisms, distinct molecular sensors are used.

Insulin, TGF-β, and high O2 signaling pathways antagonizes H2S avoidance

(A, B) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT, daf-2(e1370), daf-16(mgDf47), and daf-2(e1370); daf-16(mgDf47) double mutants (A); WT, daf-7(e1372), daf-3(mgDf90), and daf-7(e1372); daf-3(mgDf90) double mutants (B). For the daf-2 assays, WT and daf-2 mutants were maintained at 15°C. L4 animals were picked and shifted to 25°C until day-one adults, then assayed at room temperature. (C) Hypothetical model for the regulation of H2S avoidance by a starvation program involving the insulin, TGF-β, and NHR-49 pathways. (D) Locomotory speed response to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 in fed and starved WT animals. (E, F) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT and wild isolates (CB4858 and CB4855) (E); WT and npr-1(ad609) (F). (G) Hypothetical model showing how O2 regulates H2S avoidance via the inhibitory RMG circuit activity, which is modulated by O2 sensory inputs and NPR-1 signaling. (H–L) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT, npr-1(ad609), and transgenic npr-1(ad609) expressing npr-1 in RMG neurons using Cre-LoxP system (Macosko et al., 2009) (H); WT, npr-1(ad609), gcy-35(ok769), and npr-1(ad609); gcy-35(ok769) double mutants (I); WT, npr-1(ad609), tax-4(p678) and npr-1(ad609); tax-4(p678) double mutants (J); WT, npr-1(ad609), qaIs2241(genetic ablation of AQR, PQR and URX neurons) and npr-1(ad609); qaIs 2241 (K); WT, npr-1(ad609), npr-1(ad609); gcy-35(ok769) double mutants and transgenic npr-1(ad609); gcy-35(ok769) double mutants expressing gcy-35 cDNA under gcy-37 promoter, which drives gcy-35 expression in O2 sensing neurons (L). For the comparison of acute locomotory responses between strains, red bars on the x-axis represent two intervals (4– 5 minutes and 11–12 minutes, labeled a and b, respectively, in panel E) used for statistical analysis. **** = p < 0.0001, *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test.

The O2 sensing circuit antagonizes H2S avoidance in wild isolates

Wild strains of C. elegans thrive in environmental niches characterized by variable O2, CO2 and H2S levels (Adams et al., 1979; Bretscher et al., 2008; Budde & Roth, 2011; Gea, Barrena, Artola, & Sanchez, 2004; Hallem & Sternberg, 2008; Morra & Dick, 1991; Oshins, Michel, Louis, Richard, & Rynk, 2022; Patange et al., 2025; Rodriguez-Kabana et al., 1965). Compared to the laboratory reference strain N2, wild isolates display robust responses to high O2 but reduced sensitivity to CO2 stimulation (Beets et al., 2020; Bretscher et al., 2008; Carrillo, Guillermin, Rengarajan, Okubo, & Hallem, 2013; Hallem & Sternberg, 2008; Kodama-Namba et al., 2013; McGrath et al., 2009). It is thought that the unique O2 and CO2 responses of N2 strain evolved during its domestication on solid media at atmospheric gas concentrations (Sterken, Snoek, Kammenga, & Andersen, 2015). We hypothesized that wild C. elegans strains might also exhibit an attenuated response to H2S, and therefore included a set of wild strains in our candidate gene screen. Consistent with this hypothesis, wild isolates including CB4855, CB4856, and CB4858 exhibited reduced locomotory speed in H2S (Figure 3E, Figure 3— figure supplement 1J).

A naturally occurring variation in the neuropeptide receptor NPR-1 predominantly determines the differences in locomotory responses to O2 and CO2 between the laboratory N2 strain and wild isolates. The NPR-1(215F) variant found in wild isolates is less active than the NPR-1(215V) found in N2 animals (de Bono & Bargmann, 1998). We wondered whether the variations in the npr-1 gene also contribute to differences in the acute response to H2S. The absence of npr-1 was sufficient to reduce the speed response to H2S, suggesting that active NPR-1 signaling promotes H2S avoidance (Figure 3F). NPR-1(215V) primarily acts within the RMG interneurons to modulate responses to environmental stimuli by inhibiting signal output from these neurons (Figure 3G) (Laurent et al., 2015; Macosko et al., 2009). Cell-specific expression of the highly active npr-1(215V) variant in RMG neurons effectively restored the H2S-evoked locomotory speed in npr-1 null mutant worms (Figure 3H). Furthermore, optogenetic stimulation of RMG interneurons using channelrhodopsin-2 (ChR2) significantly dampened the locomotory response to H2S in N2 animals, suggesting that high RMG activity inhibits H2S responses (Figure 3—figure supplement 1K).

The perception of 21% O2 involves the soluble guanylate cyclases GCY-35/GCY-36 and the downstream cyclic nucleotide-gated (CNG) channels TAX-4/TAX-2 in the URX, AQR, and PQR O2-sensing neurons, which relay the signals to RMG (Busch et al., 2012; Cheung, Cohen, Rogers, Albayram, & de Bono, 2005; Couto, Oda, Nikolaev, Soltesz, & de Bono, 2013; Gray et al., 2004; Laurent et al., 2015; Persson et al., 2009; Zimmer et al., 2009). We investigated whether disrupting the O2 sensing machinery would have the same effect as inhibiting RMG signaling pathway by NPR-1(215V). Mutating gcy-35 and tax-4, or genetically ablating the O2 sensing neurons URX, AQR, and PQR by cell-specific expression of the pro-apoptotic gene egl-1 restored H2S-evoked speed response in npr-1 null mutants (Figure 3I–K). Specific expression of gcy-35 cDNA in O2 sensing neurons was sufficient to repress this response to H2S in npr-1; gcy-35 double mutants (Figure 3L). Finally, enhancing the presynaptic activity of URX, AQR, PQR neurons by the expression of a gain-of-function PKC-1(A160E) (Dekker, Mcintyre, & Parker, 1993; Hiroki et al., 2022) significantly dampened H2S-induced locomotory speed (Figure 3—figure supplement 1L). Taken together, these observations indicate that activation of the O2-sensing circuit inhibits the H2S-evoked avoidance response.

H2S exposure reprograms gene expression in C. elegans

Wild isolates that were relatively insensitive to H2S exposure (Figure 3E, Figure 3—figure supplement 1J) likely developed sensory and physiological adaptations to persist in environments where H2S may transiently increase (Budde & Roth, 2011; Patange et al., 2025; Rodriguez-Kabana et al., 1965; Romanelli-Cedrez et al., 2024). These adaptations to H2S are mediated, at least in part, by the reprogramming of gene expression (Miller et al., 2011a). To identify the genes whose expression was induced by prolonged H2S exposure, we conducted a comparative analysis of transcriptome profiles in animals exposed to 50 ppm or 150 ppm H2S for 1 hour, 2 hours, or 12 hours. RNA-seq analysis revealed that exposure to either 50 ppm or 150 ppm H2S for one hour was sufficient to trigger significant changes in gene expression, with 518 or 304 genes showing differential expression, respectively (Figure 4A, Supplementary file 2). Genes induced by H2S exposure for 1 or 2 hours displayed considerable overlap (Figure 4B, Figure 4—figure supplement 1A). Gene Ontology (GO) analysis revealed that biological processes such as defense against bacteria and cysteine biosynthesis from L-serine were significantly enriched after H2S exposure (Figure 4C, D, Figure 4—figure supplement 1B, C). As expected, we observed a robust induction of genes involved in H2S detoxification (Figure 4E, F) (Miller, Budde, & Roth, 2011b; Niu et al., 2011; Vora et al., 2022). H2S is detoxified in mitochondria by sulfide:quinone oxidoreductase SQRD-1 to produce persulfide, which is then metabolized by sulfur dioxygenase ETHE-1 to generate sulfite. Thiosulfate transferase further oxidizes sulfite to produce thiosulfate (Hildebrandt & Grieshaber, 2008). Glutathione S-transferases (GSTs) remove accumulated sulfur molecules during H2S oxidation (Jackson, Melideo, & Jorns, 2012). Notably, gst-19 and sqrd-1 were among the most significantly upregulated genes after one or two hours of exposure to either 50 ppm or 150 ppm H2S, while ethe-1 showed a weak increase (Figure 4E, F, Figure 4—figure supplement 1D, E, Supplementary file 2).

Prolonged H2S exposure reprograms gene expression in C. elegans

(A) Venn diagram displaying the number of differentially expressed genes after 1-hour exposure to 50 ppm or 150 ppm H2S balanced with 7% O2 in WT animals (adjusted p < 1e-10). (B) Venn diagram displaying the number of differentially expressed genes after 1-, 2-, and 12-hour exposure to 150 ppm H2S balanced with 7% O2 in WT animals (adjusted p < 1e-10). (C, D) Significantly enriched GO categories for differentially expressed genes with adjusted p < 1e-10 in WT animals exposed to 50 ppm (C) or 150 ppm (D) H2S balanced with 7% O2 for 1 hour. (E, F) Volcano plots showing the differentially expressed genes with adjusted p < 1e-10 in WT animals exposed to 50 ppm (E) or 150 ppm (F) H2S balanced with 7% O2 for 1 hour. A set of genes involved in H2S detoxification, cysteine metabolism and stress response were highlighted in red.

H2S exposure is known to activate the HIF-1 pathway (Budde & Roth, 2010, 2011; Ma, Vozdek, Bhatla, & Horvitz, 2012; Miller et al., 2011a; Miller & Roth, 2007; Vora et al., 2022). In line with this, a proportion of genes with HIF-1 regulated promoters displayed increased expression after 1 or 2 hours of H2S exposure (Figure 4—figure supplement 1F). Among these were the aforementioned detoxifying enzymes GST-19, SQRD-1, and ETHE-1, as well as CYSL-1, CYSL-2, and CDO-1, which are involved in cysteine metabolism (Supplementary file 2). Consistent with earlier studies, we also observed differential expression of a set of SKN-1 targets in response to chronic H2S exposure (Figure 4—figure supplement 1G) (Horsman et al., 2019; Miller et al., 2011a; Niu et al., 2011). This included the heat shock protein genes hsp-16.2 and hsp-16.41 (Figure 4E, F, Figure 4—figure supplement 1D, E). Surprisingly, we detected a rapid increase in the expression of genes associated with intracellular H2S production, including the methanethiol oxidase-encoding gene semo-1 and the mercaptopyruvate sulfurtransferase-encoding gene mpst-3 (Figure 4E, F, Figure 4—figure supplement 1D, E) (Philipp et al., 2022; Qabazard et al., 2014). Elevated expression of 3-mercaptopyruvate sulfurtransferase has similarly been reported in sulfide spring fish, suggesting a conserved response mechanism to H2S (Kelley et al., 2016; Mathew, Schlipalius, & Ebert, 2011; Qabazard et al., 2014). However, the functional implications of increased expression of H2S synthesis genes remains unclear.

When exposed to H2S for 12 hours, the differentially expressed genes showed reduced overlap with those identified at earlier time points (Figure 4B, Figure 4—figure supplement 1A), suggesting that distinct defense strategies may be employed at various stages of H2S exposure. Furthermore, substantial overlap was observed between the sets of genes induced by 50 ppm and 150 ppm H2S after 1h or 2h exposure. However, this overlap decreased significantly after 12 hours exposure (Figure 4—figure supplement 1H–J). Additionally, we noticed that 150 ppm H2S specifically triggered the expression of the HIF-1 regulator rhy-1 and the HIF-1 target nhr-57 at all exposure time points, a response that was less pronounced at 50 ppm H2S (Figures 4E, F, Figure 4—figure supplement 1D, E, K and L). Together, these findings suggest that multiple defense mechanisms are induced to adapt to the stress imposed by H2S, which can diverge over time and at different H2S concentrations.

Locomotory speed response to H2S is modulated by HIF-1-induced detoxification

Given that HIF-1 signaling is robustly activated shortly after H2S exposure (Figure 4—figure supplement 1F) (Budde & Roth, 2010, 2011; Miller et al., 2011a), we next sought to determine how HIF-1 signaling might contribute to H2S-evoked avoidance response, and began by examining the effects of HIF-1 stabilization. Acclimating animals in 1% O2 for 12 hours or longer significantly decreased the speed response to H2S, indicating that HIF-1 stabilization inhibits the H2S-evoked avoidance behavior (Figure 5A). To confirm this observation, we examined several mutants with stabilized HIF-1. The conserved proline-4-hydroxylase PHD/EGL-9 and the von Hippel-Lindau (VHL) tumor suppressor protein are required for HIF-1 degradation (Kaelin & Ratcliffe, 2008; Semenza, 2010). Disruption of either egl-9 or vhl-1 leads to HIF-1 stabilization under normoxic conditions. Similar to the effect of prolonged hypoxia, both mutants exhibited a markedly reduced locomotory speed in H2S (Figure 5B). Interestingly, the omega-turn and reversal responses to H2S were also substantially inhibited in egl-9 mutants (Figure 5—figure supplement 1A, B). In contrast, the locomotory speed response to 1% O2 remained largely unaffected in egl-9 deficient animals (Figure 5—figure supplement 1C), suggesting a selective impairment of the H2S response. In addition, expressing the non-degradable forms of HIF-1, hif-1(P621A or P621G), under the pan-neuronal unc-14 promoter or under the endogenous hif-1 promoter, was sufficient to inhibit H2S evoked speed responses (Figure 5C). Together, these findings indicate that the activation of HIF-1 signaling during prolonged H2S or hypoxia exposure mediates an adaptive response associated with reduced behavioral responses to H2S, and that HIF-1 stabilization in neurons alone is sufficient to promote this behavioral adaptation.

Acute response to H2S is modulated by HIF-1 signaling

(A) Locomotory speed responses of WT animals to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 after incubation of the animals for 0, 12, 24, or 36 hours in 1% O2. (B, C) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT, egl-9(sa307), and vhl-1(ok161) (B); WT and transgenic animals expressing the non-degradable form of HIF-1 (P621A or P621G) under the pan-neuronal unc-14 promoter or the endogenous hif-1 promoter, respectively (C). (D–F) Avoidance responses of WT and hif-1(ia4) to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2, including locomotory speed (D), reorientation (omega-turn) (E), and reversal (F). (G, H) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT, egl-9(sa307), hif-1(ia4), and egl-9(sa307); hif-1(ia4) double mutants (G); WT, vhl-1(ok161), hif-1(ia4), and vhl-1(ok161); hif-1(ia4) double mutants (H). (I) Average locomotory speed during 2 minutes of exposure to near-UV light for WT, hif-1(ia4), and egl-9(sa307) before (Ctrl = control) and after 30 minutes of exposure to 150 ppm H2S balanced with 7% O2. (J–L) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT and sqrd-1(tm3378) (J); WT and ethe-1(yum2895) (K); WT, cysl-1(ok762), cysl-2(ok3516), and cysl-3(yum4) (L). For the comparison of acute locomotory responses between strains, red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes, labeled a and b, respectively, in panels A, B, C, and L) used for statistical analysis. *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test.

On the other hand, animals lacking HIF-1 signaling have previously been shown to display increased sensitivity to H2S and become paralyzed rapidly upon exposure (Budde & Roth, 2010; Horsman et al., 2019). Consistent with this, hif-1 mutants exhibited a brief increase in speed upon H2S exposure, which rapidly declined to baseline levels (Figure 5D). They also displayed a stronger initial omega-turn response and greater inhibition of the reversal rate than WT (Figure 5E, F), confirming that animals deficient in HIF-1 signaling are sensitized to H2S exposure. This transient avoidance behavior was also observed in egl-9; hif-1 and vhl-1; hif-1 double mutants (Figure 5G, H), suggesting that this phenotype is linked to the absence of HIF-1 signaling. Importantly, hif-1 mutants displayed a robust speed response to other stimuli, such as near-UV light and 1% O2 stimulation (Figure 5I, Figure5—figure supplement 1D), suggesting that the brief avoidance response is specific to H2S rather than reflecting a general locomotory defect. Further supporting increased sensitivity to H2S in hif-1 mutants, the locomotory speed response of hif-1 mutants to subsequent near-UV light stimulation was nearly abolished after 30 minutes of H2S exposure, whereas egl-9 mutants and WT animals remained fully responsive (Figure 5I). The reduced movement and lack of responsiveness to other stressors in hif-1 mutants resemble the stress-induced sleep-like behavior previously observed in response to noxious heat (Byrne Rodgers & Ryu, 2020). Prompted by this observation, we next explored whether HIF-1-induced detoxification genes were also required to maintain H2S-evoked behavioral responses (Figure 5—figure supplement 1E). Similar to hif-1 mutants, sqrd-1 and ethe-1 mutants displayed brief responses to H2S, with locomotory activity rapidly returning to baseline levels (Figure 5J, K, Figure 5—figure supplement 1F–I). However, both mutants responded normally to 1% O2, suggesting a specific hypersensitivity to H2S toxicity (Figure 5—figure supplement 1J). Furthermore, we examined the contribution of additional genes whose expression was upregulated in response to H2S, including the sulfhydrylase/cysteine synthase-encoding genes cysl-1, cysl-2, and cysl-3, as well as semo-1. Disrupting any of these genes phenocopied the responses observed in hif-1, ethe-1, or sqrd-1 mutants, while they displayed normal responses to 1% O2 (Figure 5L, Figure 5—figure supplement 1D, K–P). These observations suggest that while the HIF-1-induced H2S detoxification system is not essential for the initial sensing of H2S, it is required to maintain animals’ locomotory activity in the presence of H2S.

Labile iron pool sustains the locomotory activity in H2S

Among the most differentially regulated genes upon exposure to H2S, we observed consistent downregulation of ftn-1 and to a lesser extent ftn-2, whereas smf-3 expression was increased under specific H2S conditions (Figure 6A, B, Figure 6—figure supplement 1A–D). The ferritin-encoding genes ftn-1 and ftn-2, along with the iron transporter smf-3, are critical for maintaining intracellular labile iron homeostasis in C. elegans (Ackerman & Gems, 2012; Gourley, Parker, Jones, Zumbrennen, & Leibold, 2003; Rajan et al., 2019; Romney, Newman, Thacker, & Leibold, 2011), suggesting that H2S exposure may interfere with cellular iron storage. It has been shown that disrupting ftn-1 increases labile iron pools in the cytoplasm, while its overexpression reduces these levels (Anderson & Leibold, 2014; Romney et al., 2011). We found that ftn-1 mutants had enhanced speed response to H2S, whereas ftn-1 overexpression attenuated H2S-induced speed increase without damaging acute response to 1% O2 (Figure 6C, D, Figure 6—figure supplement 1E). In addition, smf-3 mutants, which impair iron uptake, exhibited a reduced speed response to H2S (Figure 6D). These results support the idea that H2S exposure disrupts iron homeostasis.

Labile iron pool sustains the locomotory activity in H2S

(A, B) Volcano plots showing the relative expression of the genes involved in the regulation of iron homeostasis in WT animals after 2-hour exposure in 50 ppm (A) or 150 ppm (B) H2S balanced with 7% O2. (C–G) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of indicated genotypes or treatments: WT and animals overexpressing ftn-1 genomic DNA under its own promoter (#1 and #2 indicate two independent lines) (C); WT, smf-3(ok1305), and ftn-1(ok3625) (D); WT, and WT pretreated with 100 μM 2,2′-Bipyridyl (BP) or with 5 mg/ml ferric ammonium citrate (FAC) in the presence of food for 16 hours (E); WT, hif-1(ia4), and hif-1(ia4) mutants pretreated with 5 mg/ml ferric ammonium citrate (FAC) for 16 hours (F); WT, hif-1(ia4), ftn-1(ok3625), and hif-1(ia4); ftn-1(ok3625) double mutants (G). (H) Hypothetical model of the interactions between labile iron pool and HIF-1 signaling. For the comparison of acute locomotory speed responses between strains or conditions, red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes, labeled a and b, respectively, in panels C, D, and E) used for statistical analysis. **** = p < 0.0001, *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann– Whitney U test.

To complement our genetic analysis, we pharmacologically depleted labile irons. In the presence of the iron chelator 2,2’-dipyridyl (BP), the H2S-evoked speed response was fully lost (Figure 6E), whereas locomotory speed in response to hypoxia was maintained (Figure 6—figure supplement 1F). We noticed that, similar to hif-1 mutants, both smf-3 mutants and BP-treated animals exhibited a strong initial omega-turn response to H2S (Figure 6—figure supplement 1G, I), suggesting that these animals are hypersensitive to H2S toxicity. One possibility is that low iron levels compromise H2S detoxification capacity, as the detoxification enzyme ETHE-1 requires iron for its activity (Kabil & Banerjee, 2012; Pettinati, Brem, McDonough, & Schofield, 2015). By contrast, increasing iron levels by feeding ferric ammonium citrate (FAC) or by disrupting ftn-1 significantly sustained animals’ locomotory speed in H2S (Figure 6D, E), likely by delaying H2S-mediated iron depletion. However, the reorientation and reversals were not significantly affected (Figure 6—figure supplement 1G– J). This impact of FAC supplementation and ftn-1 disruption on speed response was largely lost in the absence of hif-1 (Figure 6F, G), suggesting that the effects of iron require HIF-1-dependent detoxification (Figure 6H). Yet, increased iron availability modestly improved the locomotory activity of hif-1 mutants in H2S (Figure 6F, G), presumably by compensating for the low iron levels in hif-1 mutant animals (Rajan et al., 2019; Romney et al., 2011).

Mitochondrial function modulates H2S-evoked locomotion responses

The mitochondrial electron transport chain (ETC) is not only the primary target of H2S toxicity, but is also an essential component of the H2S detoxification pathways (Cooper & Brown, 2008; Khan et al., 1990; Nicholls & Kim, 1982; Romanelli-Cedrez et al., 2024). Based on this, we hypothesized that ETC plays a central role in the locomotory response to H2S. Indeed, mutating the genes critical for ETC function, including gas-1, clk-1, mev-1, and isp-1, impaired the H2S-evocked speed and reversal responses (Figure 7A–D, Figure 7— figure supplement 1A–D). These data suggest that a functional ETC is required to trigger and support H2S-evoked avoidance behavior or that dysfunctional ETC induces an adaptive response to cellular stressors. Consistently, exposing animals to rotenone, an ETC complex I inhibitor, robustly inhibited the speed responses to H2S at all time points we tested (Figure 7E). Interestingly, we noticed that transient rotenone exposure substantially increased animals’ basal locomotory speed (Figure 7E) (Onukwufor et al., 2022), suggesting that acute ETC interruption evokes behavioral avoidance, in a manner similar to that observed with other noxious stimuli. However, a 2-hour exposure to rotenone abolished the H2S-evocked speed response without increasing the basal locomotory speed (Figure 7E), suggesting that prolonged ETC inhibition may either impair the cellular capacity to support H2S-evoked behavioral responses or drive animals’ adaptation to cellular stress and consequently reduce their locomotory response to H2S.

Mitochondrial function is required for acute response to H2S

(A–D) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT and gas-1(fc21) (A); WT and clk-1(qm30) (B); WT and mev-1(kn1) (C); WT and isp-1(qm150) (D). (E) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for WT animals pretreated with 10 μM rotenone for 0 minute, 10 minutes, 30 minutes, 60 minutes or 120 minutes. For the comparison of acute locomotory speed responses between strains or conditions, red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes) used for statistical analysis. *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test. (F– G) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT, sod-1(tm776), sod-2(ok1030), and sod-3(tm760) (F); WT, sod-4(gk101), sod-5(tm1146), and sod-1(tm783); sod-2(ok1030); sod-3(tm760); sod-4(gk101); sod-5 (tm1146) (G). For the comparison of acute locomotory speed responses between strains or conditions, red bars on the x-axis represent the time interval (6–10 minutes) used for statistical analysis. *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test. (H) Hypothetical model of the role of mitochondria in response to toxic levels of H2S. In this model, mitochondria play a dual role in H2S-evoked avoidance behavior. A transient burst of mitochondrial ROS triggered by high H2S levels initiates locomotory avoidance, whereas sustained ROS elevation activates stress-responsive pathways, including HIF-1, NRF2/SKN-1, and DAF-16/FOXO, promoting adaptation to prolonged H2S exposure.

Given that rotenone exposure and mutations in ETC genes increase mitochondrial ROS generation (Ishii et al., 2013; Ochi et al., 2016; Ramsay & Singer, 1992; Yang & Hekimi, 2010; Zorov, Juhaszova, & Sollott, 2014), and that high levels of H2S also induce ROS production (Jia et al., 2020), we sought to determine how mitochondrial ROS contributes to speed response to high H2S. We focused on superoxide dismutases (SODs), key enzymes in superoxide detoxification. The C. elegans genome encodes five sod genes: sod-1 and sod-5 encode cytosolic Cu/ZnSODs, sod-2 and sod-3 encode mitochondrial MnSODs, and sod-4 encodes the extracellular Cu/ZnSOD isoforms (Doonan et al., 2008; Zubovych, Straud, & Roth, 2010). All sod single mutants exhibited a robust initial response to 150 ppm H2S (Figure 7F, G). Interestingly, the mitochondrial SOD mutants sod-2 and sod-3, as well as sod-5, showed an initial burst of locomotory activity followed by a rapid decline in speed in the presence of H2S (Figure 7F, G), but maintained a high locomotory speed in response to 1% O2 (Figure 7—figure supplement 1E). In the quintuple sod-1; sod-2; sod-3; sod-4; sod-5 mutant, which lacks all superoxide dismutases and presumably accumulates higher ROS levels, the H2S-evoked speed and omega-turn responses were nearly abolished, and the proportion of animals exhibiting reversals remained consistently high, a pattern resembling that of egl-9 and ETC mutants (Figure 7G, Figure 7—figure supplement 1F, G). Similar to sod-2 and sod-3, the quintuple sod mutant exhibited robust speed response to 1% O2 (Zhao et al., 2022). In addition, the quintuple mutants responded normally to near-UV light after 30 minutes of H2S exposure (Figure 7—figure supplement 1H), indicating that their neuromuscular function remains largely preserved. Therefore, high ROS levels in sod mutants do not substantially compromise the detoxification capacity required for survival in H2S. Instead, high ROS appears to promote adaptation to H2S-induced stress, presumably by activating several stress-responsive pathways such as HIF-1, NRF2/SKN-1, and DAF-16/FOXO (Figure 7H) (Lennicke & Cocheme, 2021; Patten, Germain, Kelly, & Slack, 2010).

Discussion

Animals’ behavior and physiology are profoundly influenced by the environments in which they evolved. The laboratory strain N2, which is adapted to low atmospheric CO2 and H2S concentrations, robustly avoids both gases (Beets et al., 2020; Bretscher et al., 2008; Carrillo et al., 2013; Hallem & Sternberg, 2008; Kodama-Namba et al., 2013; McGrath et al., 2009). Upon acute exposure to H2S above 75 ppm, N2 exhibits an avoidance behavior characterized by reorientation and increased locomotion. This response is significantly reduced in wild isolates. These divergent responses between the laboratory strain and wild isolates are primarily driven by variations in the npr-1 gene, which encodes a neuropeptide receptor, suggesting that C. elegans has undergone rapid evolutionary adaptation to its environment.

The behavioral response to a specific sensory stimulus is usually shaped by an interplay of multiple environmental and physiological cues. We observe that conditions dampening C. elegans’ response to CO2 also impair the response to H2S. Specifically, the speed response to H2S is significantly inhibited at high O2 levels, and in mutants with defective insulin or TGF-β pathway that signal starvation. However, despite these shared modulation, CO2 and H2S sensing involve distinct molecular mechanisms. For example, CO2 responses are mediated by the guanylate cyclase GCY-9 in BAG neurons (Hallem et al., 2011), which is dispensable for H2S sensing. In addition, acute exposure to H2S induces stronger reorientation but delayed acceleration compared to CO2, suggesting different sensing and information processing mechanisms (Bretscher et al., 2008). Therefore, even though responses to CO2 and H2S are modulated by overlapping regulatory mechanisms, the two gases are detected by distinct sensors and processed differently.

In a candidate gene survey, we excluded the potential involvement of globins, K+ channels, biogenic amines, and most guanylate cyclases in H2S responses. However, we observed that H2S avoidance requires the activity of guanylate cyclase DAF-11 in ASJ neurons, as well as neurosecretion from these neurons. Since the DAF-11 pathway and neurosecretion from ASJ neurons regulate developmental programs that modify sensory functions in C. elegans (Murakami, Koga, & Ohshima, 2001), it is not surprising that daf-11 mutants display pleiotropic phenotypes including impaired H2S and CO2 responses (Hallem & Sternberg, 2008). In addition, we did not detect an H2S-evoked calcium transient in ASJ neurons. Therefore, although ASJ neurons and DAF-11 activity are clearly required, the exact sensory neurons, circuits, and molecular triggers mediating acute H2S avoidance remain to be elucidated.

Consistent with previous reports (Horsman et al., 2019; Miller et al., 2011a), H2S exposure substantially alters the gene expression in C. elegans, including those involved in H2S detoxification and iron homeostasis pathways. The repression of ftn-1 and induction of smf-3 observed during H2S exposure suggest that H2S depletes intracellular labile iron. Iron is a critical cofactor required for the activity of many enzymes and supports a wide range of cellular processes (Dev & Babitt, 2017; Hentze, Muckenthaler, & Andrews, 2004). One important enzyme that requires labile iron as cofactor is PHD/EGL-9, which targets HIF-1 for degradation. Changes in iron availability are therefore expected to modulate EGL-9 activity, which in turn alter the expression of HIF-1-regulated genes (Liochev, 1996; Myllyharju & Kivirikko, 1997; Read, Bentley, Archer, & Dunham-Snary, 2021; Xu & Moller, 2011). However, ETHE-1, a key H2S detoxification enzyme upregulated upon HIF-1 stabilization, also requires iron for its persulfide dioxygenase activity (Kabil & Banerjee, 2012; Pettinati et al., 2015). Therefore, H2S-evoked responses under varying iron availability are determined by the combined effects of labile iron on HIF-1 signaling and on the H2S detoxification enzymes, among other factors. For instance, although reduced iron availability in smf-3 or BP-treated animals may promote the HIF-1 induced detoxification pathways by inactivating EGL-9, the actual enzymatic clearance of H2S is impaired due to low ETHE-1 activity. Supporting this, smf-3 mutants and BP-treated animals display increased sensitivity to H2S, including enhanced initial omega-turn responses and rapid inhibition of locomotion. In contrast, increased iron availability in ftn-1 mutants or FAC-treated animals likely delays iron depletion during H2S exposure, preserving ETHE-1 detoxification capacity and postponing the onset of HIF-1-mediated adaptations that are associated with reduced locomotion. This may explain the sustained high locomotory speed observed in ftn-1 mutants or FAC-treated animals. In the absence of HIF-1, iron supplementation only partially improves the locomotory response to H2S, likely because the H2S detoxification system including ETHE-1 cannot be transcriptionally induced. The modest effect observed may instead reflect correction of iron deficiency in hif-1 mutants.

At concentrations below 50 ppm, H2S is well tolerated or even preferred, with beneficial effects such as improved thermotolerance (Budde & Roth, 2010; Fawcett et al., 2015; Horsman et al., 2019; Miller & Roth, 2007; Qabazard et al., 2014). Notably, behavioral avoidance is absent below 50 ppm H2S, suggesting that escape behavior is triggered only when H2S is not efficiently detoxified. This led us to hypothesize that animals with enhanced detoxification capacity would show reduced avoidance of otherwise toxic H2S levels. Indeed, the speed response to H2S is attenuated when the detoxification program is upregulated under conditions of constitutive activation of HIF-1, such as in egl-9 or vhl-1 mutants, after prolonged exposure to 1% O2, or when HIF-1 signaling is specifically activated in neurons. These observations support the idea that the initial avoidance response to H2S is likely driven by acute toxicity, while the subsequent decline in speed reflects a rapid activation of adaptive mechanisms, particularly through stabilization of HIF-1. The fact that wild type animals remain responsive to other stimuli after prolonged H2S exposure suggests that reduced locomotory speed in H2S reflects active desensitization rather than general paralysis. However, persistently high H2S exposure is likely to exhaust cellular defense systems, leading to toxicity and paralysis. By contrast, the rapid loss of locomotion observed in hif-1 mutants and detoxification-defective mutants is mediated by a mechanism distinct from adaptation. Their loss of responsiveness to other stimuli after H2S exposure suggests that these animals likely become sensitized and rapidly intoxicated by H2S due to impaired detoxification. Therefore, H2S exposure promotes a behavioral program that includes an initial reorientation and acceleration responses followed by a progressive adaptation driven by cellular detoxification processes. A similar behavioral program has been observed in C. elegans during noxious heat exposure, which induces short-term heat avoidance followed by long-lasting cytoprotective adaptation and a gradual reduction in avoidance (Byrne Rodgers & Ryu, 2020).

The interaction of H2S with mitochondrial ETC is multifaceted, acting as an electron donor at low concentrations and becoming a potent toxicant at high levels (Szabo et al., 2014), in part by promoting superoxide generation through complex IV inhibition and reverse electron transport at complex I (Cooper & Brown, 2008; Jia et al., 2020; Khan et al., 1990; Nicholls & Kim, 1982; Romanelli-Cedrez et al., 2024). We propose that mitochondria play a dual role in H2S-evoked locomotory avoidance. On one hand, the mitochondrial ETC contributes to H2S detoxification and promotes adaptation. On the other hand, toxic levels of H2S remodel the ETC, leading to increased ROS production, which may serve as a trigger for the avoidance response. Supporting the idea that acute mitochondrial ROS generation initiates avoidance of high H2S levels, short-term rotenone exposure, known to promote mitochondrial ROS formation (Ochi et al., 2016; Ramsay & Singer, 1992; Zorov et al., 2014), substantially increases locomotory speed (Onukwufor et al., 2022). Meanwhile, the speed response to high H2S is fully suppressed by rotenone. This inhibition could result either from excessive mitochondrial ROS generated by rotenone, which may dampen the H2S-triggered ROS spike, or from direct complex I inhibition, which may disrupt other H2S signaling processes required to initiate avoidance. However, persistent mitochondrial ROS production appears to suppress high locomotory speed and inhibit responsiveness to H2S, as observed after 2-hour rotenone exposure, in mitochondrial ETC mutants, and in animals lacking all superoxide dismutases. One likely explanation is that mitochondrial ROS can activate a variety of stress-responsive pathways, including HIF-1, NRF2/SKN-1, and DAF-16/ FOXO signaling (Lennicke & Cocheme, 2021; Patten et al., 2010), priming animals’ adaptation to prolonged stress rather than causing toxicity. This is supported by the observation that even though SOD-deficient animals do not display strong initial locomotory responses to H2S, they remain responsive to other stimuli after 30 minutes of H2S exposure, suggesting that high ROS levels do not compromise general viability or the H2S detoxification capacity. Therefore, we favor a model in which mitochondrial ROS exert a biphasic effect on H2S-induced avoidance, facilitating H2S detection and avoidance under acute conditions, and contributing to locomotory inhibition when it is chronically elevated.

In summary, this study unveils a novel behavior of C. elegans in their avoidance to H2S. The speed response to H2S is intricately shaped by environmental context, integrating inputs from other sensory cues such as food availability and O2 levels. Although we did not identify the specific neural circuit mediating H2S signaling, our findings suggest that H2S-induced locomotion arises from acute remodeling of mitochondrial ETC and mitochondrial ROS production. We emphasize the vital role of the HIF-1-induced detoxification pathways and iron homeostasis in protecting against H2S-induced mitochondrial toxicity. In future work, it will be important to delineate the neural circuits underlying H2S-evoked locomotory avoidance and to clarify how H2S-induced changes in mitochondrial respiration and labile iron pools shape these responses.

Materials and Methods

Strains

C. elegans were maintained using standard protocols (Brenner, 1974). Strains used in this study are listed in Supplementary file 1 and Supplementary file 3.

Preparation of H2S gas mixture

Different H2S concentrations were created as previously described (Miller & Roth, 2007). Briefly, the H2S-containing gas mixture were prepared by diluting 5000 ppm H2S in nitrogen (N2) with 7% O2 balanced with N2. The gas flow was tuned by Sierra Smart-Trak 100 mass flow controllers. H2S concentration in the mixture was measured using two H2S detectors (MSA ALTAIR 2X gas detector for H2S and Clip Single Gas Detector, SDG, CROWCON). The pre-defined gas mixtures of 7% O2, 1% O2 and 5% CO2 with N2 were purchased from Air Liquide Gas AB. 5000 ppm H2S stock in N2 was obtained from Linde Gas AB. Gas mixtures in all experiments were hydrated before use.

Molecular Biology

The Multisite Gateway system (Thermo Fisher Scientific, United States) was used to generate expression vectors. Promoters, including gcy-37 (2.7kb), trx-1 (1 kb), ftn-1(2 kb), daf-11(3 kb), odr-3(2.7 kb), gpa-11(3 kb), sra-6(3 kb), odr-1(2.4 kb), flp-21(4.1kb), and ocr-2(2.4 kb) were amplified from N2 genomic DNA and cloned into pDONR P4P1 using BP clonase. gcy-35, tax-4, and pkc-1 cDNAs were amplified using the first strand cDNA library as the template, while daf-11 and ftn-1 genomic sequences were amplified using genomic DNA as the template. The entry clones of these genes pDONR 221 were generated using BP reaction. To generate the gain of function mutation of pkc-1 (A160E), Q5 Site-Directed Mutagenesis Kit (NEB) was used according to manufacturer instructions. All expression vectors were generated using LR reaction. The primer sequences were displayed in Supplementary file 4.

To generate transgenic animals, the daf-11 expression vectors were injected at the concentration of 20 ng/μl supplemented with 50 ng/μl of a coelomocyte co-injection marker (unc-122p::GFP) and 30 ng/μl of 1kb DNA ladder. For tax-4 gene, the injection mixtures were prepared using 40 ng/μl of tax-4 expression vectors, 50 ng/μl of a coelomocyte co-injection marker and 10 ng/μl of 1kb DNA ladder. The rest of expression constructs were injected at 50 ng/μl together with 50 ng/μl of a coelomocyte marker.

CRISPR/Cas9 genome editing

Genes were disrupted using CRISPR/Cas9 mediated genome editing as described (Dokshin, Ghanta, Piscopo, & Mello, 2018; Ghanta & Mello, 2020). The strategy involved the utilization of homology-directed insertion of custom designed single strand DNA template (ssODN), which had two homology arms of 35bp flanking the targeted PAM site. Between two homology arms, a short sequence containing a unique restriction enzyme cutting site as well as in-frame and out-of-frame stop codons, was included. The insertion of ssODN template would delete 16 bases of coding sequence, ensuring the proper gene disruption. To prepare the injection cocktail, 0.5 μl of Cas9 protein (IDT) was mixed with 5 μl of 0.4μg/μl tracrRNA (IDT, United States) and 2.8 μl of 0.4 μg/μl crRNA (IDT). The mixture was incubated at 37°C for at least 10 minutes before 2.2 μl of 1μg/μl ssODN (or 500 ng dsDNA) and 2 μl of 0.6 μg/μl rol-6 co-injection marker were added. Nuclease-free water was used to bring the final volume to 20 μl. The injection mixture was centrifuged for 2 minutes before use.

Behavioral assays

H2S evoked locomotion activity was monitored as described previously (Laurent et al., 2015; Zhao et al., 2022). Briefly, OP50 bacteria were seeded on the assay plates 16 hours before use. The border of bacterial lawn was removed using a PDMS stamp. For each assay, 25 to 30 day-one adult animals were picked onto assay plates, allowed to settle down for 15 minutes, and subsequently sealed within microfluidic chambers. A syringe pump (PHD 2000, Harvard Apparatus) was employed to deliver gas mixtures into the microfluidic chamber at a constant flow of 3 ml/min. The rapid switch between different gas mixtures were controlled by Telfon valves coupled with ValveBank Perfusion Controller (AutoMate Scientific). The locomotory activity at different gas mixtures were monitored using a high-resolution camera (FLIR) mounted on a Zeiss Stemi 508 microscope. Videos were captured at a rate of 2 frames per second, starting with a 2-minute recording in 7% O2, followed by 10 or 11 minutes in H2S, and ending with 7% O2. For the long-term recording in H2S, video capture commenced with a 2-minute period in 7% O2, followed by a duration of 148 minutes in 150 ppm H2S, and concluded with a final 2-minute interval in 7% O2. For each strain or condition, 3 to 4 replicates were performed.

Near-UV light experiments were conducted to assess whether animals remained responsive following a 30-minute incubation in 150 ppm H2S balanced with 7% O2. For each experiment, 6-minute videos were recorded. The first 2 minutes captured the baseline locomotion of the animals under white light, followed by 2 minutes of exposure to near-UV light (435 nm, 0.7 mW/mm2), and concluded with 2 minutes of white light. 3 to 4 replicates were performed, with approximately 15 animals recorded per replicate.

The H2S gradient experiment was performed in a PDMS chamber connected to a pump delivering gas at 0.5 ml/min. To establish a gradient ranging from 150 ppm to 0 ppm H2S, we pumped 150 ppm H2S balanced with 7% O2 into one side and 7% O2 into the opposite side. After 25 minutes, the number of animals in each of the 5 sections was counted (Figure 1G). Four replicates were performed, each using 30 day-one adult animals.

To assess the effects of rotenone on H2S evoked locomotory response, day-one adult animals were subjected to 10 μM rotenone for 10min, 30min, 1 hour and 2 hours on the drug containing plates. Subsequently, the rotenone-treated animals were assayed on the standard assay plates without the drug. To explore the impact of ferric ammonium citrate (FAC) (F5879, Sigma-Aldrich) and 2,2′-Bipyridyl (BP) (D216305, Sigma-Aldrich) on behavioral response to H2S, L4 animals were exposed to 100 μM FAC or 5 mg/ml BP for 16 hours in the presence of bacterial food. The FAC or BP treated animals were assayed on FAC or BP containing plates, respectively.

Optogenetic experiments were performed as previously outlined (Zhao et al., 2022). Transgenic L4 animals were exposed to 100 μM all-trans retinal (ATR) (R2500, Sigma) in the dark for 16 hours. The ATR-fed animals were assayed with the illumination of 70 μW/mm2 blue light during the whole assay period. Blue light was emitted from an ultra-high-power LED lamp (UHP-MIC-LED-460, Prizmatix), and the light intensity was determined using a PM50 Optical Power Meter (ThorLabs). To minimize the effect of transmitted light on the transgenic animals, a long-pass optical filter was used to eliminate the lights with short wavelengths during picking. All videos of behavioral analysis were analyzed using a home-made MATLAB program Zentracker (https://github.com/wormtracker/zentracker). For all behavior analyses, at least three assays were performed for each strain with more than 80 worms in total.

Ca2+ Imaging

L4 animals expressing the GCaMP6s sensor in ASJ neurons were picked 24 hours before imaging. The agarose pads were prepared 1 hour before the experiment. Animals were incubated in 15 µl of 10 mM levamisole for 40 minutes prior to imaging. Between 15 and 20 animals were immobilized using 10 mM levamisole diluted in a concentrated OP50 bacterial suspension and mounted on a 2% agarose pad in M9 buffer on a glass slide. The animals were then covered with a microfluidic chamber. Imaging lasted for 16 minutes. O2 (7%) was pumped into the microfluidic chamber during the first and last 4 minutes. From minute 4 to 12, a mixture of 7% O2 and 150 ppm H2S was pumped.

Imaging was performed on a Nikon AZ100 microscope equipped with a TwinCam adapter (Cairn Research, UK), fitted with two DMK 33U monochrome cameras (The Imaging Source, Germany). A 2x AZ Plan Fluor objective with 1x zoom was used, with an exposure time of 300 ms. Excitation of GCaMP and RFP was provided by a CoolLED pE-300 (CoolLED, UK). The TwinCam adapter was mounted with emission filters for green fluorescent protein (510/20 nm) and red fluorescent protein (590/35 nm), along with a DC/T510LPXRXTUf2 dichroic mirror. Imaging data were analyzed using Neuron Analyzer, a custom-written Matlab program for processing image stacks (code available at https://github.com/neuronanalyser/neuronanalyser).

RNA extraction and sequencing

To obtain RNA samples for RNA-seq, thirty young adult animals were picked to each fresh plate and allowed to lay eggs for 2 hours, after which the adult animals were removed and eggs were allowed to develop into day-one adults. Day-one animals were subsequently challenged with H2S for three different time periods (1 hour, 2 hours and 12 hours). For each time period, five plates of day-one animals were exposed to 50 ppm H2S, 150 ppm H2S or 7% O2 as a control group. Subsequently, the animals were immediately collected and rinsed three times with M9 buffer. After washing, the worm pellet was frozen in liquid nitrogen. Animals were homogenized using Bullet Blender (Next Advance) in the presence of Qiazol Lysis Reagent and 0.5mm Zirconia beads at 4°C. RNA was prepared using RNeasy Plus Universal Mini Kit (Qiagen). Six independent RNA samples were prepared for each condition. The 2100 Bioanalyzer instrument (Agilent) was used for the RNA quality control with a microfluidic chip specific for RNA (Agilent RNA 6000 Nano Kit). Then 1 μg of qualified RNA samples in RNase-free ddH2O was used for library preparation. Library was constructed and sequenced by Novogene.

RNA-seq analysis

The RNA-seq data were aligned using STAR v2.7.9a (Dobin et al., 2013) and gene expression counts extracted by featureCounts v2.0.3 (Liao, Smyth, & Shi, 2014). Differential expression was calculated using DESeq2 (Love, Huber, & Anders, 2014). GO analysis was performed using the EnrichR R package (Kuleshov et al., 2016) on genes having p.adj < 1e-10. For brevity, several highly similar GO categories were manually omitted (full EnrichR output available at Github repository).

eLife Assessment

This paper provides valuable insights into the consequences of hydrogen sulfide (H2S) exposure on the behavior and physiology of the nematode C. elegans. While solid evidence supports most of the paper's findings, the evidence that H2S is detected by the nervous system to mediate behavioral avoidance is incomplete. The paper provides a wide range of intriguing observations that could serve as a foundation for future work to synthesize these disparate results or provide insight into the mechanisms of H2S detection in C. elegans.

Reviewer #3 (Public review):

Summary:

The manuscript explores behavioral responses of C. elegans to hydrogen sulfide, which is known to exert remarkable effects on animal physiology in a range of contexts. The possibility of genetic and precise neuronal dissection of responses to H2S motivates the study of responses in C. elegans. The revised manuscript does not seem to have significantly addressed what was lacking in the initial version.

The authors have added further characterization of possible ASJ sensing of H2S by calcium imaging but ASJ does not appear to be directly involved. Genetic and parallel analysis of O2 and CO2 responsive pathways do not reveal further insights regarding potential mechanisms underlying H2S sensing. Gene expression analysis extends prior work. Finally, the authors have examined how H2S-evoked locomotory behavioral responses are affected in mutants with altered stress and detoxification response to H2S, most notably hif-1 and egl-9. These data, while examining locomotion, are more suggestive that observed effects on animal locomotion are secondary to altered organismal toxicity as opposed to specific behavioral responedse

Overall, the manuscript provides a wide range of intriguing observations, but mechanistic insight or a synthesis of disparate data is lacking.

Reviewer #4 (Public review):

Summary:

The authors establish a behavioral paradigm for avoidance of H2S and conduct a large candidate screen to identify genetic requirements. They follow up by genetically dissecting a large number of implicated pathways - insulin, TGF-beta, oxygen/HIF-1, and mitochondrial ROS, which have varied effects on H2S avoidance. They additionally assay whole-animal gene expression changes induced by varying concentrations and durations of H2S exposure.

Strengths:

The implicated pathways are tested extensively through mutants of multiple pathway molecules. The authors address previous reviewer concerns by directly testing the ability of ASJ to respond to H2S via calcium imaging. This allows the authors to revise their previous conclusion and determine that ASJ does not directly respond to H2S and likely does not initiate the behavioral response.

Weaknesses:

Despite the authors focus on acute perception of H2S, I don't think the experiments tell us much about perception. I think they indicate pathways that modulate the behavior when disrupted, especially because most manipulations used broadly affect physiology on long timescales. For instance, genetic manipulation of ASJ signaling, oxygen sensing, HIF-1 signaling, mitochondrial function, as well as starvation are all expected to constitutively alter animal physiology, which could indirectly modulate responses to H2S. The authors rule out effects on general locomotion in some cases, but other physiological changes could relatively specifically modulate the H2S response without being involved in its perception.

I am actually not convinced that H2S is directly perceived by the C. elegans nervous system at all. As far as I can tell, the avoidance behavior could be a response to H2S-induced tissue damage rather than the gas itself.

Author response:

The following is the authors’ response to the previous reviews.

Reviewer #3 (Public review):

Summary:

The manuscript explores behavioral responses of C. elegans to hydrogen sulfide, which is known to exert remarkable effects on animal physiology in a range of contexts. The possibility of genetic and precise neuronal dissection of responses to H2S motivates the study of responses in C. elegans. The revised manuscript does not seem to have significantly addressed what was lacking in the initial version.

The authors have added further characterization of possible ASJ sensing of H2S by calcium imaging but ASJ does not appear to be directly involved. Genetic and parallel analysis of O2 and CO2 responsive pathways do not reveal further insights regarding potential mechanisms underlying H2S sensing. Gene expression analysis extends prior work. Finally, the authors have examined how H2S-evoked locomotory behavioral responses are affected in mutants with altered stress and detoxification response to H2S, most notably hif-1 and egl-9. These data, while examining locomotion, are more suggestive that observed effects on animal locomotion are secondary to altered organismal toxicity as opposed to specific behavioral responedse

Overall, the manuscript provides a wide range of intriguing observations, but mechanistic insight or a synthesis of disparate data is lacking.

We thank the reviewer for the valuable feedback. We agree that while our investigation provides broad coverage, it does not fully resolve the mechanisms of H2S perception. As both reviewers noted, the avoidance response to high levels of H2S is most likely driven by its toxicity, particularly at the level of mitochondria, rather than by direct perception of H2S. We also favor this model and have revised the results and discussion to highlight this interpretation, while acknowledging that other mechanisms cannot be excluded (main changes lines 387-402 and 535-547).

Building on this view, our observations point toward mitochondrial ROS transients as the trigger for H2S avoidance. First, toxic levels of H2S are known to promote ROS production (1). Second, similar to acute H2S, brief exposure to rotenone, an ETC complex I inhibitor that rapidly generates mitochondrial ROS, triggers locomotory responses (Figure 7E) (Lines 393-396). Third, regardless of duration, rotenone exposure inhibits H2S-evoked avoidance (Figure 7E) (Lines 389-391), likely by preventing or dampening H2S-evoked mitochondrial ROS bursts when ETC function is impaired and ROS is already high. Notably, animals subjected to prolonged rotenone exposure, ETC mutants, and quintuple sod mutants, each experiencing chronically high ROS levels, fail to respond to H2S and display reduced locomotory activity, presumably due to ROS toxicity and/or activation of stress-adaptive mechanisms (Figure 7).

Consistent with the activation of stress-responsive pathways, H2S exposure alters expression of genes controlled by SKN-1 and HIF-1 signaling. Both pathways are ROS-sensitive and promote adaptation to chronic ROS production (2-4). Their activation, as in egl-9, render these animals insensitive to H2S-evoked ROS transients (Figure 5B) (Lines 303-305). Conversely, mutants defective in these adaptive pathways, such as hif-1, still show initial locomotory responses to H2S, but rapidly lose activity during prolonged H2S exposure (Figure 5D) (Lines 318-319). These observations suggest that HIF-1 pathway is dispensable for initiating the response to H2S evoked ROS transients, but essential for protecting against ROS toxicity.

In this context, the neural circuit we examined, such as ASJ neurons, is not directly involved in H2S perception (Line 165-169 and 448-457). Instead, it likely modulates a circuit that is responsive to ROS toxicity. This circuit is also influenced by ambient O2 levels, the state of O2 sensing circuit, and nutrient status, in a manner reminiscent of the CO2 responses (5, 6).

Reviewer #4 (Public review):

Summary:

The authors establish a behavioral paradigm for avoidance of H2S and conduct a large candidate screen to identify genetic requirements. They follow up by genetically dissecting a large number of implicated pathways - insulin, TGF-beta, oxygen/HIF-1, and mitochondrial ROS, which have varied effects on H2S avoidance. They additionally assay whole-animal gene expression changes induced by varying concentrations and durations of H2S exposure.

Strengths:

The implicated pathways are tested extensively through mutants of multiple pathway molecules. The authors address previous reviewer concerns by directly testing the ability of ASJ to respond to H2S via calcium imaging. This allows the authors to revise their previous conclusion and determine that ASJ does not directly respond to H2S and likely does not initiate the behavioral response.

We thank the reviewer for the supportive comments.

Weaknesses:

Despite the authors focus on acute perception of H2S, I don't think the experiments tell us much about perception. I think they indicate pathways that modulate the behavior when disrupted, especially because most manipulations used broadly affect physiology on long timescales. For instance, genetic manipulation of ASJ signaling, oxygen sensing, HIF-1 signaling, mitochondrial function, as well as starvation are all expected to constitutively alter animal physiology, which could indirectly modulate responses to H2S. The authors rule out effects on general locomotion in some cases, but other physiological changes could relatively specifically modulate the H2S response without being involved in its perception.

I am actually not convinced that H2S is directly perceived by the C. elegans nervous system at all. As far as I can tell, the avoidance behavior could be a response to H2S-induced tissue damage rather than the gas itself.

We thank the reviewer for the valuable insights, and fully agree that the H2S may not be directly perceived by C. elegans. Please see detailed responses below.

Reviewer #4 (Recommendations for the authors):

The clarity of the paper is improved in this version. My main issue has to do with "perception" of H2S. At times the authors suggest that hydrogen sulfide should be perceived by a neural circuit ("we did not specifically identify the neural circuit mediating H2S signaling"), while at other times they discuss the possibility that it is not directly perceived neuronally ("Supporting the idea that acute mitochondrial ROS generation initiates avoidance of high H2S levels,"). The authors should clearly state their model for H2S perception. Do they think there is a receptor and sensory neuron for H2S (not identified in this paper)? If not, what does it mean for there to be a neural circuit mediating the response? To me, it looks more like what is being "perceived" by a neural circuit is ROS-induced toxicity, not H2S itself.

To drill down on direct modulation of acute perception, are any of the pathway manipulations used in this paper performed on the timescale of perception? Rotenone for 10 mins is close to that timescale, and in fact it increases speed independently of H2S, consistent with ROSinduced toxicity, not H2S being the signal that induces the behavior. Optogenetic activation of RMG could also be on the acute timescale. Can the authors clarify for how long blue light was on the worms before the start of the assay? Or was it turned on at the same time as video acquisition commenced? This could be evidence that RMG acutely modulates this behavioral response.

I feel that the ASJ calcium imaging data should be in the main figure given its importance in revising the original model.

We thank the reviewer for the valuable advice.

As suggested, ASJ calcium imaging data are displayed in the main figure (Figure 2I) (Line 167).

As both reviewers noted, our initial presentation was not sufficiently clear regarding the mechanism underlying H2S avoidance. We agree with the reviewer that H2S avoidance is unlikely mediated by direct perception via a H2S-specific receptor, but likely arises from acute mitochondrial dysfunction and ROS generation.

ROS

In line with the reviewer’s perspective, our observations point toward mitochondrial ROS transients as the trigger for H2S avoidance. First, toxic levels of H2S are known to promote ROS production (1). Second, similar to acute H2S, brief exposure to rotenone, an ETC complex I inhibitor that rapidly generates mitochondrial ROS, triggers locomotory responses (Figure 7E) (Lines 393-396). Third, regardless of duration, rotenone exposure inhibits H2S-evoked avoidance (Figure 7E) (Lines 389-391), likely by preventing or dampening H2S-evoked mitochondrial ROS bursts when ETC function is impaired and ROS is already high. Notably, animals subjected to prolonged rotenone exposure, ETC mutants, and quintuple sod mutants, each experiencing chronically high ROS levels, fail to respond to H2S and display reduced locomotory activity, presumably due to ROS toxicity and/or activation of stress-adaptive mechanisms (Figure 7). We revised the Results and Discussion to present the model more consistently (main changes lines 387-402 and 535-547).

Consistent with the activation of stress-responsive pathways, H2S exposure alters expression of genes controlled by SKN-1 and HIF-1 signaling. Both pathways are ROS-sensitive and promote adaptation to chronic ROS production (2-4). Their activation, as in egl-9, render these animals insensitive to H2S-evoked ROS transients (Figure 5B) (Lines 303-305). Conversely, mutants defective in these adaptive pathways, such as hif-1, still show initial locomotory responses to H2S, but rapidly lose activity during prolonged H2S exposure (Figure 5D) (Lines 318-319). These observations suggest that HIF-1 pathway is dispensable for initiating the response to H2 Sevoked ROS transients, but essential for protecting against ROS toxicity.

ASJ neurons

ASJ neurons and DAF-11 signaling are required for H2S-evoked behavioral responses. However, ASJ does not exhibit an H2S-evoked calcium transient. It suggests that ASJ neurons do not directly detect H2S (Line 165-169 and 448-457), but likely modulate the circuit responsive to ROS toxicity. This circuit can also be modulated by ambient O2 levels, the state of O2 sensing circuit, and nutrient status, in a manner reminiscent of the CO2 responses (5, 6).

O2 sensing circuit

Consistent with the reviewer’s view, we favor the model that H2S avoidance is likely induced by ROS transients. We believe that the state of O2 sensing circuit, similar to ASJ neurons, modulates the neural circuit that is responsive to H2S-evoked ROS toxicity. This circuit is inhibited as long as O2 sensing circuit is active. In the RMG optogenetic experiment, channelrhodopsin was photo-stimulated as soon as the assay was initiated at 7% O2 (Methods Lines 633-634 and Figure legend Lines 1177-1178), therefore RMG remained active throughout the assay including at 7% O2. Our interpretation is that RMG activation inhibits this ROSresponsive circuit and H2S avoidance. However, these observations do not resolve if H2S is acutely and directly perceived. The modulation of H2S response by O2 circuit was discussed between Lines 437-447.

References

(1) J. Jia et al., SQR mediates therapeutic effects of H(2)S by targeting mitochondrial electron transport to induce mitochondrial uncoupling. Sci Adv 6, eaaz5752 (2020).

(2) S. J. Lee, A. B. Hwang, C. Kenyon, Inhibition of Respiration Extends C. elegans Life Span via Reactive Oxygen Species that Increase HIF-1 Activity. Current Biology 20, 2131-2136 (2010).

(3) C. Lennicke, H. M. Cocheme, Redox metabolism: ROS as specific molecular regulators of cell signaling and function. Mol Cell 81, 3691-3707 (2021).

(4) D. A. Patten, M. Germain, M. A. Kelly, R. S. Slack, Reactive oxygen species: stuck in the middle of neurodegeneration. J Alzheimers Dis 20 Suppl 2, S357-367 (2010).

(5) A. J. Bretscher, K. E. Busch, M. de Bono, A carbon dioxide avoidance behavior is integrated with responses to ambient oxygen and food in Caenorhabditis elegans. Proc Natl Acad Sci U S A 105, 8044-8049 (2008).

(6) E. A. Hallem, P. W. Sternberg, Acute carbon dioxide avoidance in Caenorhabditis elegans. Proc Natl Acad Sci U S A 105, 8038-8043 (2008).

Version 3

Abstract

Hydrogen sulfide (H2S) acts as an energy source, a toxin, and a gasotransmitter across diverse biological contexts. We use the robust locomotory responses of Caenorhabditis elegans to high levels of H2S to elucidate the molecular mechanisms underlying its acute and adaptive responses. We find that the H2S-evoked behavioral response is shaped by multiple environmental factors including oxygen (O2) levels and nutritional state, and is modulated by various pathways such as insulin, TGF-β, and HIF-1 signaling, as well as by input from O2-sensing neurons. Prolonged exposure to H2S activates HIF-1 signaling, leading to the upregulation of stress-responsive genes, including those involved in H2S detoxification. This promotes an adaptive state in which locomotory speed is reduced in H2S, while responsiveness to other stimuli is preserved. In mutants deficient in HIF-1 signaling, iron storage, and detoxification mechanisms, animals display a robust initial response but rapidly enter a sleep-like behavior characterized by reduced mobility and diminished responsiveness to subsequent sensory stimuli. Furthermore, while acute production of mitochondria-derived reactive O2 species (ROS) appears to initiate the avoidance response to H2S, persistently high ROS promotes an adaptive state, likely by activating various stress-response pathways, without substantially compromising cellular H2S detoxification capacity. Taken together, our study provides comprehensive molecular insights into the mechanisms through which C. elegans modulates and adapts its response to H2S exposure.

Introduction

During the late Proterozoic era, the rise in oxygen (O2) levels eliminated hydrogen sulfide (H2S) from most habitats. Yet, low O2 levels, along with high concentrations of hydrogen sulfide (H2S) and carbon dioxide (CO2), can persist in enclosed environments where bacteria actively decompose organic matter (Olson & Straub, 2016). H2S can easily permeate biological membranes and interfere with various cellular processes. One of the most detrimental effects of H2S is the disruption of cellular respiration by remodeling the mitochondrial electron transport chain and inhibiting cytochrome c oxidase (COX) (Cooper & Brown, 2008; Khan et al., 1990; Nicholls & Kim, 1982; Romanelli-Cedrez, Vairoletti, & Salinas, 2024).

The detection and response to chemical signals are crucial for various organisms to interact effectively with their environment. Animals living in H2S-rich environments have evolved mechanisms to either avoid or adapt to these conditions, conferring them survival advantages. Species that encounter periodic increases in H2S often exhibit avoidance behaviors. For example, benthic species undertake daily vertical migrations between sulfidic and non-sulfidic waters (Abel, Koenig, & Davis, 1987; Salvanes, Utne-Palm, Currie, & Braithwaite, 2011). Certain metazoan species have seized the ecological opportunities presented by H2S-rich environments through the development of cellular adaptations. Livebearing fish species, for instance, have evolved COX subunits that are more resistant to H2S and more efficient at H2S detoxification (Kelley et al., 2016; Pfenninger et al., 2014).

Free-living nematodes feed on bacteria that thrive in decaying organic matter, such as compost. This complex and dynamic environment, where species like C. elegans are commonly found, may exhibit significant fluctuations in H2S levels over short distances (Adams, Farwell, Pack, & Bamesberger, 1979; Budde & Roth, 2011; Morra & Dick, 1991; Patange, Breen, Arsuffi, & Ruvkun, 2025; Rodriguez-Kabana, Jordan, & Hollis, 1965; Romanelli-Cedrez et al., 2024). Evidence suggests that C. elegans has evolved tolerance to prolonged H2S exposure by reprogramming gene expression through H2S-induced stabilization of the hypoxia-inducible factor HIF-1 (Budde & Roth, 2010, 2011; Horsman, Heinis, & Miller, 2019; Ma, Vozdek, Bhatla, & Horvitz, 2012; Miller, Budde, & Roth, 2011; Powell-Coffman, 2010; Topalidou & Miller, 2017), and by developing an H2S-resistant electron transport chain (Romanelli-Cedrez et al., 2024). In particular, low concentrations of H2S are not only well tolerated but can also be beneficial. For instance, exposure to 50 ppm H2S has been shown to extend the lifespan and enhance thermotolerance (Fawcett, Hoyt, Johnson, & Miller, 2015; Miller & Roth, 2007). Moreover, endogenous H2S production in C. elegans modulates its interactions with actinobacteria (Patange et al., 2025). However, exposure to concentrations exceeding 150 ppm proved lethal (Budde & Roth, 2010; Horsman et al., 2019).

In addition to long-term adaptive mechanisms, C. elegans exhibits an acute behavioral response to H2S exposure (Budde & Roth, 2011). In this study, we investigated whether and how the nematode C. elegans acutely avoids environments rich in H2S, how it adapts at the molecular and physiological levels to H2S exposure, and how these adaptations alter the avoidance response.

Results

C. elegans increases locomotory activity in response to acute H2S exposure

In response to aversive cues, C. elegans initiates a pirouette followed by an increase of its locomotory speed to escape the noxious stimuli. When acutely exposed to 150 ppm H2S, the laboratory N2 strain exhibited an increased locomotory speed and enhanced turning behavior, while generally decreasing the frequency of reversals, as a strategy to escape the noxious stimuli (Figure 1A–C, Figure 1—Video 1). The maximum speed was reached after 6–8 minutes in H2S and returned to the baseline over a period of approximately 1 hour (Figure 1D). The reduced locomotory speed after 30 minutes in 150 ppm H2S was reversible when these animals were immediately challenged with another stimulus, such as near-UV light (Figure 1E), suggesting that neuromuscular function remains preserved. The magnitude and dynamics of the H2S-induced avoidance responses were concentration-dependent. Under our assay conditions, 50 ppm H2S did not elicit an acute speed increase, and responses to 75 ppm were variable. In contrast, exposure to 150 ppm H2S consistently triggered a robust and reproducible escape response (Figure 1F), consistent with earlier reports that H2S at 50 ppm exerts beneficial effects, while H2S at 150 ppm is toxic to C. elegans (Budde & Roth, 2010; Fawcett et al., 2015; Horsman et al., 2019; Miller & Roth, 2007). To investigate whether low levels of H2S might be preferable or even attractive to the animals, we exposed wild type animals to a linear gradient of H2S (Figure 1G; See methods for details) (Bretscher, Busch, & de Bono, 2008; Chang, Chronis, Karow, Marletta, & Bargmann, 2006; Gray et al., 2004). Animals were allowed to move freely on a thin layer of bacteria within a microfluidic chamber containing an H2S gradient ranging from 150 ppm at one end to 0 ppm at the other. As expected, C. elegans robustly avoids 150 ppm H2S but does not accumulate at the extreme end of 0 ppm (Figure 1G, H). The highest proportion of animals was found in the region with approximately 40 ppm H2S. These data suggest that H2S acts as a potent repellent for C. elegans at high concentrations but as an attractant at low levels.

Acute locomotory responses of C. elegans to hydrogen sulfide

(A) Changes in the locomotory speed of WT animals (N2 laboratory strain) evoked by a switch from 7% O2 to 150 ppm H2S balanced with 7% O2, followed by a return to 7% O2. (B) Proportion of WT animals undergoing reorientation movements (omega turns) during a switch from 7% O2 to 150 ppm H2S balanced with 7% O2, followed by a return to 7% O2. (C) Fraction of WT animals doing backward locomotion (reversals) during a switch from 7% O2 to 150 ppm H2S balanced with 7% O2, followed by a return to 7% O2. (D) Locomotory speed of WT was recorded for 2 minutes at 7% O2, followed by 148 minutes at 150 ppm H2S balanced with 7% O2, and then at 7% O2. (E) Locomotory speed of WT animals in response to near-UV light (435 nm, 0.7 mW/mm2) exposure, before and directly after 30-minute preincubation in 150 ppm H2S balanced with 7% O2. Locomotory activity was recorded for 2 minutes before, during and after UV light exposure. (F) Locomotory speed of WT animals to different concentrations of H2S balanced with 7% O2. Red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes, labeled a and b, respectively) used for statistical analysis. **** = p < 0.0001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test. (G) Distribution of WT animals in a microfluidic device after 25 minutes of aerotaxis. The gas inputs and the five chamber sections used for scoring are indicated. (H) Distribution of WT animals in a gradient of H2S (150 ppm to 0 ppm). The bins correspond to different sections of the microfluidic chamber described in (G). N= 4 aerotaxis assays.

Candidate gene screen identifies signaling pathways required for H2S-evoked locomotion

To investigate the molecular mechanisms underlying the H2S-evoked locomotory response in C. elegans, we conducted a candidate gene screen for mutants that failed to increase their locomotory speed upon exposure to an acute rise of H2S level to 150 ppm. Our analysis prioritized genes that are either known to be, or potentially involved in, sensory responses to gaseous stimuli in both C. elegans and mammals (Supplementary file 1). We began by examining mutants affecting globins, guanylate cyclases, and cyclic nucleotide-gated (CNG) channels, which are involved in O2 or CO2 sensing in C. elegans, as well as wild isolates and other previously characterized mutants with defective responses to gas stimuli (Supplementary file 1) (Bretscher et al., 2008; Gray et al., 2004; Hallem & Sternberg, 2008; Persson et al., 2009). Given the essential role of K+ channels in acute hypoxia sensing in mammals, we also screened all potassium (K+) channel mutants (Gao, Gonzalez-Rodriguez, Ortega-Saenz, & Lopez-Barneo, 2017; Weir, Lopez-Barneo, Buckler, & Archer, 2005). To explore the neuronal components involved in H2S-evoked avoidance, we assayed mutants with defects in ciliogenesis or cilia-mediated signaling, as well as animals deficient in neurotransmission involving classical neurotransmitters, neuropeptides, or biogenic amines. Finally, considering that mitochondrial function is central to H2S detoxification, we included mitochondrial electron transport chain (ETC) mutants and key factors implicated in H2S and superoxide clearance. Together, these functionally diverse categories allowed us to survey a wide spectrum of potential mechanisms contributing to the behavioral response to H2S.

In the screen, a collection of mutants displayed reduced speed response, whereas only a few exhibited a lack of omega-turn and/or reversal responses to H2S (Supplementary file 1). The reversal inhibition upon H2S exposure and its rebound following H2S withdrawal appeared to vary substantially across experiments for N2, suggesting that differences in population density, food availability, and animals’ physiological state prior to the assay may matter. Nevertheless, our screen revealed that signaling pathways, including cGMP, insulin and TGF-β signaling, contributed to H2S-evoked avoidance (Supplementary file 1). A naturally occurring variant of the neuropeptide receptor gene npr-1, previously implicated in the modulation of O2 and CO2 responses, also affected locomotory speed in H2S (Supplementary file 1). In addition, mitochondrial components, including those involved in respiration and H2S detoxification, played critical roles in triggering and maintaining escape behavior in H2S. In contrast, we did not find evidence that globins, K+ channels, or biogenic amine signaling contributed significantly to the behavioral avoidance to H2S (Supplementary file 1).

cGMP signaling in ASJ cilia contributes to H2S avoidance

One group of mutants exhibiting reduced locomotory speed in H2S had defects in ciliogenesis (Figure 2A–C). This group included daf-19 mutants lacking cilia, dyf-3 mutants with reduced cilia length, and dyf-7 mutants in which ciliogenesis occurs but the cilia are not anchored to their sensilla and are not exposed to the external environment (Heiman & Shaham, 2009; Murayama, Toh, Ohshima, & Koga, 2005; Starich et al., 1995; Swoboda, Adler, & Thomas, 2000). These findings suggest that ciliogenesis and exposure of cilia to the external environment are essential for animals to escape H2S exposure. In addition, these mutants exhibited consistently high reversal rate, likely associated with their reduced locomotory speed (Figure 2—figure supplement 1A–F). Cilia-mediated sensory responses often involve the activation of guanylate cyclases and the opening of cGMP-gated channels (Ferkey, Sengupta, & L’Etoile, 2021). We observed significantly attenuated locomotory speed in response to H2S in tax-4 or tax-2 mutants, which lack key subunits of cilia-enriched cGMP-gated channels (Figure 2D, E). In a comprehensive survey of all guanylate cyclase mutants, the daf-11 strain emerged as the only mutant showing speed response defects in the avoidance of H2S (Figure 2D, supplementary file 1). Similar to the cilia mutants, animals deficient in daf-11 had low speed, attenuated omega turns, and constantly high reversal rate (Figure 2D, Figure 2—figure supplement 1G, H). Selective expression of daf-11 in ASJ neurons, but not in other neurons, partially restored the speed response to acute H2S (Figures 2F, Figure 2—figure supplement 2A–D). In addition, the H2S response defect of tax-4 mutants was also rescued by selective expression of its cDNA in ASJ neurons (Figure 2G), suggesting that the receptor guanylate cyclase DAF-11 and the CNG channel TAX-4 act at least partially in ASJ neurons to promote acute avoidance to H2S. To validate the role of ASJ neurons in H2S responses, we blocked neurotransmission from ASJ using the catalytic domain of Tetanus Toxin (TeTx), which impairs neurosecretion by specifically cleaving synaptobrevin (Schiavo et al., 1992). Cell-specific expression of TeTx in ASJ significantly inhibited the speed response to H2S (Figure 2H). Surprisingly, using the Ca2+ sensor GCaMP6s to monitor a direct response of ASJ neurons to H2S, we observed no H2S-induced Ca2+ transients under any of the tested conditions, whereas CO2-evoked Ca2+ increases were readily detected (Figure 2I, Figure 2—figure supplement 2E). These findings suggest that although ASJ neuronal activity and neurosecretion contribute to the H2S responses, ASJ neurons are unlikely to play a role in directly detecting H2S.

Neurosecretion from ASJ neurons contributes to H2S avoidance

(A–H) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT and daf-19(m86) (A); WT and dyf-3(m185) (B); WT and dyf-7(m539) (C); WT, daf-11(m47) and tax-4(p678) (D); WT and tax-2(p691) (E); WT, daf-11(m47) and transgenic daf-11(m47) expressing daf-11 genomic DNA in ASJ neurons (two independent lines) (F); WT, tax-4(p678) and transgenic tax-4(p678) expressing tax-4 cDNA in ASJ neurons (two independent lines) (G); WT and transgenic WT expressing the catalytic domain of the tetanus toxin (TeTx) in ASJ neurons (H). For the comparison of acute locomotory speed responses between strains, red bars on the x-axis represent two intervals (4– 5 minutes and 11–12 minutes, labeled a and b, respectively, in panels D, F, and G) used for statistical analysis. **** = p < 0.0001, *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test. (I) Calcium transients evoked in ASJ neurons of WT animals in response to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals (N=42). The slow decline in GCaMP6s fluorescence is due to photobleaching.

Starvation modulates H2S avoidance

In the candidate gene screen, another set of mutants with reduced speed responses to H2S had previously been shown to be defective in response to CO2 (Hallem & Sternberg, 2008). Specifically, the H2S-evoked speed response was reduced in mutants deficient in nutrient-sensitive signaling, including the insulin receptor DAF-2, the TGF-β ligand DAF-7, and the nuclear hormone receptor NHR-49 (Figure 3A–C, Figure 3—figure supplement 1A). Similarly, mutants lacking other CO2 response modulators, such as the calcineurin subunits TAX-6 and CNB-1, also failed to respond to H2S (Figure 3—figure supplement 1B, C). In addition to the abolished speed response, we also observed that the H2S-evoked omega-turn response was nearly abolished in daf-2 mutants, and high proportion of animals exhibited persistently high reversal rate regardless of H2S stimulation (Figure 3—figure supplement 1D–F). The H2S avoidance defects observed in daf-2 and daf-7 single mutants were fully suppressed in daf-2; daf-16 and daf-7; daf-3 double mutants (Figure 3A, B). Therefore, it is likely that insulin and TGF-β signaling modulate H2S responses by regulating the expression of relevant genes via DAF-16 and DAF-3 transcription factors, respectively. Given that H2S-evoked speed response was modulated by the nutrient-sensitive pathways, we further explored whether the locomotory response to H2S was sensitive to the nutrient state. Similar to the response to CO2 (Bretscher et al., 2008; Hallem & Sternberg, 2008), we observed that a 24-hour period of starvation suppressed the speed response to H2S (Figure 3D). However, the CO2 sensor GCY-9 was dispensable for H2S-evoked avoidance (Figure 3—figure supplement 1G–I). These findings suggest that while acute avoidance of CO2 and H2S share common modulatory pathways, the responses are initiated through distinct mechanisms.

Insulin, TGF-β, and high O2 signaling pathways antagonizes H2S avoidance

(A, B) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT, daf-2(e1370), daf-16(mgDf47), and daf-2(e1370); daf-16(mgDf47) double mutants (A); WT, daf-7(e1372), daf-3(mgDf90), and daf-7(e1372); daf-3(mgDf90) double mutants (B). For the daf-2 assays, WT and daf-2 mutants were maintained at 15°C. L4 animals were picked and shifted to 25°C until day-one adults, then assayed at room temperature. (C) Hypothetical model for the regulation of H2S avoidance by a starvation program involving the insulin, TGF-β, and NHR-49 pathways. (D) Locomotory speed response to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 in fed and starved WT animals. (E, F) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT and wild isolates (CB4858 and CB4855) (E); WT and npr-1(ad609) (F). (G) Hypothetical model showing how O2 regulates H2S avoidance via the inhibitory RMG circuit activity, which is modulated by O2 sensory inputs and NPR-1 signaling. (H–L) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT, npr-1(ad609), and transgenic npr-1(ad609) expressing npr-1 in RMG neurons using Cre-LoxP system (Macosko et al., 2009) (H); WT, npr-1(ad609), gcy-35(ok769), and npr-1(ad609); gcy-35(ok769) double mutants (I); WT, npr-1(ad609), tax-4(p678) and npr-1(ad609); tax-4(p678) double mutants (J); WT, npr-1(ad609), qaIs2241(genetic ablation of AQR, PQR and URX neurons) and npr-1(ad609); qaIs 2241 (K); WT, npr-1(ad609), npr-1(ad609); gcy-35(ok769) double mutants and transgenic npr-1(ad609); gcy-35(ok769) double mutants expressing gcy-35 cDNA under gcy-37 promoter, which drives gcy-35 expression in O2 sensing neurons (L). For the comparison of acute locomotory responses between strains, red bars on the x-axis represent two intervals (4– 5 minutes and 11–12 minutes, labeled a and b, respectively, in panel E) used for statistical analysis. **** = p < 0.0001, *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test.

The O2 sensing circuit antagonizes H2S avoidance in wild isolates

Wild strains of C. elegans thrive in environmental niches characterized by variable O2, CO2 and H2S levels (Adams et al., 1979; Bretscher et al., 2008; Budde & Roth, 2011; Gea, Barrena, Artola, & Sanchez, 2004; Hallem & Sternberg, 2008; Morra & Dick, 1991; Oshins, Michel, Louis, Richard, & Rynk, 2022; Patange et al., 2025; Rodriguez-Kabana et al., 1965). Compared to the laboratory reference strain N2, wild isolates display robust responses to high O2 but reduced sensitivity to CO2 stimulation (Beets et al., 2020; Bretscher et al., 2008; Carrillo, Guillermin, Rengarajan, Okubo, & Hallem, 2013; Hallem & Sternberg, 2008; Kodama-Namba et al., 2013; McGrath et al., 2009). It is thought that the unique O2 and CO2 responses of N2 strain evolved during its domestication on solid media at atmospheric gas concentrations (Sterken, Snoek, Kammenga, & Andersen, 2015). We hypothesized that wild C. elegans strains might also exhibit an attenuated response to H2S, and therefore included a set of wild strains in our candidate gene screen. Consistent with this hypothesis, wild isolates including CB4855, CB4856, and CB4858 exhibited reduced locomotory speed in H2S (Figure 3E, Figure 3— figure supplement 1J).

A naturally occurring variation in the neuropeptide receptor NPR-1 predominantly determines the differences in locomotory responses to O2 and CO2 between the laboratory N2 strain and wild isolates. The NPR-1(215F) variant found in wild isolates is less active than the NPR-1(215V) found in N2 animals (de Bono & Bargmann, 1998). We wondered whether the variations in the npr-1 gene also contribute to differences in the acute response to H2S. The absence of npr-1 was sufficient to reduce the speed response to H2S, suggesting that active NPR-1 signaling promotes H2S avoidance (Figure 3F). NPR-1(215V) primarily acts within the RMG interneurons to modulate responses to environmental stimuli by inhibiting signal output from these neurons (Figure 3G) (Laurent et al., 2015; Macosko et al., 2009). Cell-specific expression of the highly active npr-1(215V) variant in RMG neurons effectively restored the H2S-evoked locomotory speed in npr-1 null mutant worms (Figure 3H). Furthermore, optogenetic stimulation of RMG interneurons using channelrhodopsin-2 (ChR2) significantly dampened the locomotory response to H2S in N2 animals, suggesting that high RMG activity inhibits H2S responses (Figure 3—figure supplement 1K).

The perception of 21% O2 involves the soluble guanylate cyclases GCY-35/GCY-36 and the downstream cyclic nucleotide-gated (CNG) channels TAX-4/TAX-2 in the URX, AQR, and PQR O2-sensing neurons, which relay the signals to RMG (Busch et al., 2012; Cheung, Cohen, Rogers, Albayram, & de Bono, 2005; Couto, Oda, Nikolaev, Soltesz, & de Bono, 2013; Gray et al., 2004; Laurent et al., 2015; Persson et al., 2009; Zimmer et al., 2009). We investigated whether disrupting the O2 sensing machinery would have the same effect as inhibiting RMG signaling pathway by NPR-1(215V). Mutating gcy-35 and tax-4, or genetically ablating the O2 sensing neurons URX, AQR, and PQR by cell-specific expression of the pro-apoptotic gene egl-1 restored H2S-evoked speed response in npr-1 null mutants (Figure 3I–K). Specific expression of gcy-35 cDNA in O2 sensing neurons was sufficient to repress this response to H2S in npr-1; gcy-35 double mutants (Figure 3L). Finally, enhancing the presynaptic activity of URX, AQR, PQR neurons by the expression of a gain-of-function PKC-1(A160E) (Dekker, Mcintyre, & Parker, 1993; Hiroki et al., 2022) significantly dampened H2S-induced locomotory speed (Figure 3—figure supplement 1L). Taken together, these observations indicate that activation of the O2-sensing circuit inhibits the neural pathway required for H2S-evoked avoidance response.

H2S exposure reprograms gene expression in C. elegans

Wild isolates that were relatively insensitive to H2S exposure (Figure 3E, Figure 3—figure supplement 1J) likely evolved adaptations to persist in environments where H2S levels may transiently increase (Budde & Roth, 2011; Patange et al., 2025; Rodriguez-Kabana et al., 1965; Romanelli-Cedrez et al., 2024). These adaptations to H2S are mediated, at least in part, by the reprogramming of gene expression (Miller et al., 2011). To identify the genes whose expression was induced by prolonged H2S exposure, we conducted a comparative analysis of transcriptome profiles in animals exposed to 50 ppm or 150 ppm H2S for 1 hour, 2 hours, or 12 hours. RNA-seq analysis revealed that exposure to either 50 ppm or 150 ppm H2S for one hour was sufficient to trigger significant changes in gene expression, with 518 or 304 genes showing differential expression, respectively (Figure 4A, Supplementary file 2). Genes induced by H2S exposure for 1 or 2 hours displayed considerable overlap (Figure 4B, Figure 4—figure supplement 1A). Gene Ontology (GO) analysis revealed that biological processes such as defense against bacteria and cysteine biosynthesis from L-serine were significantly enriched after H2S exposure (Figure 4C, D, Figure 4—figure supplement 1B, C). As expected, we observed a robust induction of genes involved in H2S detoxification (Figure 4E, F) (Horsman & Miller, 2016; Miller et al., 2011; Niu et al., 2011; Vora et al., 2022). H2S is detoxified in mitochondria by sulfide:quinone oxidoreductase SQRD-1 to produce persulfide, which is then metabolized by sulfur dioxygenase ETHE-1 to generate sulfite. Thiosulfate transferase further oxidizes sulfite to produce thiosulfate (Hildebrandt & Grieshaber, 2008). Glutathione S-transferases (GSTs) remove accumulated sulfur molecules during H2S oxidation (Jackson, Melideo, & Jorns, 2012). Notably, gst-19 and sqrd-1 were among the most significantly upregulated genes after one or two hours of exposure to either 50 ppm or 150 ppm H2S, while ethe-1 showed a weak increase (Figure 4E, F, Figure 4—figure supplement 1D, E, Supplementary file 2).

Prolonged H2S exposure reprograms gene expression in C. elegans

(A) Venn diagram displaying the number of differentially expressed genes after 1-hour exposure to 50 ppm or 150 ppm H2S balanced with 7% O2 in WT animals (adjusted p < 1e-10). (B) Venn diagram displaying the number of differentially expressed genes after 1-, 2-, and 12-hour exposure to 150 ppm H2S balanced with 7% O2 in WT animals (adjusted p < 1e-10). (C, D) Significantly enriched GO categories for differentially expressed genes with adjusted p < 1e-10 in WT animals exposed to 50 ppm (C) or 150 ppm (D) H2S balanced with 7% O2 for 1 hour. (E, F) Volcano plots showing the differentially expressed genes with adjusted p < 1e-10 in WT animals exposed to 50 ppm (E) or 150 ppm (F) H2S balanced with 7% O2 for 1 hour. A set of genes involved in H2S detoxification, cysteine metabolism and stress response were highlighted in red.

H2S exposure is known to activate the HIF-1 pathway (Budde & Roth, 2010, 2011; Horsman et al., 2019; Ma et al., 2012; Miller et al., 2011; Powell-Coffman, 2010; Topalidou & Miller, 2017). In line with this, a proportion of genes with HIF-1 regulated promoters displayed increased expression after 1 or 2 hours of H2S exposure (Figure 4—figure supplement 1F). Among these were the aforementioned detoxifying enzymes GST-19, SQRD-1, and ETHE-1, as well as CYSL-1, CYSL-2, and CDO-1, which are involved in cysteine metabolism (Supplementary file 2). Consistent with earlier studies, we also observed differential expression of a set of SKN-1 targets in response to chronic H2S exposure (Figure 4—figure supplement 1G) (Horsman et al., 2019; Miller et al., 2011; Niu et al., 2011). This included the heat shock protein genes hsp-16.2 and hsp-16.41 (Figure 4E, F, Figure 4—figure supplement 1D, E). Surprisingly, we detected a rapid increase in the expression of genes associated with intracellular H2S production, including the methanethiol oxidase-encoding gene semo-1 and the mercaptopyruvate sulfurtransferase-encoding gene mpst-3 (Figure 4E, F, Figure 4—figure supplement 1D, E) (Philipp et al., 2022; Qabazard et al., 2014). Elevated expression of 3-mercaptopyruvate sulfurtransferase has similarly been reported in sulfide spring fish, suggesting a conserved response mechanism to H2S (Kelley et al., 2016; Mathew, Schlipalius, & Ebert, 2011; Qabazard et al., 2014). However, the functional implications of increased expression of H2S synthesis genes remains unclear.

When exposed to H2S for 12 hours, the differentially expressed genes showed reduced overlap with those identified at earlier time points (Figure 4B, Figure 4—figure supplement 1A), suggesting that distinct defense strategies may be employed at various stages of H2S exposure. Furthermore, substantial overlap was observed between the sets of genes induced by 50 ppm and 150 ppm H2S after 1h or 2h exposure. However, this overlap decreased significantly after 12 hours exposure (Figure 4—figure supplement 1H–J). Additionally, we noticed that 150 ppm H2S specifically triggered the expression of the HIF-1 regulator rhy-1 and the HIF-1 target nhr-57 at all exposure time points, a response that was less pronounced at 50 ppm H2S (Figures 4E, F, Figure 4—figure supplement 1D, E, K and L). Together, these findings suggest that multiple defense mechanisms are induced to adapt to the stress imposed by H2S, which can diverge over time and at different H2S concentrations.

Locomotory speed response to H2S is modulated by HIF-1-induced detoxification

Given that HIF-1 signaling is robustly activated shortly after H2S exposure (Figure 4—figure supplement 1F) (Budde & Roth, 2010, 2011; Miller et al., 2011), we next sought to determine how HIF-1 signaling might contribute to H2S-evoked avoidance response, and began by examining the effects of HIF-1 stabilization. Acclimating animals in 1% O2 for 12 hours or longer significantly decreased the speed response to H2S, indicating that HIF-1 stabilization inhibits the H2S-evoked avoidance behavior (Figure 5A). To confirm this observation, we examined several mutants with stabilized HIF-1. The conserved proline-4-hydroxylase PHD/EGL-9 and the von Hippel-Lindau (VHL) tumor suppressor protein are required for HIF-1 degradation (Kaelin & Ratcliffe, 2008; Semenza, 2010). Disruption of either egl-9 or vhl-1 leads to HIF-1 stabilization under normoxic conditions. Similar to the effect of prolonged hypoxia, both mutants exhibited a markedly reduced locomotory speed in H2S (Figure 5B). Interestingly, the omega-turn and reversal responses to H2S were also substantially inhibited in egl-9 mutants (Figure 5—figure supplement 1A, B). In contrast, the locomotory speed response to 1% O2 remained largely unaffected in egl-9 deficient animals (Figure 5—figure supplement 1C), suggesting a selective impairment of the H2S response. In addition, expressing the non-degradable forms of HIF-1, hif-1(P621A or P621G), under the pan-neuronal unc-14 promoter or under the endogenous hif-1 promoter, was sufficient to inhibit H2S evoked speed responses (Figure 5C). Together, these findings indicate that the activation of HIF-1 signaling during prolonged H2S or hypoxia exposure mediates an adaptive response associated with reduced behavioral responses to H2S, and that HIF-1 stabilization in neurons alone is sufficient to promote this behavioral adaptation.

Acute response to H2S is modulated by HIF-1 signaling

(A) Locomotory speed responses of WT animals to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 after incubation of the animals for 0, 12, 24, or 36 hours in 1% O2. (B, C) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT, egl-9(sa307), and vhl-1(ok161) (B); WT and transgenic animals expressing the non-degradable form of HIF-1 (P621A or P621G) under the pan-neuronal unc-14 promoter or the endogenous hif-1 promoter, respectively (C). (D–F) Avoidance responses of WT and hif-1(ia4) to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2, including locomotory speed (D), reorientation (omega-turn) (E), and reversal (F). (G, H) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT, egl-9(sa307), hif-1(ia4), and egl-9(sa307); hif-1(ia4) double mutants (G); WT, vhl-1(ok161), hif-1(ia4), and vhl-1(ok161); hif-1(ia4) double mutants (H). (I) Average locomotory speed during 2 minutes of exposure to near-UV light for WT, hif-1(ia4), and egl-9(sa307) before (Ctrl = control) and after 30 minutes of exposure to 150 ppm H2S balanced with 7% O2. (J–L) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT and sqrd-1(tm3378) (J); WT and ethe-1(yum2895) (K); WT, cysl-1(ok762), cysl-2(ok3516), and cysl-3(yum4) (L). For the comparison of acute locomotory responses between strains, red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes, labeled a and b, respectively, in panels A, B, C, and L) used for statistical analysis. *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test.

On the other hand, animals lacking HIF-1 signaling have previously been shown to display increased sensitivity to H2S and become paralyzed rapidly upon exposure (Budde & Roth, 2010; Horsman et al., 2019). Consistent with this, hif-1 mutants exhibited a brief increase in speed upon H2S exposure, which rapidly declined to baseline levels (Figure 5D). They also displayed a stronger initial omega-turn response and greater inhibition of the reversal rate than WT (Figure 5E, F), confirming that animals deficient in HIF-1 signaling are sensitized to H2S exposure. This transient avoidance behavior was also observed in egl-9; hif-1 and vhl-1; hif-1 double mutants (Figure 5G, H), suggesting that this phenotype is linked to the absence of HIF-1 signaling. Importantly, hif-1 mutants displayed a robust speed response to other stimuli, such as near-UV light and 1% O2 stimulation (Figure 5I, Figure5—figure supplement 1D), suggesting that the brief avoidance response is specific to H2S rather than reflecting a general locomotory defect. Further supporting increased sensitivity to H2S in hif-1 mutants, the locomotory speed response of hif-1 mutants to subsequent near-UV light stimulation was nearly abolished after 30 minutes of H2S exposure, whereas egl-9 mutants and WT animals remained fully responsive (Figure 5I). The reduced movement and lack of responsiveness to other stressors in hif-1 mutants resemble the stress-induced sleep-like behavior previously observed in response to noxious heat (Byrne Rodgers & Ryu, 2020). Prompted by this observation, we next explored whether HIF-1-induced detoxification genes were also required to maintain H2S-evoked behavioral responses (Figure 5—figure supplement 1E). Similar to hif-1 mutants, sqrd-1 and ethe-1 mutants displayed brief responses to H2S, with locomotory activity rapidly returning to baseline levels (Figure 5J, K, Figure 5—figure supplement 1F–I). However, both mutants responded normally to 1% O2, suggesting a specific hypersensitivity to H2S toxicity (Figure 5—figure supplement 1J). Furthermore, we examined the contribution of additional genes whose expression was upregulated in response to H2S, including the sulfhydrylase/cysteine synthase-encoding genes cysl-1, cysl-2, and cysl-3, as well as semo-1. Disrupting any of these genes phenocopied the responses observed in hif-1, ethe-1, or sqrd-1 mutants, while they displayed normal responses to 1% O2 (Figure 5L, Figure 5—figure supplement 1D, K–P). These observations suggest that while the HIF-1-induced H2S detoxification system is not essential for initiating the response to H2S, it is required to maintain animals’ locomotory activity in the presence of H2S.

Labile iron pool sustains the locomotory activity in H2S

Among the most differentially regulated genes upon exposure to H2S, we observed consistent downregulation of ftn-1 and to a lesser extent ftn-2, whereas smf-3 expression was increased under specific H2S conditions (Figure 6A, B, Figure 6—figure supplement 1A–D). The ferritin-encoding genes ftn-1 and ftn-2, along with the iron transporter smf-3, are critical for maintaining intracellular labile iron homeostasis in C. elegans (Ackerman & Gems, 2012; Gourley, Parker, Jones, Zumbrennen, & Leibold, 2003; Rajan et al., 2019; Romney, Newman, Thacker, & Leibold, 2011), suggesting that H2S exposure may interfere with cellular iron storage. It has been shown that disrupting ftn-1 increases labile iron pools in the cytoplasm, while its overexpression reduces these levels (Anderson & Leibold, 2014; Romney et al., 2011). We found that ftn-1 mutants had enhanced speed response to H2S, whereas ftn-1 overexpression attenuated H2S-induced speed increase without damaging acute response to 1% O2 (Figure 6C, D, Figure 6—figure supplement 1E). In addition, smf-3 mutants, which impair iron uptake, exhibited a reduced speed response to H2S (Figure 6D). These results support the idea that H2S exposure disrupts iron homeostasis.

Labile iron pool sustains the locomotory activity in H2S

(A, B) Volcano plots showing the relative expression of the genes involved in the regulation of iron homeostasis in WT animals after 2-hour exposure in 50 ppm (A) or 150 ppm (B) H2S balanced with 7% O2. (C–G) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of indicated genotypes or treatments: WT and animals overexpressing ftn-1 genomic DNA under its own promoter (#1 and #2 indicate two independent lines) (C); WT, smf-3(ok1305), and ftn-1(ok3625) (D); WT, and WT pretreated with 100 μM 2,2′-Bipyridyl (BP) or with 5 mg/ml ferric ammonium citrate (FAC) in the presence of food for 16 hours (E); WT, hif-1(ia4), and hif-1(ia4) mutants pretreated with 5 mg/ml ferric ammonium citrate (FAC) for 16 hours (F); WT, hif-1(ia4), ftn-1(ok3625), and hif-1(ia4); ftn-1(ok3625) double mutants (G). (H) Hypothetical model of the interactions between labile iron pool and HIF-1 signaling. For the comparison of acute locomotory speed responses between strains or conditions, red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes, labeled a and b, respectively, in panels C, D, and E) used for statistical analysis. **** = p < 0.0001, *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann– Whitney U test.

To complement our genetic analysis, we pharmacologically depleted labile irons. In the presence of the iron chelator 2,2’-dipyridyl (BP), the H2S-evoked speed response was fully lost (Figure 6E), whereas locomotory speed in response to hypoxia was maintained (Figure 6—figure supplement 1F). We noticed that, similar to hif-1 mutants, both smf-3 mutants and BP-treated animals exhibited a strong initial omega-turn response to H2S (Figure 6—figure supplement 1G, I), suggesting that these animals are hypersensitive to H2S toxicity. One possibility is that low iron levels compromise H2S detoxification capacity, as the detoxification enzyme ETHE-1 requires iron for its activity (Kabil & Banerjee, 2012; Pettinati, Brem, McDonough, & Schofield, 2015). By contrast, increasing iron levels by feeding ferric ammonium citrate (FAC) or by disrupting ftn-1 significantly sustained animals’ locomotory speed in H2S (Figure 6D, E), likely by delaying H2S-mediated iron depletion. However, the reorientation and reversals were not significantly affected (Figure 6—figure supplement 1G– J). This impact of FAC supplementation and ftn-1 disruption on speed response was largely lost in the absence of hif-1 (Figure 6F, G), suggesting that the effects of iron require HIF-1-dependent detoxification (Figure 6H). Yet, increased iron availability modestly improved the locomotory activity of hif-1 mutants in H2S (Figure 6F, G), presumably by compensating for the low iron levels in hif-1 mutant animals (Rajan et al., 2019; Romney et al., 2011).

Mitochondrial function modulates H2S-evoked locomotion responses

The mitochondrial electron transport chain (ETC) is not only the primary target of H2S toxicity, but is also an essential component of the H2S detoxification pathways (Cooper & Brown, 2008; Khan et al., 1990; Nicholls & Kim, 1982; Romanelli-Cedrez et al., 2024). Based on this, we hypothesized that ETC plays a central role in the locomotory response to H2S. Indeed, mutating the genes critical for ETC function, including gas-1, clk-1, mev-1, and isp-1, impaired the H2S-evocked speed and reversal responses (Figure 7A–D, Figure 7— figure supplement 1A–D). These data suggest either that dysfunctional ETC induces an adaptive response to cellular stressors or that a functional ETC is required to trigger and support H2S-evoked avoidance behavior. Consistent with the later model, exposing animals to rotenone, an ETC complex I inhibitor, robustly inhibited the speed responses to H2S at all time points we tested (Figure 7E). Similar to ETC mutants, WT animals exposed to rotenone for 2 hours also showed reduced basal locomotory activity (Figure 7E).

Mitochondrial function is required for acute response to H2S

(A–D) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT and gas-1(fc21) (A); WT and clk-1(qm30) (B); WT and mev-1(kn1) (C); WT and isp-1(qm150) (D). (E) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for WT animals pretreated with 10 μM rotenone for 0 minute, 10 minutes, 30 minutes, 60 minutes or 120 minutes. For the comparison of acute locomotory speed responses between strains or conditions, red bars on the x-axis represent two intervals (4–5 minutes and 11–12 minutes) used for statistical analysis. *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test. (F– G) Locomotory speed responses to a switch from 7% O2 to 150 ppm H2S balanced with 7% O2 for animals of the indicated genotype: WT, sod-1(tm776), sod-2(ok1030), and sod-3(tm760) (F); WT, sod-4(gk101), sod-5(tm1146), and sod-1(tm783); sod-2(ok1030); sod-3(tm760); sod-4(gk101); sod-5 (tm1146) (G). For the comparison of acute locomotory speed responses between strains or conditions, red bars on the x-axis represent the time interval (6–10 minutes) used for statistical analysis. *** = p < 0.001, ** = p < 0.01, * = p < 0.05, ns = not significant, Mann–Whitney U test. (H) Hypothetical model of the role of mitochondria in response to toxic levels of H2S. In this model, mitochondria play a dual role in H2S-evoked avoidance behavior. A transient burst of mitochondrial ROS triggered by high H2S levels initiates locomotory avoidance, whereas sustained ROS elevation activates stress-responsive pathways, including HIF-1, NRF2/SKN-1, and DAF-16/FOXO, promoting adaptation to prolonged H2S exposure.

In contrast to prolonged rotenone exposure, we noticed that transient rotenone exposure substantially increased animals’ basal locomotory speed (Figure 7E) (Onukwufor et al., 2022), suggesting that acute ETC interruption evokes behavioral avoidance, in a manner similar to that observed with other noxious stimuli. Since rotenone rapidly and persistently stimulates mitochondrial ROS production (Ochi et al., 2016; Ramsay & Singer, 1992; Zorov, Juhaszova, & Sollott, 2014), the transient locomotory increase during short rotenone exposure (Figure 7E) is likely driven by these ROS bursts, whereas reduced locomotory activity during prolonged exposure is presumably caused by chronically elevated ROS production. As toxic levels of H2S also induce ROS (Jia et al., 2020), this raised the possibility that H2S-evoked behavioral responses were modulated by the mitochondrial ROS. Therefore, we sought to further explore how mitochondrial ROS contributes to speed response to high H2S. We focused on superoxide dismutases (SODs), key enzymes in superoxide detoxification. The C. elegans genome encodes five sod genes: sod-1 and sod-5 encode cytosolic Cu/ZnSODs, sod-2 and sod-3 encode mitochondrial MnSODs, and sod-4 encodes the extracellular Cu/ZnSOD isoforms (Doonan et al., 2008; Zubovych, Straud, & Roth, 2010). All sod single mutants exhibited a robust initial response to 150 ppm H2S (Figure 7F, G). Interestingly, the mitochondrial SOD mutants sod-2 and sod-3, as well as sod-5, showed an initial burst of locomotory activity followed by a rapid decline in speed in the presence of H2S (Figure 7F, G), but maintained a high locomotory speed in response to 1% O2 (Figure 7—figure supplement 1E). In the quintuple sod-1; sod-2; sod-3; sod-4; sod-5 mutant, which lacks all superoxide dismutases and presumably accumulates higher ROS levels, the H2S-evoked speed and omega-turn responses were nearly abolished, and the proportion of animals exhibiting reversals remained consistently high, a pattern resembling that of egl-9 and ETC mutants (Figure 7G, Figure 7—figure supplement 1F, G). Similar to sod-2 and sod-3, the quintuple sod mutant exhibited robust speed response to 1% O2 (Zhao et al., 2022). In addition, the quintuple mutants responded normally to near-UV light after 30 minutes of H2S exposure (Figure 7—figure supplement 1H), indicating that their neuromuscular function remains largely preserved. These observations suggest that either constitutively high ROS levels in quintuple mutants dampen H2S-evoked ROS transients, or they promote adaptation to H2S-induced stress, presumably via activation of stress-responsive pathways such as HIF-1, NRF2/SKN-1, and DAF-16/FOXO, ultimately reducing the locomotory responses (Figure 7H) (Lee, Hwang, & Kenyon, 2010; Lennicke & Cocheme, 2021; Patten, Germain, Kelly, & Slack, 2010).

Discussion

Animals’ behavior and physiology are profoundly influenced by the environments in which they evolved. The laboratory strain N2, which is adapted to low atmospheric CO2 and H2S concentrations, robustly avoids both gases (Beets et al., 2020; Bretscher et al., 2008; Carrillo et al., 2013; Hallem & Sternberg, 2008; Kodama-Namba et al., 2013; McGrath et al., 2009). Upon acute exposure to H2S above 75 ppm, N2 exhibits an avoidance behavior characterized by reorientation and increased locomotion. This response is significantly reduced in wild isolates. These divergent responses between the laboratory strain and wild isolates are primarily driven by variations in the npr-1 gene, which encodes a neuropeptide receptor, suggesting that C. elegans has undergone rapid evolutionary adaptation to its environment.

The behavioral response to a specific sensory stimulus is usually shaped by an interplay of multiple environmental and physiological cues. We observe that conditions dampening C. elegans’ response to CO2 also impair the response to H2S. Specifically, the speed response to H2S is significantly inhibited at high O2 levels, and in mutants with defective insulin or TGF-β pathway that signal starvation. However, despite these shared modulations, CO2 and H2S responses involve distinct molecular mechanisms. For example, CO2 responses are mediated by the guanylate cyclase GCY-9 in BAG neurons (Hallem et al., 2011), which is dispensable for H2S responses. In addition, acute exposure to H2S induces a delayed acceleration compared to CO2. These data suggest that while responses to the two gases share a downstream neural circuit that can be dynamically modulated by the state of O2 sensing circuit, they are triggered by distinct mechanisms.

In a candidate gene survey, we excluded the potential involvement of globins, K+ channels, biogenic amines, and most guanylate cyclases in H2S responses. However, we observed that H2S avoidance requires the activity of guanylate cyclase DAF-11 in ASJ neurons, as well as neurosecretion from these neurons. Since the DAF-11 pathway and neurosecretion from ASJ neurons regulate developmental programs that modify sensory functions in C. elegans (Murakami, Koga, & Ohshima, 2001), it is not surprising that daf-11 mutants display pleiotropic phenotypes including impaired H2S and CO2 responses (Hallem & Sternberg, 2008). In addition, we did not detect an H2S-evoked calcium transient in ASJ neurons. Therefore, although ASJ neurons and DAF-11 activity are clearly required, how H2S avoidance is triggered remains to be elucidated.

Consistent with previous reports (Horsman et al., 2019; Miller et al., 2011), H2S exposure substantially alters the gene expression in C. elegans, including those involved in H2S detoxification and iron homeostasis pathways. The repression of ftn-1 and induction of smf-3 observed during H2S exposure suggest that H2S depletes intracellular labile iron. Iron is a critical cofactor required for the activity of many enzymes and supports a wide range of cellular processes (Dev & Babitt, 2017; Hentze, Muckenthaler, & Andrews, 2004). One important enzyme that requires labile iron as cofactor is PHD/EGL-9, which targets HIF-1 for degradation. Changes in iron availability are therefore expected to modulate EGL-9 activity, which in turn alter the expression of HIF-1-regulated genes (Liochev, 1996; Myllyharju & Kivirikko, 1997; Read, Bentley, Archer, & Dunham-Snary, 2021; Xu & Moller, 2011). However, ETHE-1, a key H2S detoxification enzyme upregulated upon HIF-1 stabilization, also requires iron for its persulfide dioxygenase activity (Kabil & Banerjee, 2012; Pettinati et al., 2015). Therefore, H2S-evoked responses under varying iron availability are determined by the combined effects of labile iron on HIF-1 signaling and on the H2S detoxification enzymes, among other factors. For instance, although reduced iron availability in smf-3 or BP-treated animals may promote the HIF-1 induced detoxification pathways by inactivating EGL-9, the actual enzymatic clearance of H2S is impaired due to low ETHE-1 activity. Supporting this, smf-3 mutants and BP-treated animals display increased sensitivity to H2S, including enhanced initial omega-turn responses and rapid inhibition of locomotion. In contrast, increased iron availability in ftn-1 mutants or FAC-treated animals likely delays iron depletion during H2S exposure, preserving ETHE-1 detoxification capacity and postponing the onset of HIF-1-mediated adaptations that are associated with reduced locomotion. This may explain the sustained high locomotory speed observed in ftn-1 mutants or FAC-treated animals. In the absence of HIF-1, iron supplementation only partially improves the locomotory response to H2S, likely because the H2S detoxification system including ETHE-1 cannot be transcriptionally induced. The modest effect observed may instead reflect correction of iron deficiency in hif-1 mutants.

At concentrations below 50 ppm, H2S is well tolerated or even preferred, with beneficial effects such as improved thermotolerance (Fawcett et al., 2015; Miller & Roth, 2007; Qabazard et al., 2014). Notably, behavioral avoidance is absent below 50 ppm H2S, suggesting that escape behavior is triggered only when H2S is not efficiently detoxified. This led us to hypothesize that animals with enhanced detoxification capacity would show reduced avoidance of otherwise toxic H2S levels. Indeed, the speed response to H2S is attenuated when the detoxification program is upregulated under conditions of constitutive activation of HIF-1, such as in egl-9 or vhl-1 mutants, after prolonged exposure to 1% O2, or when HIF-1 signaling is specifically activated in neurons. These observations support the idea that the initial avoidance response to H2S is triggered by neuronal detection of acute toxicity, while the subsequent decline in speed reflects a rapid activation of adaptive mechanisms, particularly through stabilization of HIF-1. The fact that wild type animals remain responsive to other stimuli after prolonged H2S exposure suggests that reduced locomotory speed in H2S reflects active desensitization rather than general paralysis. However, persistently high H2S exposure is likely to exhaust cellular defense systems, leading to toxicity and paralysis. By contrast, the rapid loss of locomotion observed in hif-1 mutants and detoxification-defective mutants is mediated by a mechanism distinct from adaptation. Their loss of responsiveness to other stimuli after H2S exposure suggests that these animals likely become sensitized and rapidly intoxicated by H2S due to impaired detoxification. Therefore, H2S exposure promotes a behavioral program that includes an initial reorientation and acceleration responses followed by a progressive adaptation driven by cellular detoxification processes. A similar behavioral program has been observed in C. elegans during noxious heat exposure, which induces short-term heat avoidance followed by long-lasting cytoprotective adaptation and a gradual reduction in avoidance (Byrne Rodgers & Ryu, 2020).

The interaction of H2S with mitochondrial ETC is multifaceted, acting as an electron donor at low concentrations and becoming a potent toxicant at high levels (Szabo et al., 2014), in part by promoting superoxide generation through complex IV inhibition and reverse electron transport at complex I (Cooper & Brown, 2008; Jia et al., 2020; Khan et al., 1990; Nicholls & Kim, 1982; Romanelli-Cedrez et al., 2024). We propose that mitochondria play a dual role in H2S-evoked locomotory avoidance. On one hand, the mitochondrial ETC contributes to H2S detoxification and promotes adaptation. On the other hand, toxic levels of H2S remodel the ETC, leading to increased ROS production, which may serve as a trigger for the avoidance response. Supporting the idea that acute mitochondrial ROS generation initiates avoidance of high H2S levels, short-term rotenone exposure, known to promote mitochondrial ROS formation (Ochi et al., 2016; Ramsay & Singer, 1992; Zorov et al., 2014), substantially increases locomotory speed (Onukwufor et al., 2022). Meanwhile, the speed response to high H2S is fully suppressed by rotenone. This inhibition could result either from excessive mitochondrial ROS generated by rotenone, which may dampen the H2S-triggered ROS spike, or from direct complex I inhibition, which may disrupt other H2S signaling processes required to initiate avoidance. However, persistent mitochondrial ROS production appears to suppress high locomotory speed and inhibit responsiveness to H2S, as observed after 2-hour rotenone exposure, in mitochondrial ETC mutants, and in animals lacking all superoxide dismutases. One likely explanation is that mitochondrial ROS can activate a variety of stress-responsive pathways, including HIF-1, NRF2/SKN-1, and DAF-16/ FOXO signaling (Lee et al., 2010; Lennicke & Cocheme, 2021; Patten et al., 2010), priming animals’ adaptation to prolonged stress rather than causing toxicity. This is supported by the observation that even though SOD-deficient animals do not display strong initial locomotory responses to H2S, they remain responsive to other stimuli after 30 minutes of H2S exposure, suggesting that high ROS levels do not compromise general viability or the H2S detoxification capacity. Therefore, we favor a model in which mitochondrial ROS exert a biphasic effect on H2S-induced avoidance, facilitating H2S avoidance under acute conditions, and contributing to locomotory inhibition when it is chronically elevated. Overall, lack of a clear sensory machinery, the slow increase of locomotory speed in H2S (Figure 1D), the rotenone-evoked speed responses, and the strong modulation of H2S responses by mitochondrial ETC inhibition suggest that H2S may not be directly perceived by C. elegans. Instead, acute avoidance to H2S is likely initiated by ROS-induced toxicity.

In summary, this study unveils a novel behavior of C. elegans in their avoidance to H2S. The speed response to H2S is intricately shaped by environmental context, integrating inputs from other sensory cues such as food availability and O2 levels. Our findings suggest that H2S-induced locomotion arises from acute remodeling of mitochondrial ETC and associated ROS production, while highlighting the vital role of the HIF-1-induced detoxification pathways and iron homeostasis in protecting against H2S-induced mitochondrial toxicity. Further work is needed to validate this model and to elucidate the neural circuits mediating the behavioral response to ROS-induced toxicity.

Materials and Methods

Strains

C. elegans were maintained using standard protocols (Brenner, 1974). Strains used in this study are listed in Supplementary file 1 and Supplementary file 3.

Preparation of H2S gas mixture

Different H2S concentrations were created as previously described (Miller & Roth, 2007). Briefly, the H2S-containing gas mixture were prepared by diluting 5000 ppm H2S in nitrogen (N2) with 7% O2 balanced with N2. The gas flow was tuned by Sierra Smart-Trak 100 mass flow controllers. H2S concentration in the mixture was measured using two H2S detectors (MSA ALTAIR 2X gas detector for H2S and Clip Single Gas Detector, SDG, CROWCON). The pre-defined gas mixtures of 7% O2, 1% O2 and 5% CO2 with N2 were purchased from Air Liquide Gas AB. 5000 ppm H2S stock in N2 was obtained from Linde Gas AB. Gas mixtures in all experiments were hydrated before use.

Molecular Biology

The Multisite Gateway system (Thermo Fisher Scientific, United States) was used to generate expression vectors. Promoters, including gcy-37 (2.7kb), trx-1 (1 kb), ftn-1(2 kb), daf-11(3 kb), odr-3(2.7 kb), gpa-11(3 kb), sra-6(3 kb), odr-1(2.4 kb), flp-21(4.1kb), and ocr-2(2.4 kb) were amplified from N2 genomic DNA and cloned into pDONR P4P1 using BP clonase. gcy-35, tax-4, and pkc-1 cDNAs were amplified using the first strand cDNA library as the template, while daf-11 and ftn-1 genomic sequences were amplified using genomic DNA as the template. The entry clones of these genes pDONR 221 were generated using BP reaction. To generate the gain of function mutation of pkc-1 (A160E), Q5 Site-Directed Mutagenesis Kit (NEB) was used according to manufacturer instructions. All expression vectors were generated using LR reaction. The primer sequences were displayed in Supplementary file 4.

To generate transgenic animals, the daf-11 expression vectors were injected at the concentration of 20 ng/μl supplemented with 50 ng/μl of a coelomocyte co-injection marker (unc-122p::GFP) and 30 ng/μl of 1kb DNA ladder. For tax-4 gene, the injection mixtures were prepared using 40 ng/μl of tax-4 expression vectors, 50 ng/μl of a coelomocyte co-injection marker and 10 ng/μl of 1kb DNA ladder. The rest of expression constructs were injected at 50 ng/μl together with 50 ng/μl of a coelomocyte marker.

CRISPR/Cas9 genome editing

Genes were disrupted using CRISPR/Cas9 mediated genome editing as described (Dokshin, Ghanta, Piscopo, & Mello, 2018; Ghanta & Mello, 2020). The strategy involved the utilization of homology-directed insertion of custom designed single strand DNA template (ssODN), which had two homology arms of 35bp flanking the targeted PAM site. Between two homology arms, a short sequence containing a unique restriction enzyme cutting site as well as in-frame and out-of-frame stop codons, was included. The insertion of ssODN template would delete 16 bases of coding sequence, ensuring the proper gene disruption. To prepare the injection cocktail, 0.5 μl of Cas9 protein (IDT) was mixed with 5 μl of 0.4μg/μl tracrRNA (IDT, United States) and 2.8 μl of 0.4 μg/μl crRNA (IDT). The mixture was incubated at 37°C for at least 10 minutes before 2.2 μl of 1μg/μl ssODN (or 500 ng dsDNA) and 2 μl of 0.6 μg/μl rol-6 co-injection marker were added. Nuclease-free water was used to bring the final volume to 20 μl. The injection mixture was centrifuged for 2 minutes before use.

Behavioral assays

H2S evoked locomotion activity was monitored as described previously (Laurent et al., 2015; Zhao et al., 2022). Briefly, OP50 bacteria were seeded on the assay plates 16 hours before use. The border of bacterial lawn was removed using a PDMS stamp. For each assay, 25 to 30 day-one adult animals were picked onto assay plates, allowed to settle down for 15 minutes, and subsequently sealed within microfluidic chambers. A syringe pump (PHD 2000, Harvard Apparatus) was employed to deliver gas mixtures into the microfluidic chamber at a constant flow of 3 ml/min. The rapid switch between different gas mixtures were controlled by Telfon valves coupled with ValveBank Perfusion Controller (AutoMate Scientific). The locomotory activity at different gas mixtures were monitored using a high-resolution camera (FLIR) mounted on a Zeiss Stemi 508 microscope. Videos were captured at a rate of 2 frames per second, starting with a 2-minute recording in 7% O2, followed by 10 or 11 minutes in H2S, and ending with 7% O2. For the long-term recording in H2S, video capture commenced with a 2-minute period in 7% O2, followed by a duration of 148 minutes in 150 ppm H2S, and concluded with a final 2-minute interval in 7% O2. For each strain or condition, 3 to 4 replicates were performed.

Near-UV light experiments were conducted to assess whether animals remained responsive following a 30-minute incubation in 150 ppm H2S balanced with 7% O2. For each experiment, 6-minute videos were recorded. The first 2 minutes captured the baseline locomotion of the animals under white light, followed by 2 minutes of exposure to near-UV light (435 nm, 0.7 mW/mm2), and concluded with 2 minutes of white light. 3 to 4 replicates were performed, with approximately 15 animals recorded per replicate.

The H2S gradient experiment was performed in a PDMS chamber connected to a pump delivering gas at 0.5 ml/min. To establish a gradient ranging from 150 ppm to 0 ppm H2S, we pumped 150 ppm H2S balanced with 7% O2 into one side and 7% O2 into the opposite side. After 25 minutes, the number of animals in each of the 5 sections was counted (Figure 1G). Four replicates were performed, each using 30 day-one adult animals.

To assess the effects of rotenone on H2S evoked locomotory response, day-one adult animals were subjected to 10 μM rotenone for 10min, 30min, 1 hour and 2 hours on the drug containing plates. Subsequently, the rotenone-treated animals were assayed on the standard assay plates without the drug. To explore the impact of ferric ammonium citrate (FAC) (F5879, Sigma-Aldrich) and 2,2′-Bipyridyl (BP) (D216305, Sigma-Aldrich) on behavioral response to H2S, L4 animals were exposed to 100 μM FAC or 5 mg/ml BP for 16 hours in the presence of bacterial food. The FAC or BP treated animals were assayed on FAC or BP containing plates, respectively.

Optogenetic experiments were performed as previously outlined (Zhao et al., 2022). Transgenic L4 animals were exposed to 100 μM all-trans retinal (ATR) (R2500, Sigma) in the dark for 16 hours. The ATR-fed animals were assayed under continuous illumination of 70 μW/mm2 blue light, which was switched on at the start of recording at 7% O2. Blue light was emitted from an ultra-high-power LED lamp (UHP-MIC-LED-460, Prizmatix), and the light intensity was determined using a PM50 Optical Power Meter (ThorLabs). To minimize the effect of transmitted light on the transgenic animals, a long-pass optical filter was used to eliminate the lights with short wavelengths during picking. All videos of behavioral analysis were analyzed using a home-made MATLAB program Zentracker (https://github.com/wormtracker/zentracker). For all behavior analyses, at least three assays were performed for each strain with more than 80 worms in total.

Ca2+ Imaging

L4 animals expressing the GCaMP6s sensor in ASJ neurons were picked 24 hours before imaging. The agarose pads were prepared 1 hour before the experiment. Animals were incubated in 15 µl of 10 mM levamisole for 40 minutes prior to imaging. Between 15 and 20 animals were immobilized using 10 mM levamisole diluted in a concentrated OP50 bacterial suspension and mounted on a 2% agarose pad in M9 buffer on a glass slide. The animals were then covered with a microfluidic chamber. Imaging lasted for 16 minutes. O2 (7%) was pumped into the microfluidic chamber during the first and last 4 minutes. From minute 4 to 12, a mixture of 7% O2 and 150 ppm H2S was pumped.

Imaging was performed on a Nikon AZ100 microscope equipped with a TwinCam adapter (Cairn Research, UK), fitted with two DMK 33U monochrome cameras (The Imaging Source, Germany). A 2x AZ Plan Fluor objective with 1x zoom was used, with an exposure time of 300 ms. Excitation of GCaMP and RFP was provided by a CoolLED pE-300 (CoolLED, UK). The TwinCam adapter was mounted with emission filters for green fluorescent protein (510/20 nm) and red fluorescent protein (590/35 nm), along with a DC/T510LPXRXTUf2 dichroic mirror. Imaging data were analyzed using Neuron Analyzer, a custom-written Matlab program for processing image stacks (code available at https://github.com/neuronanalyser/neuronanalyser).

RNA extraction and sequencing

To obtain RNA samples for RNA-seq, thirty young adult animals were picked to each fresh plate and allowed to lay eggs for 2 hours, after which the adult animals were removed and eggs were allowed to develop into day-one adults. Day-one animals were subsequently challenged with H2S for three different time periods (1 hour, 2 hours and 12 hours). For each time period, five plates of day-one animals were exposed to 50 ppm H2S, 150 ppm H2S or 7% O2 as a control group. Subsequently, the animals were immediately collected and rinsed three times with M9 buffer. After washing, the worm pellet was frozen in liquid nitrogen. Animals were homogenized using Bullet Blender (Next Advance) in the presence of Qiazol Lysis Reagent and 0.5mm Zirconia beads at 4°C. RNA was prepared using RNeasy Plus Universal Mini Kit (Qiagen). Six independent RNA samples were prepared for each condition. The 2100 Bioanalyzer instrument (Agilent) was used for the RNA quality control with a microfluidic chip specific for RNA (Agilent RNA 6000 Nano Kit). Then 1 μg of qualified RNA samples in RNase-free ddH2O was used for library preparation. Library was constructed and sequenced by Novogene.

RNA-seq analysis

The RNA-seq data were aligned using STAR v2.7.9a (Dobin et al., 2013) and gene expression counts extracted by featureCounts v2.0.3 (Liao, Smyth, & Shi, 2014). Differential expression was calculated using DESeq2 (Love, Huber, & Anders, 2014). GO analysis was performed using the EnrichR R package (Kuleshov et al., 2016) on genes having p.adj < 1e-10. For brevity, several highly similar GO categories were manually omitted (full EnrichR output available at Github repository).